---
title: "Overview of imaging mass cytometry datasets"
date: "Created: Nov 02, 2020; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
author:
- name: Nils Eling
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nils.eling@dqbm.uzh.ch
- name: Nicolas Damond
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nicolas.damond@dqbm.uzh.ch
- name: Tobias Hoch
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: tobias.hoch@dqbm.uzh.ch
output:
    BiocStyle::html_document:
        toc: true
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
vignette: >
    %\VignetteIndexEntry{User's Guide}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

```{r library, echo=FALSE}
library(cytomapper)
library(SingleCellExperiment)
# library(imcdatasets)
source("../R/DamondPancreasData.R")
source("../R/listDatasets.R")
```

# Introduction

The `r Biocpkg("imcdatasets")` package provides access to publicly available imaging mass cytometry (IMC) datasets.  

Each dataset is composed of :
* Single-cell data in the form of a `SingleCellExperiment` object.  
* Multichannel images in the form of a `CytoImageList` object.
* Cell masks in the form of a `CytoImageList` object.

Please refer to the `r Biocpkg("SingleCellExperiment")` and `r Biocpkg("cytomapper")` documentation for further information on how to work with `SingleCellExperiment` objects and with the `cytomapper`package.

Users can import the datasets by calling a single function. For example:  

```{r import-dataset}
sce <- DamondPancreasData(data.type = "sce")
sce
```

Import multichannel images

```{r import-images}
images <- DamondPancreasData(data.type = "images")
images
```

Import cell masks

```{r import-masks}
masks <- DamondPancreasData(data.type = "masks")
masks
```


# Available datasets

The `listDatasets()` function returns all available datasets in `r Biocpkg("imcdatasets")`, along with associated information and the R function call to load them.

```{r list-datasets}
datasets <- listDatasets()
datasets <- as.data.frame(datasets)
datasets$FunctionCall <- sprintf("`%s`", datasets$FunctionCall)
knitr::kable(datasets)
```

# References

```{r session-info}
sessionInfo()
```

