---
title: "Accessing IMC datasets"
date: "Created: Nov 02, 2020; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
author:
- name: Nicolas Damond
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nicolas.damond@dqbm.uzh.ch
- name: Nils Eling
  affiliation: Department for Quantitative Biomedicine, University of Zurich
  email: nils.eling@dqbm.uzh.ch
output:
    BiocStyle::html_document:
        toc_float: yes
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
vignette: >
    %\VignetteIndexEntry{"Accessing IMC datasets"}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE,
                      fig.retina = 0.75)
library(BiocStyle)
```

```{r library, echo=FALSE}
library(cytomapper)
library(SingleCellExperiment)
library(imcdatasets)
```

# Introduction

The `r Biocpkg("imcdatasets")` package provides access to publicly available
datasets generated using imaging mass cytometry (IMC) [@giesen2014imc].

IMC is a technology that enables measurement of up to 50 markers from tissue 
sections at a resolution of 1 $\mu m$ @giesen2014imc. In classical processing 
pipelines, such as the [ImcSegmentationPipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline), 
the multichannel images are segmented to generate cells masks. These masks are 
then used to extract single cell features from the multichannel images.  

Each dataset in `imcdatasets` is composed of three elements that can be 
retrieved separately:  
1. Single-cell data in the form of a `SingleCellExperiment` class object (named 
`XYZ_sce.rds`).  
2. Multichannel images in the form of a `CytoImageList` class object (named 
`XYZ_images.rds`).  
3. Cell segmentation masks in the form of a `CytoImageList` class object (named
`XYZ_masks.rds`).  
Here, `XYZ` refers to the name of the dataset.

# Available datasets

The `listDatasets()` function returns all available datasets in `imcdatasets`, 
along with associated information. The `FunctionCall` column gives the name of
the R function that enables to load the dataset.

```{r list-datasets}
datasets <- listDatasets()
datasets <- as.data.frame(datasets)
datasets$FunctionCall <- sprintf("`%s`", datasets$FunctionCall)
knitr::kable(datasets)
```

# Retrieving data

Users can import the datasets by calling a single function and specifying the 
type of data to retrieve. The following examples highlight accessing the dataset
provided by _Damond, N. et al., A Map of Human Type 1 Diabetes Progression by 
Imaging Mass Cytometry_ [@damond2019pancreas].

__Importing single-cell expression data and metadata__

```{r import-dataset}
sce <- DamondPancreas2019Data(data_type = "sce")
sce
```

__Importing multichannel images__

```{r import-images}
images <- DamondPancreas2019Data(data_type = "images")
images
```

__Importing cell segmentation masks__

```{r import-masks}
masks <- DamondPancreas2019Data(data_type = "masks")
masks
```

__On disk storage__

Objects containing multi-channel images and segmentation masks can furthermore
be stored on disk rather than in memory. Nevertheless, they need to be loaded
into memory once before writing them to disk. This process takes longer than
keeping them in memory but reduces memory requirements during downstream
analysis.

To write images or masks to disk, set `on_disk = TRUE` and specify a path
where images/masks will be stored as .h5 files:

```{r on_disk}
# Create temporary location
cur_path <- tempdir()

masks <- DamondPancreas2019Data(data_type = "masks", on_disk = TRUE,
                                h5FilesPath = cur_path)
masks
```


__Accessing data objects by name__

Alternately, data objects can be imported using their name as a function:

```{r access-by-name}
sce <- DamondPancreas2019_sce()
sce

images <- DamondPancreas2019_images()
images

masks <- DamondPancreas2019_masks()
masks
```


# Dataset info and metadata

Additional information about each dataset is available in the help page:

```{r function-help}
?DamondPancreas2019Data
```

The metadata associated with a specific data object can be displayed as follows:

```{r access-metadata}
DamondPancreas2019_sce(metadata = TRUE)
DamondPancreas2019_images(metadata = TRUE)
DamondPancreas2019_masks(metadata = TRUE)
```

# Usage

The `SingleCellExperiment` class objects can be used for data analysis. For more
information, please refer to the `r Biocpkg("SingleCellExperiment")` package 
and to the [Orchestrating Single-Cell Analysis with Bioconductor](http://bioconductor.org/books/release/OSCA/)
workflow.  

The `CytoImageList` class objects can be used for plotting cell and pixel 
information. Some typical use cases are given below. For more information, 
please see the `r Biocpkg("cytomapper")` package and the
[associated vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/cytomapper/inst/doc/cytomapper.html)
.  

__Subsetting the images and masks__

```{r usage-subset}
cur_images <- images[1:5]
cur_masks <- masks[1:5]
```

__Plotting pixel information__  

The `images` objects can be used to display pixel-level data.

```{r usage-pixel}
plotPixels(
    cur_images,
    colour_by = c("CDH", "CD99", "H3"),
    bcg = list(
        CD99 = c(0,2,1),
        CDH = c(0,8,1),
        H3 = c(0,5,1)
    )
)
```

__Plotting cell information__  

The `masks` and `sce` objects can be combined to display cell-level data.

```{r usage-cell}
plotCells(
    cur_masks, object = sce,
    img_id = "ImageNumber", cell_id = "CellNumber",
    colour_by = c("CD8a", "PIN"),
    exprs_values = "exprs"
)
```

__Outlining cells on images__  

Cell information can be displayed on top of images by combining the `images`, 
`masks` and `sce` objects.

```{r usage-outline}
plotPixels(
    cur_images, mask = cur_masks, object = sce,
    img_id = "ImageNumber", cell_id = "CellNumber",
    outline_by = "CellType",
    colour_by = c("H3", "CD99", "CDH"),
    bcg = list(
        CD99 = c(0,2,1),
        CDH = c(0,8,1),
        H3 = c(0,5,1)
    )
)
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
