---
title: "Prepare single-cell data, images and masks from the Jackson and Fischer dataset"
author: "Jana Fischer and Nicolas Damond"
date: "Created: Mar 15, 2021; Compiled: `r BiocStyle::doc_date()`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: console
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

This script loads an example dataset including single-cell data, single-cell masks and IMC tiffs of 100 images (1 image per patient) from the Jackson and Fischer 2020 breast cancer dataset:  

[Jackson, H.W., Fischer, J.R. et al. The single-cell pathology landscape of breast cancer. Nature 578, 615–620 (2020)](https://doi.org/10.1038/s41586-019-1876-x).  

The data is obtained from [zenodo](https://doi.org/10.5281/zenodo.3518284) and processed to create a [SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) object. The [cytomapper package](https://www.bioconductor.org/packages/release/bioc/html/cytomapper.html) is used to read in the images and masks and create `CytoImageList` objects.  


Loading required R packages

```{r libraries, include = FALSE}
library(data.table)
library(SingleCellExperiment)
library(stringr)
library(cytomapper)
library(plyr)
```

Setting the working and output directories

```{r directories}
workdir <- tempdir()
Sys.setenv(workdir = workdir)

outdir <- file.path("inst", "extdata", "JacksonFischer2020")
if(!(dir.exists(outdir)))
    dir.create(outdir)
```

#Single-cell and sample/patient metadata

##Download and unzip single-cell, sample and patient metadata

The data is downloaded from [zenodo](https://doi.org/10.5281/zenodo.3518284).
The first zip folder contains the main single-cell, sample and patient metadata. The single-cell locations and cluster labels are downloaded as separate zip folders because they were uploaded to zenodo separately at a later time point.

###Single cell data and metadata

```{r download single-cell and metadata, message=FALSE}
timeout <- getOption('timeout')
options(timeout=1000)

#Download the zipped folder with single-cell and metadata
url.cells <- ("https://zenodo.org/record/4607374/files/SingleCell_and_Metadata.zip?download=1")
download.file(url.cells, destfile = file.path(
  workdir, "SingleCell_and_Metadata.zip"))

#Unzip file in same directory
system2("unzip", args = c(
  "-o", # include override flag
  file.path(workdir, "SingleCell_and_Metadata.zip"), "-d", workdir),
  stdout = TRUE)

#Remove zip file
file.remove(file.path(workdir, "SingleCell_and_Metadata.zip"))
```

###Single cell cluster labels

```{r download single-cell cluster labels, message=FALSE}
#Download the zipped folder with single-cell cluster labels
url.cluster <- ("https://zenodo.org/record/4607374/files/singlecell_cluster_labels.zip?download=1")
download.file(url.cluster, destfile = file.path(
  workdir, "singlecell_cluster_labels.zip"))

#Unzip file in same directory
system2("unzip", args = c(
  "-o", # include override flag
  file.path(workdir, "singlecell_cluster_labels.zip"), "-d", workdir),
  stdout = TRUE)

#Remove zip file
file.remove(file.path(workdir, "singlecell_cluster_labels.zip"))
```

###Single cell locations

```{r download single-cell locations, message=FALSE}
#Download the zipped folder with single-cell locations
url.clocations <- ("https://zenodo.org/record/4607374/files/singlecell_locations.zip?download=1")
download.file(url.clocations, destfile = file.path(
  workdir, "singlecell_locations.zip"))

#Unzip file in same directory
system2("unzip", args = c(
  "-o", # include override flag
  file.path(workdir, "singlecell_locations.zip"), "-d", workdir), stdout = TRUE)

#Remove zip file
file.remove(file.path(workdir, "singlecell_locations.zip"))
```

##Read single-cell, sample and patient metadata

Basel dataset

```{r read single-cell and metadata}
#Single cell expressions and spatial features
cells <- fread(
  file.path(workdir, "Data_publication/BaselTMA/SC_dat.csv"))

#Sample and patient metadata
meta <- fread(
  file.path(workdir, "Data_publication/BaselTMA/Basel_PatientMetadata.csv"))

#Panel info
panel <- fread(
  file.path(workdir, "Data_publication/Basel_Zuri_StainingPanel.csv"))

#Cluster labels (combine PhenoGraph cluster labels and metacluster labels)
PG_clusters <- fread(
  file.path(workdir, "Cluster_labels/PG_basel.csv"), header = T)
Meta_clusters <- fread(
  file.path(workdir, "Cluster_labels/Basel_metaclusters.csv"), header = T)
clusters <- merge(PG_clusters, Meta_clusters, by = "id", all.x = T)
PG_clusters <- NULL
Meta_clusters <- NULL

#Single-cell locations
locations <- fread(file.path(workdir, "Basel_SC_locations.csv"), header = T)
```

##Select a set of 100 patients as an example dataset

One primary tumor core per patient

```{r select patients}
#Exclude healthy tissue samples and controls
meta <- meta[diseasestatus == "tumor",]

#Exclude patients for which the tumor core was split across 2 images (machine failure)
meta[, ncores := .N,by = "PID"]
meta <- meta[ncores == 1,]

#Randomly sample 100 patients
set.seed(2)
meta <- meta[PID %in% sample(meta$PID, 100, replace = F),]
```

##Subset single-cell data to cores of selected patients

```{r subset single-cell data}
cells <- cells[core %in% meta$core,]
locations <- locations[core %in% meta$core,]
clusters <- clusters[core %in% meta$core,]
```

##Convert cells to single-cell experiment

```{r convert to SCE}
spatial_channels <- c("Area","Eccentricity","Solidity","Extent","EulerNumber",
                      "Perimeter","MajorAxisLength","MinorAxisLength",
                      "Orientation","Percent_Touching","Number_Neighbors")

#Spatial channels go into colData
spatial <- cells[channel %in% spatial_channels,]

#Marker expression channels go into counts
cells <- cells[!channel %in% spatial_channels,]

#Normalize marker counts between 0 and 1 with the 99th percentile
cells[, quant := quantile(mc_counts, 0.99), by = "channel"]
cells[, quant_norm := mc_counts/quant, by = "channel"]
cells[quant_norm > 1, quant_norm := 1]
cells[quant_norm < 0, quant_norm := 0]

#Arcsinh transform marker counts
cells[,arcsinh := asinh(mc_counts), by = "channel"]

#Raw counts to SCE format
cells_raw <- dcast.data.table(cells, "core + CellId + id ~channel",
                              value.var = "mc_counts")
rnames <- cells_raw$id
cells_raw <- as.matrix(cells_raw[, -c("id", "CellId", "core")])
rownames(cells_raw) <- rnames
cells_raw <- t(cells_raw)

#Normalized counts to SCE format
cells_quant <- dcast.data.table(cells, "core + CellId + id ~channel",
                                value.var = "quant_norm")
rnames <- cells_quant$id
cells_quant <- as.matrix(cells_quant[, -c("id", "CellId", "core")])
rownames(cells_quant) <- rnames
cells_quant <- t(cells_quant)

#Asinh counts to SCE format
cells_asinh <- dcast.data.table(cells, "core + CellId + id ~channel",
                                value.var = "arcsinh")
rnames <- cells_asinh$id
cells_asinh <- as.matrix(cells_asinh[, -c("id", "CellId", "core")])
rownames(cells_asinh) <- rnames
cells_asinh <- t(cells_asinh)

#Make sce
sce <- SingleCellExperiment(
  assays = list(counts = cells_raw,
                exprs = cells_asinh,
                quant_norm = cells_quant)
)
cells_asinh <- NULL
cells_quant <- NULL
cells_raw <- NULL
cells <- NULL
```

##Add spatial single-cell features, clusters and metadata to colData

```{r colData}
#Wide format spatial single-cell info
singlecell_meta <- dcast.data.table(
  spatial, "core + CellId + id ~channel", value.var = "mc_counts")

#Add single-cell locations
singlecell_meta <- merge(
  singlecell_meta, locations[, c("Location_Center_X","Location_Center_Y","id")],
  by = "id")

#Add clusters
singlecell_meta <- merge(
  singlecell_meta, clusters[,c("PhenoGraphBasel","cluster","id")],
  by = "id")
setnames(singlecell_meta, "cluster", "metacluster")

#Add sample and patient metadata
meta <- meta[,ImageNb := 1:.N]
singlecell_meta <- merge(
  singlecell_meta,
  meta[,-c("treatment", "DiseaseStage", "AllSamplesSVSp4.Array", "ncores")],
  by = "core", all.x = T)

#Set ImageName, ImageNb and CellNb for cytomapper package
# to map to CytoImageList objects
setnames(singlecell_meta, "FileName_FullStack", "ImageName")
setnames(singlecell_meta, "CellId", "CellNb")

#Make sure order matches id order of sce
singlecell_meta <- singlecell_meta[order(match(id, colnames(sce))),]

#Add colData to sce
rown <- singlecell_meta$id
singlecell_meta <- DataFrame(singlecell_meta)
rownames(singlecell_meta) <- rown
colData(sce) <- singlecell_meta
```

##Clean marker names and add row metadata

```{r rowData}
#Exclude gas channels
sce <- sce[rownames(sce)[!str_detect(rownames(sce),"Hg") & !str_detect(rownames(sce),"In115") & !str_detect(rownames(sce),"I127") & !str_detect(rownames(sce),"Pb") & !str_detect(rownames(sce),"Xe") & !str_detect(rownames(sce),"Ar")]]

#Add row metadata
rows <- data.table(original = rownames(sce))
rows[,marker := sub(".*Di ", "",original)]
rows[,metal := sub("Di .*", "",original)]
rows[,weight := sub(".*[A-Za-z ]", "",metal)]
rows[,metal := gsub("[0-9]+","",metal)]
rows[,`Metal Tag` := paste0(metal,weight)]

#Add metal weight to non-unique names
rows[marker == "Rutheni", marker := paste0("Rutheni_", weight)]
rows[marker == "Iridium", marker := paste0("Iridium_", weight)]

#Clarify unclear names
rows[marker == "cleaved", marker := "CC3/cPARP"]
rows[marker == "cerbB", marker := "HER2"]
rows[`Metal Tag` == "Nd143", marker := "CK19"]
rows[`Metal Tag` == "Sm147", marker := "CK14"]
rows[`Metal Tag` == "Pr141", marker := "CK5"]
rows[`Metal Tag` == "La139", marker := "H3K27me3"]
rows[marker == "Carboni" ,marker := "CAIX"]
rows[`Metal Tag` == "Nd144", marker := "CK8/18"]
rows[`Metal Tag` == "Yb174", marker := "CK7"]

#Add missing metal info
rows[marker == "phospho Histone", `Metal Tag` := "Eu153"]
rows[marker == "phospho S6", `Metal Tag` := "Er170"]
rows[marker == "phospho mTOR", `Metal Tag` := "Yb173"]

#Add channel number from panel
rows <- merge(rows, unique(panel[, c("Metal Tag", "FullStack")]),
              by = "Metal Tag", all.x = T)

#Add rowData to sce
rows <- rows[order(match(original, rownames(sce)))]
rownames(sce) <- rows$marker
rowd <- DataFrame(rows[, -"marker"])
rownames(rowd) <- rows$marker
rowData(sce) <- rowd
```

##Save out sce object

Save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save sce}
saveRDS(sce, file.path(outdir, "sce.rds"))
```

## Clean up

Remove the downloaded objects to save storage space.

```{r clean-up1, message = FALSE}
file.remove(file.path(workdir, "Basel_SC_locations.csv"),
            file.path(workdir, "Zurich_SC_locations.csv"))
unlink(file.path(workdir, "Data_publication"), recursive = TRUE)
unlink(file.path(workdir, "Cluster_labels"), recursive = TRUE)
```

#Image tiffs and masks

##Download and unzip tiffs and masks

```{r download images and masks, message=FALSE}
#Download the zipped folder with tiffs and masks
url.imgs <- ("https://zenodo.org/record/4607374/files/OMEandSingleCellMasks.zip?download=1")
download.file(url.imgs, destfile = file.path(
  workdir, "OMEandSingleCellMasks.zip"))

#Unzip main folder
system2("unzip", args = c(
  "-o", # include override flag
  file.path(workdir, "OMEandSingleCellMasks.zip"),"-d", workdir),
  stdout = TRUE)

#Remove zip file
file.remove(file.path(workdir, "OMEandSingleCellMasks.zip"))

#Unzip subfolder with masks
system2("unzip", args = c(
  "-o", # include override flag
  file.path(workdir, "OMEnMasks", "Basel_Zuri_masks.zip"),"-d", workdir),
  stdout = TRUE)

#Unzip subfolder with tiffs
system2("unzip", args = c(
  "-o", # include override flag
  file.path(workdir, "OMEnMasks", "ome.zip"), "-d", workdir),
  stdout = TRUE)

#Remove zip files
file.remove(file.path(workdir, "OMEnMasks/Basel_Zuri_masks.zip"))
file.remove(file.path(workdir, "OMEnMasks/ome.zip"))
```

##Load and process masks and images into cytomapper image list

The masks are 16-bit images and need to be re-scaled in order to obtain integer cell IDs. Also, occasionally a single-cell ID is skipped in the masks but in the single-cell data the CellNbs were renamed sequentially. Therefore, the single-cell IDs in the masks also have to renamed sequentially in order to correspond to the CellNbs from the single-cell data.

```{r load and process masks}
#Load the masks as a CytoImageList object
masks <- loadImages(file.path(workdir, "Basel_Zuri_masks/"),
                    pattern = "_full_maks.tiff")

#Metadata names according to full filename in sce
mcols(masks)$ImageName <- gsub("_full_maks", "_full.tiff", names(masks))

#Subset to selected 100 images corresponding to sce
masks <- masks[mcols(masks)$ImageName %in% unique(sce$ImageName)]

#Scale masks
masks <- scaleImages(masks, value = (2 ^ 16) - 1)

#Rename single-cell IDs sequentially in each mask
for (n in names(masks)){
  imageData(masks[[n]]) = mapvalues(
    imageData(masks[[n]]),
    sort(unique(as.integer(imageData(masks[[n]])))),
    0:(length(unique(as.integer(imageData(masks[[n]]))))-1)
  )
}

#Add ImageNb to metadata
mcols(masks) <- merge(mcols(masks),
                       unique(colData(sce)[, c("ImageName", "ImageNb")]),
                       by = "ImageName")

#Save masks as CytoImageList object
saveRDS(masks, file.path(outdir, "masks.rds"))
```

Loading all tiffs would need too much memory -> only keep tiffs that we want to load in folder and delete rest

```{r load and process tiffs}
#Delete tiff stacks of images we don't want to read from folder
infold = list.files(file.path(workdir, "ome/"))
file.remove(file.path(workdir, "ome/",
                      infold[!infold %in% unique(sce$ImageName)]))

#Load the tiffs as a CytoImageList object
images <- loadImages(file.path(workdir, "ome/"), pattern = "_full.tiff")

#Metadata names according to full filename in sce
mcols(images)$ImageName <- paste0(names(images), ".tiff")

#Double check that the images correspond to filenames in the sce
images <- images[mcols(images)$ImageName %in% sce$ImageName]

#Add ImageNb to metadata
mcols(images) <- merge(mcols(images),
                       unique(colData(sce)[, c("ImageName", "ImageNb")]),
                       by = "ImageName")

#Exclude channels that weren't used in sce
d <- as.data.table(rowData(sce))
d$rown <- rownames(sce)
d <- d[order(FullStack),]
images <- getChannels(images,d$FullStack)

#Add channel names to CytoImageList object
channelNames(images) <- d$rown

#Check if names of masks and tiffs are equal
identical(mcols(masks)$ImageName, mcols(images)$ImageName)

#Save tiff images as CytoImageList object
saveRDS(images, file.path(outdir, "images.rds"))
```

## Clean up

Remove the downloaded objects to save storage space.

```{r clean-up2, message = FALSE}
unlink(file.path(workdir, "OMEnMasks"), recursive = TRUE)
unlink(file.path(workdir, "Basel_Zuri_masks"), recursive = TRUE)
unlink(file.path(workdir, "ome"), recursive = TRUE)
options(timeout=timeout)
```

# Session information

```{r session-info}
sessionInfo()
```
