---
title: "Preparation of the Jackson, Fischer et al dataset"
author: "Jana Fischer and Nicolas Damond"
date: "Created: 15 March 2021; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: inline
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script downloads a hundred images (one image per patient), as well as the associated single-cell data and cell segmentation masks from the breast tumour Imaging Mass Cytometry (IMC) dataset described in the following publication:  

[Jackson, H.W., Fischer, J.R. et al. The single-cell pathology landscape of breast cancer. Nature 578, 615–620 (2020)](https://doi.org/10.1038/s41586-019-1876-x).  

All data are openly available from [zenodo](https://doi.org/10.5281/zenodo.3518284).  

Here, we will download single cell data and metadata, and process them to create a [SingleCellExperiment](https://bioconductor.org/packages//SingleCellExperiment.html) object. We will then download the corresponding multichannel IMC images and cell segmentation masks and format them into `CytoImageList` objects using the [cytomapper](https://bioconductor.org/packages/cytomapper) package.


# **Settings**

```{r libraries, include = FALSE}
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
dataset_name <- "JacksonFischer_2020_BreastCancer"
dataset_version <- "v2"
cat("Dataset version:", dataset_version)
```

Setting the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
dataset_dir <- file.path(".", dataset_name)
if(!(dir.exists(dataset_dir))) dir.create(dataset_dir)

outdir <- file.path(dataset_dir, dataset_version)
if(!(dir.exists(outdir))) dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


# **Single cell data**

We will download single-cell data corresponding from
[@JacksonFischer-2020-BreastCancer] from 
[zenodo](https://doi.org/10.5281/zenodo.3518284).

## Download single cell data

### Import function

Function to download and unzip files.

```{r import-function}
importData <- function(url, output_dir, filename) {
  # Download
  download.file(url, destfile = file.path(output_dir, filename))
  
  # Unzip
  system2("unzip", args = c("-o",
                            file.path(output_dir, filename),
                            "-d", output_dir),
          stdout = TRUE)
  
  # Remove zipped folder
  file.remove(file.path(output_dir, filename))
}
```

### Single cell data

The first zip folder contains the main single-cell, sample and patient metadata. The single-cell locations and cluster labels are downloaded as separate zip folders because they were uploaded to zenodo separately at a later time point.

```{r import-cell-data, results='hide'}
# Download dataset
url_cells <- ("https://zenodo.org/record/4607374/files/SingleCell_and_Metadata.zip?download=1")
zip_name <- "SingleCell_and_Metadata.zip"
download.file(url_cells, destfile = file.path(workdir, zip_name))

# Unzip required files - Basel Cohort
system2("unzip", args = c(
    "-o", file.path(workdir, zip_name),
    "Data_publication/BaselTMA/SC_dat.csv",
    "-d", workdir))
system2("unzip", args = c(
    "-o", file.path(workdir, zip_name),
    "Data_publication/BaselTMA/Basel_PatientMetadata.csv",
    "-d", workdir))

# Unzip required files - Zurich Cohort
system2("unzip", args = c(
    "-o", file.path(workdir, zip_name),
    "Data_publication/ZurichTMA/SC_dat.csv",
    "-d", workdir))
system2("unzip", args = c(
    "-o", file.path(workdir, zip_name),
    "Data_publication/ZurichTMA/Zuri_PatientMetadata.csv",
    "-d", workdir))

# Unzip panel file
system2("unzip", args = c(
    "-o", file.path(workdir, zip_name),
    "Data_publication/Basel_Zuri_StainingPanel.csv",
    "-d", workdir))

file.remove(file.path(workdir, zip_name))
```

Download single cell labels and locations

```{r import-cell-metadata, results='hide'}
# Clusters
url_cluster <- ("https://zenodo.org/record/4607374/files/singlecell_cluster_labels.zip?download=1")
importData(url_cluster, workdir, "singlecell_cluster_labels.zip")

# Cell locations
url_locations <- ("https://zenodo.org/record/4607374/files/singlecell_locations.zip?download=1")
importData(url_locations, workdir, "singlecell_locations.zip")
```


## Read in single-cell data

We read in single cell data linked to the Basel and Zurich cohorts, including 
single cell data, cell metadata, clinical data, antibody panel information, and
cell clusters.

```{r read-in-cell-data}
# Single cell expressions and spatial features
cells <- rbind(
    fread(file.path(workdir, "Data_publication/BaselTMA/SC_dat.csv")),
    fread(file.path(workdir, "Data_publication/ZurichTMA/SC_dat.csv"))
)

# Sample and clinical metadata
cell_meta_basel <- fread(file.path(workdir,
                    "Data_publication/BaselTMA/Basel_PatientMetadata.csv"))
cell_meta_zuri <- fread(file.path(workdir,
                    "Data_publication/ZurichTMA/Zuri_PatientMetadata.csv"))

cell_meta_zuri[, HER2Status := HER2]
cell_meta_zuri[ ,':=' (HER2 = NULL, ER = NULL, PR = NULL)]  

cell_meta <- as.data.table(rbind(
  data.frame(c(cell_meta_basel, sapply(
      setdiff(names(cell_meta_zuri),names(cell_meta_basel)), function(x) NA))),
  data.frame(c(cell_meta_zuri, sapply(
      setdiff(names(cell_meta_basel),names(cell_meta_zuri)), function(x) NA)))
))

# Panel information
panel <- fread(file.path(
  workdir, "Data_publication/Basel_Zuri_StainingPanel.csv"))

# Cluster labels (merge PhenoGraph cluster labels and metacluster labels)
pg_clusters_basel <- fread(file.path(
  workdir, "Cluster_labels/PG_basel.csv"), header = TRUE)
setnames(pg_clusters_basel, "PhenoGraphBasel", "PhenoGraph")

pg_clusters_zuri <- fread(file.path(
  workdir, "Cluster_labels/PG_zurich.csv"), header = TRUE)
pg_clusters <- rbind(pg_clusters_basel,
                             pg_clusters_zuri)

meta_clusters_basel <- fread(file.path(
  workdir, "Cluster_labels/Basel_metaclusters.csv"), header = TRUE)
meta_clusters_zuri <- fread(file.path(
  workdir, "Cluster_labels/Zurich_matched_metaclusters.csv"), header = TRUE)
meta_clusters <- rbind(meta_clusters_basel, meta_clusters_zuri)

clusters <- merge(pg_clusters, meta_clusters, by = "id", all.x = TRUE)

# Single-cell locations
locations_basel <- fread(file.path(
  workdir, "Basel_SC_locations.csv"), header = TRUE)
locations_zuri <- fread(file.path(
  workdir, "Zurich_SC_locations.csv"), header = TRUE)
locations <- rbind(locations_basel, locations_zuri)

# Remove data frames that are not needed anymore
remove(cell_meta_basel, cell_meta_zuri)
remove(locations_basel, locations_zuri)
remove(pg_clusters_basel, pg_clusters_zuri, pg_clusters)
remove(meta_clusters_basel, meta_clusters_zuri, meta_clusters)
```

Filter out non-breast images

```{r remove-non-breast-images}
img_to_remove <- paste(c("Liver", "control", "non-breast"), collapse = "|")
cells <- cells[grep(img_to_remove, cells$core, invert = TRUE), ]
```


## Prepare data

### Extract spatial information

We first extract channels related to spatial information, such as cell area or number of neighbors.

```{r split-cell-metadata}
spatial_channels <- c(
  "Area", "Eccentricity", "Solidity", "Extent", "EulerNumber", "Perimeter",
  "MajorAxisLength","MinorAxisLength", "Orientation", "Percent_Touching",
  "Number_Neighbors"
)

# Spatial channels will go to the colData slot of the SCE
spatial <- cells[channel %in% spatial_channels, ]

# Marker expression levels will go to the assay slot of the SCE
cells <- cells[!channel %in% spatial_channels,]
```

### Cell-level metadata

Here, we will collect all cell-specific metadata in a single `DataFrame`, which will constitute the `colData` entry of the final `SingleCellExperiment` object.

Columns are renamed for consistency with the other datasets.

```{r cell-metadata}
# Wide format spatial single-cell info
cell_metadata <- dcast.data.table(
  spatial, "core + CellId + id ~ channel", value.var = "mc_counts")

# Subset and rename columns
cell_metadata <- cell_metadata[, .(
  image_name = core,
  cell_id = id,
  neighbors_number = Number_Neighbors,
  neighbors_percent_touching = Percent_Touching,
  cell_area = Area,
  cell_perimeter = Perimeter,
  cell_eccentricity = Eccentricity,
  cell_euler_number = EulerNumber,
  cell_extent = Extent,
  cell_major_axis_length = MajorAxisLength,
  cell_minor_axis_length = MinorAxisLength,
  cell_orientation = Orientation,
  cell_solidity = Solidity
)]

# Add single-cell locations
locations <- locations[, .(
  cell_id = id,
  cell_x = Location_Center_X,
  cell_y = Location_Center_Y
)]
cell_metadata <- merge(cell_metadata, locations, by = "cell_id")

# Add clusters
clusters <- clusters[, .(
  cell_id = id,
  cell_cluster_phenograph = PhenoGraph,
  cell_metacluster = cluster
)]
cell_metadata <- merge(cell_metadata, clusters, by = "cell_id")

# Add sample and patient metadata
cell_meta <- cell_meta[, image_number := 1:.N]

# Rename columns
cell_meta <- cell_meta[, .(
  image_name = core,
  image_number = image_number,
  cells_per_image = Count_Cells,
  image_width = Width_FullStack,
  image_height = Height_FullStack,
  image_area = area,
  image_filename = FileName_FullStack,
  image_sum_area_cells = sum_area_cells,
  image_percent_tumor_cells = X.tumorcells,
  image_percent_normal_epithelial_cells = X.normalepithelialcells,
  image_percent_stroma = X.stroma,
  image_percent_inflammatory_cells = X.inflammatorycells,
  patient_id = PID,
  patient_age = age,
  patient_gender = gender,
  patient_status = Patientstatus,
  patient_disease_status = diseasestatus,
  pateint_menopausal = menopausal,
  patient_DFS_months = DFSmonth,
  patient_OS_months = OSmonth,
  patient_year_sample_collection = Yearofsamplecollection,
  TMA_location = TMALocation,
  TMA_x = TMAxlocation,
  TMA_y = yLocation,
  TMA_block_label = TMABlocklabel,
  TMA_UBTMA_location = UBTMAlocation,
  tumor_grade = grade,
  tumor_type = tumor_type,
  tumor_subtype = Subtype,
  tumor_clinical_type = clinical_type,
  tumor_location = location,
  tumor_size = tumor_size,
  tumor_primary_site = PrimarySite,
  tumor_primary_diagnosis = Ptdiagnosis,
  tumor_ER_status = ERStatus,
  tumor_HER2_status = HER2Status,
  tumor_HR_status = HR,
  tumor_PR_status = PRStatus,
  tumor_ERpos_ductal_ca = ER.DuctalCa,
  tumor_triple_neg_ductal = TripleNegDuctal,
  tumor_hormone_sensitive = hormonesensitive,
  tumor_hormone_resistant_after_sensitive = hormoneresistantaftersenstive,
  tumor_I_plus_neg = I_plus_neg,
  tumor_PTNM_M = PTNM_M,
  tumor_PTNM_N = PTNM_N,
  tumor_PTNM_T = PTNM_T,
  tumor_PTNM_radicality = PTNM_Radicality,
  tumor_microinvasion = microinvasion,
  tumor_lymphatic_invation = Lymphaticinvasion,
  tumor_venous_invasion = Venousinvasion,
  tumor_SN = SN,
  tumor_histology = histology,
  tumor_pre_surgery_Tx_type = Pre.surgeryTx,
  tumor_post_surgery_Tx_type = Post.surgeryTx,
  tumor_post_surgery_Tx = Post_surgeryTx,
  tumor_response = response
)]

cell_meta[tumor_HER2_status == "+", ]$tumor_HER2_status <- "positive"
cell_meta[tumor_HER2_status == "-", ]$tumor_HER2_status <- "negative"
cell_meta[tumor_HER2_status == "", ]$tumor_HER2_status <- NA
```

We specify the cohort (Zurich or Basel), and merge the two cell metadata data 
frames.

```{r merge-metadata}
# Specify the cohort of origin
cell_meta$patient_cohort <- ""
cell_meta[grep("BaselTMA", image_name), ]$patient_cohort <- "Basel"
cell_meta[grep("ZTMA", image_name), ]$patient_cohort <- "Zurich"

# Merg the data frames
cell_metadata <- merge(cell_metadata, cell_meta,
                       by = "image_name", all.x = TRUE)
cell_metadata <- DataFrame(cell_metadata)
```

Finally, we add unique cell ids as row names, add cell numbers, and order the cell metadata object based on `image_number` and `cell_number`.

```{r cell-rownames}
# Cell ids are used as row names
cell_metadata$cell_number <- as.integer(sub(".*_", "", cell_metadata$cell_id))
cell_metadata$cell_id <- paste(cell_metadata$image_name,
                               cell_metadata$cell_number, sep = "_")
rownames(cell_metadata) <- cell_metadata$cell_id

# Rows are ordered by image and cell numbers
cell_metadata <- cell_metadata[order(cell_metadata$image_number,
                                     cell_metadata$cell_number), ]
```

### Marker metadata

Here, we will collect all marker-related information and collect it in a `DataFrame` that will constitute the `rowData` slot of the `SingleCellExperiment` object.  

We first rename markers for consistency with other datasets.

```{r rename-markers}
# Exclude gas channels
gas_channels <- c("Hg", "In115", "I127", "Pb", "Xe", "Ar")
cells <- cells[!grepl(paste(gas_channels, collapse = "|"), cells$channel),
               .(image_name = core, cell_id = id, channel, counts = mc_counts)]

# Fix marker and metal names
cells[, full_name := sub(".*Di ", "", channel)]
cells[, metal := sub("Di .*", "", channel)]
cells[, weight := sub(".*[A-Za-z ]", "", metal)]
cells[, metal := gsub("[0-9]+", "", metal)]
cells[, metal := paste0(metal, weight)]

cells[full_name == "Rutheni", `:=` (short_name = metal, full_name = metal)]
cells[full_name == "Iridium", `:=` (short_name = metal, full_name = metal)]

# Add missing metal info
cells[full_name == "phospho Histone", `:=` (
  metal = "Eu153", full_name = "phospho-Histone H3 [S28]",
  short_name = "p_H3")]
cells[full_name == "phospho S6", `:=` (
  metal = "Er170", full_name = "phospho-S6 [S235/S236]",
  short_name = "p_S6")]
cells[full_name == "phospho mTOR", `:=` (
  metal = "Yb173", full_name = "phospho-mTOR [S2448]",
  short_name = "p_mTOR")]

# Clarify unclear names
cells[full_name == "cleaved", `:=` (
  full_name = "cleaved-PARP + cleaved-Caspase3", short_name = "cPARP_cCASP3")]
cells[full_name == "cerbB", `:=` (
  full_name = "Epidermal growth factor receptor-2", short_name = "HER2")]
cells[full_name == "Carboni", `:=` (
  full_name = "Carbonic anhydrase IX", short_name = "CA9")]

cells[metal == "In113", `:=` (full_name = "Histone H3", short_name = "H3")]
cells[metal == "La139", `:=` (full_name = "H3K27me3",
                              short_name = "H3K27me3")]
cells[metal == "Pr141", `:=` (full_name = "Cytokeratin 5",
                              short_name = "KRT5")]
cells[metal == "Nd142", `:=` (full_name = "Fibronectin", short_name = "FN1")]
cells[metal == "Nd143", `:=` (full_name = "Cytokeratin 19",
                              short_name = "KRT19")]
cells[metal == "Nd144", `:=` (full_name = "Cytokeratin 8/18",
                              short_name = "KRT8_18")]
cells[metal == "Nd145", `:=` (full_name = "Twist", short_name = "TWIST1")]
cells[metal == "Sm147", `:=` (full_name = "Cytokeratin 14",
                              short_name = "KRT14")]
cells[metal == "Nd148", `:=` (full_name = "Smooth muscle actin",
                              short_name = "SMA")]
cells[metal == "Sm149", `:=` (full_name = "Vimentin", short_name = "VIM")]
cells[metal == "Nd150", `:=` (full_name = "c-Myc", short_name = "c_Myc")]
cells[metal == "Sm152", `:=` (full_name = "CD3 epsilon", short_name = "CD3e")]
cells[metal == "Gd155", `:=` (full_name = "Slug", short_name = "SNAI2")]
cells[metal == "Tb159", `:=` (full_name = "p53", short_name = "p53")]
cells[metal == "Gd156", `:=` (full_name = "Estrogen receptor alpha",
                              short_name = "ERa")]
cells[metal == "Gd158", `:=` (full_name = "Progesterone receptor A/B",
                              short_name = "PGR")]
cells[metal == "Ho165", `:=` (full_name = "Beta-catenin",
                              short_name = "CTNNB")]
cells[metal == "Er167", `:=` (full_name = "E-Cadherin", short_name = "CDH1")]
cells[metal == "Er168", `:=` (full_name = "Ki-67", short_name = "Ki67")]
cells[metal == "Tm169", `:=` (full_name = "Epidermal growth factor receptor",
                         short_name = "EGFR")]
cells[metal == "Yb171", `:=` (full_name = "Transcription factor SOX-9",
                              short_name = "SOX9")]
cells[metal == "Yb172", `:=` (full_name = "von Willebrand factor",
                              short_name = "vWF")]
cells[metal == "Yb174", `:=` (full_name = "Cytokeratin 7",
                              short_name = "KRT7")]
cells[metal == "Lu175", `:=` (full_name = "Pan-cytokeratin",
                              short_name = "PanCK")]
cells[metal == "Ir191", `:=` (full_name = "Iridium 191", short_name = "DNA1")]
cells[metal == "Ir193", `:=` (full_name = "Iridium 193", short_name = "DNA2")]

# Add missing short names
cells[full_name %in% c("CD68", "p53", "CD44", "CD45", "GATA3", "CD20", "EpCAM"),
      short_name := full_name]

# Ruthenium channels
cells[startsWith(full_name, "Ru"),
      full_name := gsub("Ru", "Ruthenium ", full_name)]
```

We then import the panel and fix antibody clone names.

```{r assemble-panel}
cells$name <- cells$channel

# Select columns
panel <- panel[, .(
  channel = FullStack,
  metal = `Metal Tag`,
  antibody_clone = `Antibody Clone`
)]

# Fix antibody clones names
panel[metal == "Gd158", antibody_clone := "EP2 + SP2"]
panel[metal == "Yb176", antibody_clone := "F21-852 + C92-605"]
panel <- panel[!duplicated(metal), ]
panel <- panel[metal %in% unique(cells$metal)]
panel[antibody_clone == "", antibody_clone := NA]
panel[metal == "Sm147", `:=` (antibody_clone = "polyclonal_CK14")]
panel[metal == "Eu153", `:=` (antibody_clone = "HTA28")]
panel[metal == "Gd156", `:=` (antibody_clone = "polyclonal_anti_rabbit_IgG")]
panel[metal == "Er167", `:=` (antibody_clone = "36/E-Cadherin")]
panel[metal == "Yb172", `:=` (antibody_clone = "polyclonal_vWF")]

# Merge all panel information
panel <- merge(unique(cells[, .(metal, name, full_name, short_name)]),
               panel, by = "metal")
```

Finally, we convert the panel table to a `DataFrame` and add target short_names as row names.

```{r prepare-panel}
panel <- panel[order(panel$channel), ]
panel <- as(panel, "DataFrame")
rownames(panel) <- panel$short_name
```

### Counts matrix

Here, we will prepare the counts matrix that will be stored in the `assay` slot of the `SingleCellExperiment` object.

We extract marker expression values from the `cells` table and convert them to a matrix. We then order the counts matrix by `cell_id` and `short_name`.
We then convert `cell_id` to the {`image_number` `_` `cell_number`} format for consistency with other datasets.

```{r prepare-counts}
# Filter "cells" data frame
cells <- cells[!is.na(cells$short_name), ]
cells <- cells[cells$image_name %in% cell_metadata$image_name, ]

# Create count matrix
counts <- dcast.data.table(cells, "cell_id ~ short_name",
                           value.var = "counts")
row_names <- counts$cell_id
counts[, cell_id := NULL]
counts <- as.matrix(counts, rownames = NULL)
rownames(counts) <- row_names

counts <- counts[order(match(rownames(counts), rownames(cell_metadata))),
                 order(match(colnames(counts), rownames(panel)))]

# Add cell_id
cell_metadata$cell_id <- paste(cell_metadata$image_number,
                               cell_metadata$cell_number, sep = "_")
rownames(cell_metadata) <- cell_metadata$cell_id

rownames(counts) <- rownames(cell_metadata)
```


## Create SingleCellExperiment object

### Create the object

We have now obtained all metadata and feature data to create the `SingleCellExperiment` object.

```{r create-SCE}
sce <- SingleCellExperiment(
  assays = list(counts = t(counts)),
  rowData = panel,
  colData = cell_metadata
)
```

### Counts transformations

We apply two different counts transformations:
- `exprs`: arcsinh-transformed counts (cofactor = 1).
- `quant_norm`: censored + quantile-normalized counts.

```{r transform-counts}
assay(sce, "exprs") <- asinh(counts(sce) / 1)

quant <- apply(assay(sce, "counts"), 1, quantile, probs = 0.99, na.rm = TRUE)
assay(sce, "quant_norm") <- apply(assay(sce, "counts"), 2,
                                  function(x) x / quant)
assay(sce, "quant_norm")[assay(sce, "quant_norm") > 1] <- 1
assay(sce, "quant_norm")[assay(sce, "quant_norm") < 0] <- 0
```

### Subset patients and images

We first subset the Zurich cohort to one image per patient.

```{r subset-sce-zurich}
sce_zurich <- sce[, sce$patient_cohort == "Zurich"]
mainExpName(sce_zurich) <- paste(dataset_name, "Zurich", dataset_version,
    sep = "_")

# Randomly select one image per patient
coldata_zurich <- as.data.table(colData(sce_zurich))
set.seed(2)
image_sub_zurich <- coldata_zurich[
    coldata_zurich[, .I[sample(.N, 1)], by = patient_id]$V1]$image_name

# Select 100 patients from the Zurich cohort
sce_zurich <- sce_zurich[, sce_zurich$image_name %in% image_sub_zurich]
remove(coldata_zurich)
```

For the Basel cohort, we select a set of a hundred patients as an example 
dataset. For consistency, the selection is based on a subsetting vector 
obtained from the first `imcdatasets` version.  

```{r subsetting-vector-basel}
image_sub_basel <- c("BaselTMA_SP41_257_X3Y1", "BaselTMA_SP41_166_X15Y4",
                     "BaselTMA_SP41_153_X7Y5", "BaselTMA_SP41_58_X15Y1", 
                     "BaselTMA_SP41_135_X8Y5", "BaselTMA_SP41_220_X10Y5", 
                     "BaselTMA_SP41_284_X7Y2", "BaselTMA_SP41_18_X13Y5",
                     "BaselTMA_SP41_42_X14Y5", "BaselTMA_SP41_141_X11Y2",
                     "BaselTMA_SP41_177_X16Y5", "BaselTMA_SP41_45_X3Y3",
                     "BaselTMA_SP41_117_X13Y3", "BaselTMA_SP41_57_X8Y6",
                     "BaselTMA_SP41_227_X9Y6", "BaselTMA_SP41_234_X10Y6", 
                     "BaselTMA_SP41_86_X15Y3", "BaselTMA_SP41_214_X15Y6", 
                     "BaselTMA_SP41_237_X16Y6", "BaselTMA_SP41_61_X3Y4",
                     "BaselTMA_SP41_186_X5Y4", "BaselTMA_SP41_38_X4Y7",
                     "BaselTMA_SP41_114_X13Y4", "BaselTMA_SP41_11_X13Y7", 
                     "BaselTMA_SP41_112_X5Y8", "BaselTMA_SP41_53_X6Y8", 
                     "BaselTMA_SP41_129_X7Y8", "BaselTMA_SP41_203_X8Y8", 
                     "BaselTMA_SP41_101_X10Y8", "BaselTMA_SP41_255_X12Y8", 
                     "BaselTMA_SP41_267_X13Y8", "BaselTMA_SP41_48_X14Y8", 
                     "BaselTMA_SP41_212_X1Y9", "BaselTMA_SP41_63_X4Y9", 
                     "BaselTMA_SP42_13_X5Y8", "BaselTMA_SP42_51_X1Y2", 
                     "BaselTMA_SP42_120_X5Y2", "BaselTMA_SP42_217_X1Y3", 
                     "BaselTMA_SP42_10_X1Y5", "BaselTMA_SP42_160_X13Y5", 
                     "BaselTMA_SP42_98_X11Y5", "BaselTMA_SP42_91_X11Y3", 
                     "BaselTMA_SP42_192_X8Y5", "BaselTMA_SP42_185_X7Y5",
                     "BaselTMA_SP42_145_X1Y4", "BaselTMA_SP42_16_X3Y4", 
                     "BaselTMA_SP42_275_X5Y4", "BaselTMA_SP42_89_X4Y6", 
                     "BaselTMA_SP42_56_X2Y6", "BaselTMA_SP42_130_X1Y6", 
                     "BaselTMA_SP42_136_X9Y4", "BaselTMA_SP42_158_X12Y6", 
                     "BaselTMA_SP42_261_X11Y6", "BaselTMA_SP42_174_X9Y6", 
                     "BaselTMA_SP42_279_X14Y6", "BaselTMA_SP42_152_X15Y6", 
                     "BaselTMA_SP42_176_X3Y7", "BaselTMA_SP42_99_X5Y7", 
                     "BaselTMA_SP42_273_X6Y7", "BaselTMA_SP42_5_X8Y7", 
                     "BaselTMA_SP42_236_X11Y7", "BaselTMA_SP42_169_X3Y8",
                     "BaselTMA_SP42_184_X6Y8", "BaselTMA_SP42_71_X8Y8",
                     "BaselTMA_SP42_179_X13Y8", "BaselTMA_SP42_163_X1Y9", 
                     "BaselTMA_SP42_59_X3Y9", "BaselTMA_SP43_87_X15Y4", 
                     "BaselTMA_SP43_142_X5Y1", "BaselTMA_SP43_17_X11Y4", 
                     "BaselTMA_SP43_233_X13Y7", "BaselTMA_SP43_243_X13Y3", 
                     "BaselTMA_SP43_278_X3Y2", "BaselTMA_SP43_124_X10Y8", 
                     "BaselTMA_SP43_180_X5Y2", "BaselTMA_SP43_4_X15Y3", 
                     "BaselTMA_SP43_118_X13Y5", "BaselTMA_SP43_119_X9Y8", 
                     "BaselTMA_SP43_266_X7Y8", "BaselTMA_SP43_122_X12Y8", 
                     "BaselTMA_SP43_49_X1Y5", "BaselTMA_SP43_244_X3Y1", 
                     "BaselTMA_SP43_47_X16Y4", "BaselTMA_SP43_200_X9Y6", 
                     "BaselTMA_SP43_90_X7Y6", "BaselTMA_SP43_147_X6Y6",
                     "BaselTMA_SP43_207_X1Y6", "BaselTMA_SP43_93_X15Y5", 
                     "BaselTMA_SP43_97_X16Y6", "BaselTMA_SP43_66_X15Y6", 
                     "BaselTMA_SP43_235_X11Y2", "BaselTMA_SP43_107_X4Y7", 
                     "BaselTMA_SP43_170_X3Y3", "BaselTMA_SP43_209_X11Y1", 
                     "BaselTMA_SP43_50_X7Y1", "BaselTMA_SP43_43_X5Y5", 
                     "BaselTMA_SP43_199_X8Y5", "BaselTMA_SP43_115_X4Y8", 
                     "BaselTMA_SP43_226_X6Y9", "BaselTMA_SP43_269_X5Y9")
```

```{r subset-sce-basel}
# Subset the Basel cohort
sce_basel <- sce[, sce$patient_cohort == "Basel"]
mainExpName(sce_basel) <- paste(dataset_name, "Basel", dataset_version,
    sep = "_")

# Select 100 patients based on the subsetting vector above
sce_basel <- sce_basel[, sce_basel$image_name %in% image_sub_basel]
```

### Save on disk

We save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
# Full dataset
mainExpName(sce) <- paste(dataset_name, "FULL", dataset_version, sep = "_")
saveRDS(sce, file.path(outdir, "sce_full.rds"))
print(sce)

# Subampled Basel cohort
saveRDS(sce_basel, file.path(outdir, "sce_basel.rds"))
print(sce_basel)

# Zurich cohort
saveRDS(sce_zurich, file.path(outdir, "sce_zurich.rds"))
print(sce_zurich)
```

### Clean up

Finally, we remove the downloaded files and generated objects to save storage space.

```{r clean-up-cell-data, results='hide'}
remove(cells, cell_metadata, spatial, counts, locations, clusters, row_names)

file.remove(file.path(workdir, "Basel_SC_locations.csv"),
            file.path(workdir, "Zurich_SC_locations.csv"))
unlink(file.path(workdir, "Data_publication"), recursive = TRUE)
unlink(file.path(workdir, "Cluster_labels"), recursive = TRUE)
```



# **Images and masks**

Here, we will download multichannel images from [@JacksonFischer-2020-BreastCancer], as well as the corresponding cell segmentation masks. Images and masks correspond to the data in the `SingleCellExperiment` object and will be formatted into `CytoImageList` objects.

## Download images and masks

We download the zip file containing images and masks.

```{r import-images, results='hide'}
# Download images and masks
url_images <- ("https://zenodo.org/record/4607374/files/OMEandSingleCellMasks.zip?download=1")
importData(url_images, workdir, "OMEandSingleCellMasks.zip")
```


## Cell segmentation masks

We will import and process cell segmentation masks to make them compatible with the `cytomapper` package.

### Import masks

We unzip cell segmentation masks and read them into a `CytoImageList` object.

```{r import-masks, results='hide'}
# Unzip
fn <- file.path("OMEnMasks", "Basel_Zuri_masks.zip")
system2("unzip", args = c("-o", file.path(workdir, fn),
                          "*full_maks.tiff",
                          "-d", workdir),
        stdout = TRUE)

# Load all masks
masks <- loadImages(file.path(workdir, "Basel_Zuri_masks"),
                    pattern = "_full_maks.tiff")

# Clean-up
tiffs_delete <- list.files(file.path(workdir, "Basel_Zuri_masks"))
file.remove(file.path(workdir, "Basel_Zuri_masks", tiffs_delete))
file.remove(file.path(workdir, fn))
```

### Rescale

The masks are 16-bit images and need to be re-scaled in order to obtain integer cell ids.

```{r scale-masks}
# Before scaling
range(masks[[1]])

masks <- scaleImages(masks, value = (2 ^ 16) - 1)

# After scaling
range(masks[[1]])
```

### Fix cell IDs

Occasionally a single-cell ID is skipped in the masks but in the single-cell data the cell numbers were renamed sequentially. Therefore, the single-cell IDs in the masks also have to renamed sequentially in order to correspond to the cell numbers from the single-cell data.

```{r fix-cell-ids}
# Rename single-cell IDs sequentially in each mask
for (n in names(masks)){
  imageData(masks[[n]]) = plyr::mapvalues(
    imageData(masks[[n]]),
    sort(unique(as.integer(imageData(masks[[n]])))),
    0:(length(unique(as.integer(imageData(masks[[n]]))))-1)
  )
}
```

### Add mask names

Next, we add image names to masks objects, these names correspond to the 
`image_name` column in `colData(sce)`. This information is stored in the 
metadata columns of the `CytoImageList` objects and is used by `cytomapper` to 
match single cell data, images and masks.

```{r add-mask-names}
names(masks) <- gsub("_full_maks", "_full.tiff", names(masks))
mcols(masks)$image_filename <- names(masks)
masks <- masks[names(masks) %in% sce$image_filename]

# Add image metadata
mcols(masks) <- merge(
  mcols(masks),
  unique(colData(sce)[, c(
      "image_filename", "image_name", "image_number")]),
  by = "image_filename")
```

### Create cohort-specific subsets

We create two objects containing the masks from the Basel and Zurich cohorts.

We subset the masks to the hundred images contained in the
`SingleCellExperiment` objects.

```{r subset-masks-cohorts}
# Basel cohort
masks_basel <- masks[grep("BaselTMA", names(masks))]
masks_basel <- masks_basel[names(masks_basel) %in% unique(
    sce_basel$image_filename)]

names(masks_basel) <- mcols(masks_basel)$image_name

# Zurich cohort
masks_zurich <- masks[grep("ZTMA", names(masks))]
masks_zurich <- masks_zurich[names(masks_zurich) %in% unique(
    sce_zurich$image_filename)]

names(masks_zurich) <- mcols(masks_zurich)$image_name

# All masks
names(masks) <- mcols(masks)$image_name
```

### Save on disk

Finally, we will save the generated `CytoImageList` objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-masks}
# All masks
saveRDS(masks, file.path(outdir, "masks_full.rds"))
print(head(masks))
remove(masks)

# Basel cohort
saveRDS(masks_basel, file.path(outdir, "masks_basel.rds"))
print(head(masks_basel))

# Zurich cohort
saveRDS(masks_zurich, file.path(outdir, "masks_zurich.rds"))
print(head(masks_zurich))
```


## Multichannel images

We will import and process multichannel images to make them compatible with the `cytomapper` package.

For memory space reasons, we will generate one image subset for the Basel cohort and one for the Zurich cohort. The images will be subsetted to correspond to the single cell data in the `sce_basel` and `sce_zurich` objects. We will first process images from the Basel cohort and then repeat the same steps for the Zurich cohort.

### Import images

We unzip images and read them into a `CytoImageList` object.

```{r unzip-images-basel, results='hide'}
fn <- file.path("OMEnMasks", "ome.zip")
system2("unzip", args = c("-o", file.path(workdir, fn),
                          "*BaselTMA_*.tiff",
                          "-d", workdir),
        stdout = TRUE)
```

Loading all tiffs would require too much memory, so we delete all the `tiff` files that we do not want to keep. We then use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r load-images-basel, results='hide', warning=FALSE}
tiffs_all <- list.files(file.path(workdir, "ome/"))
tiff_delete <- tiffs_all[!tiffs_all %in% sce_basel$image_filename]
file.remove(file.path(workdir, "ome/", tiff_delete))

# Load the images as a CytoImageList object
images_basel <- loadImages(file.path(workdir, "ome/"), pattern = "_full.tiff")
file.remove(file.path(workdir, "ome/", tiffs_all))
```

### Add image names and numbers

Next, we add image names to image objects, these names correspond to the 
`image_name` column in `colData(sce)`. This information is stored in the 
metadata columns of the `CytoImageList` objects and is used by `cytomapper` 
to match single cell data, images and masks.

```{r add-image-names-basel}
mcols(images_basel)$image_filename <- paste0(names(images_basel), ".tiff")
images_basel <- images_basel[
    mcols(images_basel)$image_filename %in% sce_basel$image_filename]
mcols(images_basel) <- merge(
  mcols(images_basel),
  unique(colData(sce_basel)[, c(
      "image_filename", "image_name", "image_number")]),
  by = "image_filename")
```

### Subset channels

Here, we exclude image channels that are not present in the `SingleCellExperiment` object.

```{r subset-channels-basel}
panel <- rowData(sce)
panel <- panel[order(panel$channel), ]

images_basel <- getChannels(images_basel, panel$channel)
gc()
```

### Add channel names

Finally, we will add protein short names as channel names of the `images` 
object. These short names correspond to the row names of the 
`SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names-basel}
channelNames(images_basel) <- rownames(panel)
```

### Check image names

We make sure that image names and masks names are the same so that they can be matched with `cytomapper`.

```{r check-mask-image-names-basel}
print(identical(mcols(masks_basel)$image_name,
                mcols(images_basel)$image_name))
names(images_basel) <- mcols(images_basel)$image_name
```

### Save on disk

Finally, we save the generated `CytoImageList` images for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-images-basel}
saveRDS(images_basel, file.path(outdir, "images_basel.rds"))
print(head(images_basel))

remove(images_basel)
gc()
```

### Zurich cohort

To import and format images from the Zurich cohort, we perform the exact same steps as we just did for the Basel cohort.

Unzip images

```{r unzip-images-zurich, results='hide'}
system2("unzip", args = c("-o", file.path(workdir, fn),
                          "*ZTMA*.tiff",
                          "-d", workdir),
        stdout = TRUE)
```

Loading all tiffs would require too much memory, so we delete all the `tiff` files that we do not want to keep. We then use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r load-images-zurich, results='hide', warning=FALSE}
tiffs_all <- list.files(file.path(workdir, "ome/"))
tiff_delete <- tiffs_all[!tiffs_all %in% sce_zurich$image_filename]
file.remove(file.path(workdir, "ome/", tiff_delete))

# Load the images as a CytoImageList object
images_zurich <- loadImages(file.path(workdir, "ome/"), pattern = "_full.tiff")
file.remove(file.path(workdir, "ome/", tiffs_all))
```

Next, we add image names to image objects, these names correspond to the 
`image_name` column in `colData(sce)`. This information is stored in the 
metadata columns of the `CytoImageList` objects and is used by `cytomapper` 
to match single cell data, images and masks.

```{r add-image-names-zurich}
mcols(images_zurich)$image_filename <- paste0(names(images_zurich), ".tiff")
images_zurich <- images_zurich[
    mcols(images_zurich)$image_filename %in% sce_zurich$image_filename]
mcols(images_zurich) <- merge(
  mcols(images_zurich),
  unique(colData(sce_zurich)[, c(
      "image_filename", "image_name", "image_number")]),
  by = "image_filename")
```

Here, we exclude image channels that are not present in the `SingleCellExperiment` object.

```{r subset-channels-zurich}
panel <- rowData(sce)
panel <- panel[order(panel$channel), ]

images_zurich <- getChannels(images_zurich, panel$channel)
gc()
```

We add protein short names as channel names of the `images` 
object. These short names correspond to the row names of the 
`SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names-zurich}
channelNames(images_zurich) <- rownames(panel)
```

We make sure that image names and masks names are the same so that they can be matched with `cytomapper`.

```{r check-mask-image-names-zurich}
print(identical(mcols(images_zurich)$image_name,
                mcols(masks_zurich)$image_name))
names(images_zurich) <- mcols(images_zurich)$image_name
```

Finally, we save the generated `CytoImageList` images and masks objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-images-zurich}
saveRDS(images_zurich, file.path(outdir, "images_zurich.rds"))
print(head(images_zurich))

remove(images_zurich)
```

## Clean up

Remove all files from the temporary working directory.

```{r clean-up, results='hide'}
downloaded_files <- list.files(workdir)
downloaded_files <- downloaded_files[!downloaded_files %in% "BiocStyle"]
unlink(file.path(workdir, downloaded_files), recursive = TRUE)

# Reset original timeout value
options(timeout = timeout)
```


# **Session information**

```{r session-info}
sessionInfo()
```


# **References**