
# **Single cell data**

# **Introduction**

This script downloads single-cell data, multichannel images and cell segmentation masks from a  
 Imaging Mass Cytometry (IMC) dataset investigating type 1 diabetes pathogenesis.
[Steenbuck, Damond et al. Imaging mass cytometry reveals functional and immunological changes during type 1 diabetes progression in human pancreata, Nature Metabolism, 2026](https://doi.org/10.1038/s42255-026-01559-z).  

In the study, we acquired one Islet panel and Immune panel.
IMC data is provided on **Zenodo** and spread across multiple repositories, see below for details: 
- SingleCellExperiment objects containing single-cell data and metadata + cell segmentation masks: https://zenodo.org/records/14968076
- Images: Islet: https://zenodo.org/records/15919088 + Immune: https://zenodo.org/records/15074971

# **Settings**

```{r libraries, include = FALSE}
library(BiocStyle)
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
library(ExperimentHub)
```

```{r dataset-version, echo=FALSE}
dataset_name <- "SteenbuckDamond_2026_Pancreas"
dataset_version <- "v1"
cat("Dataset version:", dataset_version)
```

Setting the working and output directories

```{r directories}
# Temporary directory to unzip files
# workdir <- tempdir()
# workdir <- setwd("/mnt/central_nas/projects/type1_diabetes/nathan/T1D_Vol")
workdir <- getwd()
Sys.setenv(workdir = workdir)

# Output directory
dataset_dir <- file.path(".", dataset_name)
if(!(dir.exists(dataset_dir))) dir.create(dataset_dir)

outdir <- file.path(dataset_dir, dataset_version)
if(!(dir.exists(outdir))) dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```

# **Single cell data**

We will download single-cell data corresponding to
[@SteenbuckDamond_2026_Pancreas] from 
[Zenodo](https://zenodo.org/records/14968076).

## Download SpatialExperimentobjects

Here, we download `SpatialExperiment` objects from zenodo that contain all single-cell, sample and patient metadata.
Per panel, we download one `SpatialExperiment` object. 

### Import function

Function to download and read RDS files.

```{r import-function}
importAndReadData <- function(url, output_dir, filename) {
  # Create full path
  file_path <- file.path(output_dir, filename)
  
  # Download the RDS file
  if (!file.exists(file_path)) {
    download.file(url, destfile = file_path, mode = "wb")
  }
  
  # Read and return the RDS object
  readRDS(file_path)
}
```

### Download SCEs

```{r download-sce}
# Download SCE.
url_sce_islet <- "https://zenodo.org/records/14968076/files/spe_Islet_full.rds"
url_sce_immune <- "https://zenodo.org/records/14968076/files/spe_Immune_full.rds"

url_sces <- c(url_sce_islet, url_sce_immune)
names(url_sces) <- c("Islet", "Immune")
sce_list <- purrr::imap(url_sces, function(url, name) importAndReadData(url, workdir, paste0("sce_", name, ".rds")))
```

Remove remaining Pancreatitis images.

```{r remove-pancreatitis}
sce_list[[1]] <- sce_list[[1]][, sce_list[[1]]$donor_type != "Pancreatitis"]
sce_list[[2]] <- sce_list[[2]][, sce_list[[2]]$donor_type != "Pancreatitis"]
```

## colData

The colData entry of the SingleCellExperiment object is a DataFrame that contains observation metadata; 
i.e., cells, slides, tissue, patients, …. 
It is recommended that all column names have a prefix that indicates the level of observation 
(e.g. cell_, slide_ , tissue_, patient_, tumor_).

The following columns are required:
- `image_name` and/or `image_number`: unique image (ablated ROI) name, respectively number. Should map to the image_name/image_number column(s) in the metadata of the images and masks objects.
- `cell_number`: integer representing cell numbers. Should map to the values of cell segmentation masks.
- `cell_id`: a unique cell identifier defined as {image_number _ cell_number} (e.g., 7_138).
- `cell_x` and `cell_y`: position of the cell centroid on the image. These columns are used as SpatialCoords when converting to a SpatialExperiment object.
- In addition, colnames(sce) should be set as colData(sce)$cell_id.

```{r format-col}
colData_list <- lapply(sce_list, function(sce) data.table::as.data.table(colData(sce)))
colData_list
```
 
### colData: Patient-level metadata

-> ADD patient_ prefix to all patient-level metadata columns.
```{r coldata-patient-level}
# Add patient_ prefix to patient-level metadata columns
patient_columns <- c("case_id", "batch", "donor_type", "case_number",
                     "number_of_AAbs", "GADA", "IA2A", "mIAA", "ZnT8A", 
                     "age", "diabetes_duration", "gender", "race", "BMI", "HbA1c", "C_peptide",
                     "age_group", "age_years", "height_cm", "weight_kg", 
                     "transit_time_minutes", "ICU_time_days", "cause_of_death", 
                     "case_recovery_type", "drug_use", "class_weight") 

# check that all patient-specific columns are present:
lapply(colData_list, function(dt) {
  missing_columns <- setdiff(patient_columns, colnames(dt))
  if(length(missing_columns) > 0) {
    warning(paste("Missing patient-level metadata columns in colData:", paste(missing_columns, collapse = ", ")))
  }
})

colData_list <- lapply(colData_list, function(dt) {
  data.table::setnames(dt, old = patient_columns, new = paste0("patient_", patient_columns))
})
```

### colData: cell-level metadata

Split into 4 categories:
- cell-level metadata columns that should be kept as is (e.g. cell_area)
- cell-level metadata columns that should be renamed with a cell_ prefix (e.g. pseudotime_beta -> cell_pseudotime_beta).
- cell-level metadata columns that should be added (cell_x + cell_y).
- cell-level metadata columns that should be removed from colData (e.g. Pheno1_fastMNN_case_id etc.).

```{r coldata-cell-level}
cell_columns_present <- c("cell_area", "cell_number", "cell_major_axis_length", "cell_minor_axis_length", 
                          "cell_eccentricity", "cell_id", "cell_type", "cell_category")

cell_columns_to_rename <- c("pseudotime_beta", "distance_to_islet", "periislet", "islet_closest", "islet_parent")

cell_columns_to_add <- c("cell_x", "cell_y")

cell_column_to_remove <-c("...1", "Pheno1_fastMNN_case_id", "Leiden1_fastMNN_case_id", "Pheno1_harmony_batch",            
                          "Leiden1_harmony_batch", "Pheno1_scaled", "Leiden1_scaled", 
                          "CellType_Pheno1_scaled", "CellType_Pheno1_fastMNN_case_id", 
                          "CellType_Pheno1_harmony_batch", "CellType_Leiden1_scaled", 
                          "CellType_Leiden1_harmony_batch", "CellType_Leiden1_fastMNN_case_id",
                          "Leiden2_scaled", "Pheno2_scaled", 
                          "Leiden2_fastMNN_case_id", "Pheno2_fastMNN_case_id", 
                          "CellType_Pheno2_scaled", "CellType_Pheno2_fastMNN_case_id", 
                          "CellType_Leiden2_scaled", "CellType_Leiden2_fastMNN_case_id")


# Check if cell columns are present. 
# Note: distance_to_islet + periislet only in Immune panel and pseudotime_beta only in Islet panel.
lapply(colData_list, function(dt) {
  missing_columns <- setdiff(c(cell_columns_present, cell_columns_to_rename), colnames(dt))
  if(length(missing_columns) > 0) {
    warning(paste("Missing cell-level metadata columns in colData:", paste(missing_columns, collapse = ", ")))
  }
})

# RENAME: Add cell_ prefix to specified columns
colData_list <- lapply(colData_list, function(dt) {
  data.table::setnames(dt, old = cell_columns_to_rename, new = paste0("cell_", cell_columns_to_rename), skip_absent = TRUE)
})

# ADD: write colPairs$cell_x and cell_y to colData.
colData_list <- lapply(seq_along(colData_list), function(x) {
  dt <- colData_list[[x]]
  if(!("cell_x" %in% colnames(dt)) && "cell_x" %in% colnames(SpatialExperiment::spatialCoords(sce_list[[x]]))) {
    dt[, cell_x := SpatialExperiment::spatialCoords(sce_list[[x]])[, "cell_x"]]
  }
  if(!("cell_y" %in% colnames(dt)) && "cell_y" %in% colnames(SpatialExperiment::spatialCoords(sce_list[[x]]))) {
    dt[, cell_y := SpatialExperiment::spatialCoords(sce_list[[x]])[, "cell_y"]]
  }
  return(dt)
})

# REMOVE: columns that should not be included in colData
# NOTE: these partially do not exist in colData_list[["Immune"]].
colData_list <- lapply(colData_list, function(dt) {
  dt[, (cell_column_to_remove) := NULL]
  return(dt)
})
```

### colData: image-level metadata

- as-is columns:
- rename columns.

```{r coldata-image-level}
image_columns_present <- c("image_x", "image_y", "image_number", "image_fullname", "image_id")
image_columns_to_rename <- c("image_width_px" = "width_px", "image_height_px" = "height_px", "image_name" = "sample_id")

# Check if image columns are present
lapply(colData_list, function(dt) {
  missing_columns <- setdiff(c(image_columns_present, image_columns_to_rename), colnames(dt))
  if(length(missing_columns) > 0) {
    warning(paste("Missing image-level metadata columns in colData:", paste(missing_columns, collapse = ", ")))
  }
})

# Rename specified columns
colData_list <- lapply(colData_list, function(dt) {
  data.table::setnames(dt, old = image_columns_to_rename, new = names(image_columns_to_rename))
})
```

Others + Islet:
- islet_area
- islet_y
- islet_x
- islet_eccentricity
- islet_major_axis_length
- islet_minor_axis_length
- islet_id
- panel
-> Leave as is.

### Final check:

Check that required columns are present in colData and that colnames(sce) is set as colData(sce)$cell_id.
```{r image-names-colData}
# Check that all required columns are present in colData
required_columns <- c("image_name", "cell_number", "cell_id", "cell_x", "cell_y")
# check for colData_list
lapply(colData_list, function(dt) {
  missing_columns <- setdiff(required_columns, colnames(dt))
  if(length(missing_columns) > 0) {
    stop(paste("Missing required columns in colData:", paste(missing_columns, collapse = ", ")))
  }
})

# Write the updated colData back to the SCE objects
for(i in seq_along(sce_list)) {
  colData(sce_list[[i]]) <- DataFrame(colData_list[[i]])
  # check that colnames(sce) is set as colData(sce)$cell_id
  if(!all(colnames(sce_list[[i]]) == colData(sce_list[[i]])$cell_id)) {
    stop("colnames(sce) must be set as colData(sce)$cell_id")
  }
}
```

### Add missing information:

#### ICI vs IDIs.
Select ICI/IDIs on Islet panel and write to both colDatas.
ICI = Insulin-containing Islets (= contains beta-cells).
IDI = Insulin-deficient Islet (= contains beta-cells).

```{r ici_idis}
ici_ids <- colData(sce_list[[1]]) |> 
  tibble::as_tibble() |> 
  dplyr::select(image_id, cell_type, patient_donor_type, islet_id, cell_islet_closest, cell_distance_to_islet) |>
  dplyr::filter(cell_type == "Beta") |> 
  dplyr::filter(cell_distance_to_islet >= 0) |>
  dplyr::distinct(image_id) |> 
  dplyr::pull(image_id)

immune_ici_ids <- gsub("Islet", "Immune", ici_ids)
sce_list[[1]]$image_ici_idi <- ifelse(sce_list[[1]]$image_id %in% ici_ids, "ICI", "IDI")
sce_list[[2]]$image_ici_idi <- ifelse(sce_list[[2]]$image_id %in% immune_ici_ids, "ICI", "IDI")
```

#### Insulitic ROIs.
Select Insulitic Images on Immune panel and write to both colDatas.

```{r insulitic-rois}
insulitic_ids <- colData(sce_list[[2]]) |> 
  tibble::as_tibble() |> 
  dplyr::select(image_id, cell_type, patient_donor_type, islet_id, cell_islet_closest, cell_distance_to_islet) |>
  dplyr::filter(cell_type %in% c("T_DN", "T_CD8", "T_CD4")) |> 
  dplyr::filter(cell_distance_to_islet >= -5) |>
  dplyr::count(patient_donor_type, image_id, cell_islet_closest) |> 
  dplyr::arrange(desc(n)) |> 
  dplyr::filter(n >= 6) |> 
  dplyr::distinct(image_id) |> 
  dplyr::pull(image_id)

(head(insulitic_ids)); (length(insulitic_ids))
islet_insulitic <- gsub("Immune", "Islet", insulitic_ids)

sce_list[[1]]$image_insulitic <- ifelse(sce_list[[1]]$image_id %in% islet_insulitic, "Insulitic", "NonInsulitic")
sce_list[[2]]$image_insulitic <- ifelse(sce_list[[2]]$image_id %in% insulitic_ids, "Insulitic", "NonInsulitic")
```

## colPairs

Neighborhood information, such as a list of cells that are localized next to each other, can be stored as a SelfHits object in the colPair slot of the SingleCellExperiment object.
-> Done.

```{r}
(colPairs(sce_list[[1]]))
(colPairs(sce_list[[2]]))
```

## rowData

The rowData entry of the SingleCellExperiment is a DataFrame that contains marker (protein, RNA, probe) information.

The following columns are required in the rowData entry:
- `channel`: a unique integer that maps to the channels of the associated multichannel images.
  - Rename 1:47
- `metal`: the metal isotope used for detection, formatted as { ChemicalSymbol IsotopeMass} (e.g., Y89, In115, Yb176, Bi209).
- `name`: marker name used in the publication that describes the dataset.
- `full_name`: full marker name.
- `short_name`: abbreviated marker name, preferably following the official UniProt nomenclature.

For the `full_name` and `short_name` columns, the following guidelines apply:
- In `short_name`, all dashes, dots and spaces should removed or replaced with underscores.

For post-translationally modified proteins:
- Prefix the `full_name` with the modification type (e.g., phospho-, methyl-) and suffix it with the modified aminoacids (e.g., [S28]).
- Prefix the `short_name` with an abbreviation of the modification type (e.g., p_, me_) 
and do not indicate the modified aminoacids, unless there is a possible confusion with another target in the dataset.

```{r}
rowData(sce_list[[1]])[, "channel"] <- 1:nrow(rowData(sce_list[[1]]))
# Edit PTMs and other marker names in rowData of the SCE objects.
rowData(sce_list[[1]])["ERN1", "short_name"] <- "p_IRE1a"
rowData(sce_list[[1]])["ERN1", "name"] <- "p-IRE1-alpha"
rowData(sce_list[[1]])["HBP", "name"] <- "Hyaluronic-Acid"
rowData(sce_list[[1]])["PDL1", "name"] <- "PDL1"
rowData(sce_list[[1]])[, "full_name"] <- rowData(sce_list[[1]])[, "name"]

# Full names
rowData(sce_list[[1]])["ERN1", "full_name"] <- "phospho-IRE1-alpha[S724]"

rowData(sce_list[[1]])["CFTR", "full_name"] <- "Cystic Fibrosis Transmembrane Conductance Regulator"
rowData(sce_list[[1]])["XBP1", "full_name"] <- "X-box Binding Protein 1"
rowData(sce_list[[1]])["B2M", "full_name"] <- "Beta-2-Microglobulin"
rowData(sce_list[[1]])["PCSK1", "full_name"] <- "Proprotein Convertase Subtilisin/Kexin Type 1"
rowData(sce_list[[1]])["ProINS", "full_name"] <- "Proinsulin (Insulin precursor)"
rowData(sce_list[[1]])["ProGCG", "full_name"] <- "Proglucagon (Glucagon precursor)"
rowData(sce_list[[1]])["TXNIP", "full_name"] <- "Thioredoxin-Interacting Protein"
rowData(sce_list[[1]])["WFS1", "full_name"] <- "Wolframin ER Transmembrane Glycoprotein"
rowData(sce_list[[1]])["PDL1", "full_name"] <- "Programmed death-ligand 1"
rowData(sce_list[[1]])["NFkB", "full_name"] <- "Nuclear factor kappa-light-chain-enhancer of activated B-cells"
rowData(sce_list[[1]])["NKX2_2", "full_name"] <- "Homeobox protein Nkx-2.2"
rowData(sce_list[[1]])["ARX", "full_name"] <- "Aristaless Related Homeobox"
rowData(sce_list[[1]])["CD9", "full_name"] <- "CD9 (Tetraspanin)"
rowData(sce_list[[1]])["CAIX", "full_name"] <- "Carbonic Anhydrase IX"
rowData(sce_list[[1]])["IAPP", "full_name"] <- "Islet Amyloid Polypeptide (Amylin)"
rowData(sce_list[[1]])["ATPIF1", "full_name"] <- "ATPase Inhibitory Factor 1"
rowData(sce_list[[1]])["NKX6_1", "full_name"] <- "Homeobox protein Nkx-6.1"
rowData(sce_list[[1]])["PAX6", "full_name"] <- "Paired box protein Pax-6"
rowData(sce_list[[1]])["CD155", "full_name"] <- "Poliovirus Receptor (PVR)"
rowData(sce_list[[1]])["MX1", "full_name"] <- "Interferon-induced GTP-binding protein Mx1"
rowData(sce_list[[1]])["HLA_ABC", "full_name"] <- "Human Leukocyte Antigen Class I (HLA-A, HLA-B, HLA-C)"
rowData(sce_list[[1]])["ADAR", "full_name"] <- "Adenosine Deaminase Acting on RNA 1-A"
rowData(sce_list[[1]])["PCSK2", "full_name"] <- "Proprotein Convertase Subtilisin/Kexin Type 2"
rowData(sce_list[[1]])["PPY", "full_name"] <- "Pancreatic Polypeptide"
rowData(sce_list[[1]])["DNA1", "full_name"] <- "Ir191"
rowData(sce_list[[1]])["DNA3", "full_name"] <- "Ir193"

# For Immune panel:
rowData(sce_list[[2]])[, "channel"] <- 1:nrow(rowData(sce_list[[2]]))
rowData(sce_list[[2]])["CD3e", "name"] <- "CD3e"
rowData(sce_list[[2]])["PD1", "name"] <- "PD1"
rowData(sce_list[[2]])[, "full_name"] <- rowData(sce_list[[2]])[, "name"]

# Full names. 
rowData(sce_list[[2]])["MPO", "full_name"] <- "Myeloperoxidase"
rowData(sce_list[[2]])["SMA", "full_name"] <- "Alpha-Smooth Muscle Actin (ACTA2)"
rowData(sce_list[[2]])["CD16", "full_name"] <- "Fc Gamma Receptor III (FCGR3A / FCGR3B)"
rowData(sce_list[[2]])["CD11b", "full_name"] <- "Integrin Alpha M (ITGAM, part of Mac-1)"
rowData(sce_list[[2]])["HLA_DR", "full_name"] <- "Major Histocompatibility Complex Class II, DR"
rowData(sce_list[[2]])["CD20", "full_name"] <- "B-Lymphocyte Antigen CD20 (MS4A1)"
rowData(sce_list[[2]])["CD45RA", "full_name"] <- "Protein Tyrosine Phosphatase Receptor Type C Isoform RA (PTPRC)"
rowData(sce_list[[2]])["CD56", "full_name"] <- "Neural Cell Adhesion Molecule 1 (NCAM1)"
rowData(sce_list[[2]])["CD54", "full_name"] <- "Intercellular Adhesion Molecule 1 (ICAM-1)"
rowData(sce_list[[2]])["CD11c", "full_name"] <- "Integrin Alpha X (ITGAX)"
rowData(sce_list[[2]])["ATP5A", "full_name"] <- "ATP Synthase F1 Subunit Alpha"
rowData(sce_list[[2]])["CD73", "full_name"] <- "Ecto-5-Nucleotidase (NT5E)"
rowData(sce_list[[2]])["GranB", "full_name"] <- "Granzyme B (GZMB)"
rowData(sce_list[[2]])["LDHA", "full_name"] <- "Lactate Dehydrogenase A"
rowData(sce_list[[2]])["TMEM173", "full_name"] <- "Transmembrane Protein 173 (STING1)"
rowData(sce_list[[2]])["CD45RO", "full_name"] <- "Protein Tyrosine Phosphatase Receptor Type C Isoform RO (PTPRC)"
rowData(sce_list[[2]])["HK1", "full_name"] <- "Hexokinase 1 (HK1)"
rowData(sce_list[[2]])["CD8a", "full_name"] <- "CD8 alpha (CD8A)"
rowData(sce_list[[2]])["FoxP3", "full_name"] <- "Forkhead Box Protein P3"
rowData(sce_list[[2]])["Arg1", "full_name"] <- "Arginase 1 (ARG1)"
rowData(sce_list[[2]])["CD103", "full_name"] <- "Integrin Alpha E (ITGAE)"
rowData(sce_list[[2]])["CD204", "full_name"] <- "Macrophage Scavenger Receptor 1 (MSR1)"
rowData(sce_list[[2]])["TIM3", "full_name"] <- "T-cell Immunoglobulin and Mucin-domain Containing-3 (HAVCR2)"
rowData(sce_list[[2]])["CD303", "full_name"] <- "C-type Lectin Domain Family 4 Member C (CLEC4C, BDCA-2)"
rowData(sce_list[[2]])["CD206", "full_name"] <- "Mannose Receptor C-Type 1 (MRC1)"
rowData(sce_list[[2]])["CD66b", "full_name"] <- "Carcinoembryonic Antigen-related Cell Adhesion Molecule 8 (CEACAM8)"
rowData(sce_list[[2]])["CD15", "full_name"] <- "Lewis X Antigen (SSEA-1, carbohydrate epitope)"
```


## Assays

The `assays` slot of the `SingleCellExperiment` contains counts matrices representing marker expression levels 
per cell and channel.

It should at least contain a counts matrix with `raw` ion counts. 
The assays slot can also contain additional matrices with commonly used counts transformations, 
or counts transformations that were used in the publication that describes the dataset. 
All counts transformations must be documented in the .R function used to load the dataset. 

Common examples include:
    `exprs`: asinh-transformed counts. For IMC, a cofactor of 1 is typically used.
    `quant_norm`: counts censored (e.g., at the 99th percentile) and scaled from 0 to 1. Here, named as `scaled`.

```{r}
assayNames(sce_list[[1]])
assayNames(sce_list[[2]])
```

## Subsample-Images.

### ADD missing colData-information.

Generate fine-grained approximate disease stages.

keep only: 
- Stage 1: Control, 
- Stage 2: sAAb+_NonInsulitic_ICI, 
- Stage 3: mAAb+_NonInsulitic_ICI, 
- Stage 4: Onset_NonInsulitic_ICI, 
- Stage 5: mAAb+_Insulitic_ICI, Onset_Insulitic_ICI, 
- Stage 6: mAAb+_Insulitic_IDI, Onset_Insulitic_IDI
- Stage 7: mAAb+_NonInsulitic_IDI, Onset_NonInsulitic_IDI
- Stage 8: LongDuration_NonInsulitic_IDI


```{r recode-stages}
tt <- colData(sce_list[[1]])[, c("image_name", "patient_donor_type", "image_insulitic", "image_ici_idi")] |> 
  tibble::as_tibble() |> 
  dplyr::distinct(image_name, patient_donor_type, image_insulitic, image_ici_idi)
  
tt <- tt |> 
  dplyr::mutate(stages = paste(patient_donor_type, image_insulitic, image_ici_idi, sep = "_"))  |> 
  dplyr::distinct(image_name, stages) |>
  dplyr::filter(stages %in% c("NoDiabetes_NonInsulitic_ICI", 
                              "sAAb+_NonInsulitic_ICI", 
                              "mAAb+_NonInsulitic_ICI", 
                              "RecentOnset_NonInsulitic_ICI", 
                              "mAAb+_Insulitic_ICI", "RecentOnset_Insulitic_ICI", 
                              "mAAb+_Insulitic_IDI", "RecentOnset_Insulitic_IDI",
                              "mAAb+_NonInsulitic_IDI", "RecentOnset_NonInsulitic_IDI",
                              "LongDuration_NonInsulitic_IDI")) |> 
  dplyr::mutate(stages = dplyr::case_when(
    stages %in% c("mAAb+_Insulitic_ICI", "RecentOnset_Insulitic_ICI") ~ "mAAb+/RecentOnset_Insulitic_ICI",
    stages %in% c("mAAb+_Insulitic_IDI", "RecentOnset_Insulitic_IDI") ~ "mAAb+/RecentOnset_Insulitic_IDI",
    stages %in% c("mAAb+_NonInsulitic_IDI", "RecentOnset_NonInsulitic_IDI") ~ "mAAb+/RecentOnset_NonInsulitic_IDI",
    TRUE ~ stages
  ))
tt |> dplyr::count(stages)
```


### Subssample SCEs.

```{r select-sce-list}
# From each stage select 15 images.
set.seed(222)
sub_images_islet <- tt |> 
  dplyr::group_by(stages) |> 
  dplyr::slice_sample(n = 15, replace = FALSE) |> 
  dplyr::ungroup()  |> 
  dplyr::pull(image_name)

sub_images_immune <- gsub("Islet", "Immune", sub_images_islet)

sce_list_sub <- lapply(sce_list, function(sce) sce[, sce$image_name %in% sub_images_islet | sce$image_name %in% sub_images_immune])
# set colnames()
colnames(sce_list_sub[[1]]) <- sce_list_sub[[1]]$cell_id
colnames(sce_list_sub[[2]]) <- sce_list_sub[[2]]$cell_id
```

### Retrieve Image-names

```{r image-names}
sce_list_sub[[1]]$image_name |> unique() -> image_names_islet
sce_list_sub[[2]]$image_name |> unique() -> image_names_immune

sce_list[[1]]$image_name |> unique() -> image_names_islet_full
sce_list[[2]]$image_name |> unique() -> image_names_immune_full
```

```{r}
# Coerce to SingleCellExperiment
sce_list[[1]] <- as(sce_list[[1]], "SingleCellExperiment")
sce_list[[2]] <- as(sce_list[[2]], "SingleCellExperiment")
```

## Add NICHE-information.

```{r add-niche-information}
run <- TRUE
if (run) {
  fn  <- "/mnt/central_nas/projects/type1_diabetes/nathan/T1D_Vol/temp_analysis/analysis/16_Neighborhood_cells_Immune/CN_annot.rds"
  tt <- readRDS(fn)

  cl_temp <- SummarizedExperiment::colData(sce_list[[2]]) |> 
    tibble::as_tibble() |> 
    dplyr::left_join(tt, by = "cell_id") |> 
    S4Vectors::DataFrame()
  
  SummarizedExperiment::colData(sce_list[[2]]) <- cl_temp
}
```

## Save.

We save the `SingleCellExperiment` object for upload to `ExperimentHub`.

```{r save-sces}
# Save the updated SCE objects as RDS files
names(sce_list) <- c("islet", "immune")
lapply(names(sce_list), function(name) {
    message(file.path(outdir, paste0("sce_full_", name, ".rds")))
    saveRDS(sce_list[[name]], file.path(outdir, paste0("sce_full_", name, ".rds")))
})
```

We also save the subsampled SCE objects as RDS files for upload to "ExperimentHub".
```{r save-sces-sub}  
sce_list_sub <- lapply(sce_list, function(sce) sce[, sce$image_name %in% sub_images_islet | sce$image_name %in% sub_images_immune])
# set colnames()
colnames(sce_list_sub[[1]]) <- sce_list_sub[[1]]$cell_id
colnames(sce_list_sub[[2]]) <- sce_list_sub[[2]]$cell_id

# Save the updated SCE objects as RDS files
names(sce_list_sub) <- c("islet", "immune")
lapply(names(sce_list_sub), function(name) {
  message(file.path(outdir, paste0("sce_", name, ".rds")))
  saveRDS(sce_list_sub[[name]], file.path(outdir, paste0("sce_", name, ".rds")))
})

message(colnames(colData(sce_list_sub[[2]])))
```


### Clean-Up:
```{r clean-up}
rm(sce_list, colData_list)
```


# Cell Masks:

Are located at https://zenodo.org/records/14968076/files/mask_cells.zip

## Download and unzip cell segmentation masks.
```{r download-cell-masks}
url_masks <- "https://zenodo.org/records/14968076/files/mask_cells.zip"
dest_dir <- file.path(workdir, "mask_cells.zip")
exdir <- file.path(workdir, "mask_cells")

if (!file.exists(dest_dir)) {
  download.file(url_masks, destfile = dest_dir, mode = "wb")
}
if (!dir.exists(exdir)) {
  unzip(dest_dir, exdir = exdir, overwrite = TRUE)
}
```

## Create a CytoImageList object: Masks.

Cell segmentation masks are stored in a CytoImageList object named masks.

The values of the masks should be integers mapping to the cell_number column of colData(sce). 
This information is used by cytomapper to associate `single cell data` and `cell segmentation masks`.

The metadata slot `(mcols(masks))` must contain an `image_name` column that maps to the image_name column of `colData(sce)`, 
and to the image_name column of `mcols(images)`. 

This information is used by cytomapper to associate:
- multichannel images
- cell segmentation masks
- `single cell data`.

```{r create-cytoimagelist-masks}
temp_dir <- tempdir()
workdir_masks <- file.path(workdir, "mask_cells", "mask_cells")
panels <- c("islet", "immune")

# Create a CytoImageList object for each panel
mask_list <- lapply(panels, function(panel) {
  # List all files in the directory for the current panel
  # panel <- "Islet"
  panel_dir <- file.path(workdir_masks, panel, "whole-cell")
  mask_files <- list.files(panel_dir, full.names = TRUE)

  # Create a CytoImageList object from the mask files
  tt <- cytomapper::loadImages(mask_files, pattern = "\\.tiff$", on_disk = FALSE)
  names(tt) <- gsub("_ROI_", "_", names(tt))
  return(tt)
})
```

Scale Masks to Integer Values.
```{r scale-masks}
image_names_full <- list("islet" = image_names_islet_full, "immune" = image_names_immune_full)

mask_list_2 <- lapply(seq_along(mask_list), function(i) {
  # i <- 2
  mask_sublist <- mask_list[[i]]
  mask_sublist <- cytomapper::getImages(mask_sublist, names(cytomapper::getImages(mask_sublist)) %in% image_names_full[[i]])
  # scale to integers. 
  mask_sublist <- cytomapper::scaleImages(mask_sublist, 2^16-1)
  mcols(mask_sublist)[["image_name"]] <- names(cytomapper::getImages(mask_sublist))
  return(mask_sublist)
})
```

```{r subset-masks}
## Generate SUBSET using the same image-ids as for SCE-SUB.
image_names_sub <- list("islet" = image_names_islet, "immune" = image_names_immune)

mask_list_sub <- lapply(seq_along(mask_list_2), function(i) {
  # i <- 2
  mask_sublist <- mask_list_2[[i]]
  mask_sublist <- cytomapper::getImages(mask_sublist, names(cytomapper::getImages(mask_sublist)) %in% image_names_sub[[i]])
  mcols(mask_sublist)[["image_name"]] <- names(cytomapper::getImages(mask_sublist))
  return(mask_sublist)
})
```

## Save the CytoImageList objects as RDS files for upload to `ExperimentHub`.
```{r save-masks}
## Save the CytoImageList objects as RDS files
panels <- c("islet", "immune")
lapply(seq_along(mask_list), function(i) {
    print(paste0("Saving masks_full_", panels[i], ".rds"))
    print(file.path(getwd(), outdir, paste0("masks_full_", panels[i], ".rds")))
    saveRDS(mask_list[[i]], file.path(outdir, paste0("masks_full_", panels[i], ".rds")))
})

## Save the subsetted CytoImageList objects as RDS files
lapply(seq_along(mask_list_sub), function(i) {
    print(paste0("Saving masks_", panels[i], ".rds"))
    print(file.path(getwd(), outdir, paste0("masks_",panels[i], ".rds")))
    saveRDS(mask_list_sub[[i]], file.path(outdir, paste0("masks_",panels[i], ".rds")))
})
```

## Clean-up:
```{r clean-up-masks}
rm(mask_list, mask_list_2)
```


# Images:
Processed images (.tiffs):

Data is available in the following repositories:

- Immune panel: https://zenodo.org/records/15971803
- Islet panel: https://zenodo.org/records/15919088


```{r download-images}
exdir <- file.path(workdir, "Islet_img")
dest_dir <- file.path(workdir, "Islet_img.zip")

if (!dir.exists(dest_dir)) {
  url_img <- "https://zenodo.org/records/15919088/files/Islet.zip"
  download.file(url_img, destfile = dest_dir, mode = "wb")
}
if (!dir.exists(exdir)) {
  unzip(dest_dir, exdir = exdir, overwrite = TRUE)
}

exdir <- file.path(workdir, "Immune_img")
dest_dir <- file.path(workdir, "Immune_img.zip")

if (!dir.exists(dest_dir)) {
  url_img <- "https://zenodo.org/records/15971803/files/Immune.zip"
  download.file(url_img, destfile = dest_dir, mode = "wb")
}
if (!dir.exists(exdir)) {
  unzip(dest_dir, exdir = exdir, overwrite = TRUE)
}
```

```{r read-in-image-subset}
temp_dir <- tempdir()
panels <- c("islet", "immune")
image_names <- list("islet" = image_names_islet, "immune" = image_names_immune)

# Create a CytoImageList object for each panel
img_list <- purrr::map2(panels, image_names, function(panel, image_name) {
  # List all files in the directory for the current panel
  # panel <- "Islet"; image_name <- image_names_islet
  panel_dir <- file.path(workdir, paste(panel, "_img", sep = ""), panel)
  img_files <- list.files(panel_dir, full.names = TRUE)
  
  # add .tiff to end and add ROI after Case-ID and ROI tag (e.g. 1234_Islet -> 1234_Islet_ROI) to match the image_name in colData and in masks.
  image_name_2 <- paste0(image_name, ".tiff")
  image_name_2 <- gsub("_(Islet|Immune)", "_\\1_ROI", image_name_2)
  # add full path to image_name_2 (panel_dir)
  image_name_2 <- file.path(panel_dir, image_name_2)

  img_files <- img_files[img_files %in% image_name_2]

  # Create a CytoImageList object from the specified image files.
  tt <- cytomapper::loadImages(img_files, on_disk = FALSE)

  # remove _ROI_ from mask_list names. 
  names(tt) <- gsub("_ROI_", "_", names(tt))
  mcols(tt)[["image_name"]] <- names(cytomapper::getImages(tt))
  tt
})
```

Channel names of the images object (channelNames(images)) must map to rownames(sce) (marker short names).
We will adjust `channelNames` of the `CytoImageList` image subset to match the `SingleCellExperiment` object.

```{r channelnames}
# Update channelNames, so they match the SCE object
img_list <- purrr::map2(img_list, sce_list_sub, function(images_sub, sce_new_sub) {
  cytomapper::channelNames(images_sub) <- rownames(sce_new_sub)
  images_sub
})

img_list
```

## Save:

```{r save-images}
# Save the CytoImageList Image-Object as RDS files
lapply(seq_along(img_list), function(i) {
  print(paste0("Saving images_", panels[i], ".rds"))
  print(file.path(getwd(), outdir, paste0("images_", panels[i], ".rds")))
  saveRDS(img_list[[i]], file.path(outdir, paste0("images_", panels[i], ".rds")))
})
```