---
title: "Preparation of the Damond et al dataset"
author: "Nicolas Damond and Nils Eling"
date: "Created: 02 November 2020; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: console
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script downloads a hundred example images, as well as the associated single-cell data and cell segmentation masks from the pancreas Imaging Mass Cytometry (IMC) dataset described in the following publication:  

[Damond et al. A Map of Human Type 1 Diabetes Progression by Imaging Mass Cytometry. Cell Metabolism. 2019 Mar 5;29(3):755-768](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6821395).  

All data are openly available from [Mendeley data](http://dx.doi.org/10.17632/cydmwsfztj.2). The images and masks were created using the [imctools](https://github.com/BodenmillerGroup/imctools) package and the [IMC segmentation pipeline](https://github.com/BodenmillerGroup/ImcSegmentationPipeline).

Here, we will download single cell data and metadata, and process them to create a [SingleCellExperiment](https://bioconductor.org/packages//SingleCellExperiment.html) object. We will then download the corresponding multichannel IMC images and cell segmentation masks and format them into `CytoImageList` objects using the [cytomapper](https://bioconductor.org/packages/cytomapper) package.


# **Settings**

```{r libraries, include = FALSE}
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
cat("Dataset version: v1")
```

Set the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
outdir <- file.path(".", "Damond_2019_Pancreas")
if(!(dir.exists(outdir)))
    dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


# **Single cell data**

We will download a subset of single-cell data corresponding to 100 images from [@Damond-2019-Pancreas] from [Mendeley data](http://dx.doi.org/10.17632/cydmwsfztj.2).

## Download single cell data

### Import function

Function to download and unzip files.

```{r import-function}
importData <- function(url, output_dir, filename) {
  # Download
  download.file(url, destfile = file.path(output_dir, filename))
  
  # Unzip
  system2("unzip", args = c("-o",
                            file.path(output_dir, filename),
                            "-d", output_dir),
          stdout = TRUE)
  
  # Remove zipped folder
  file.remove(file.path(output_dir, filename))
}
```

### Single cell data

The `CellSubset` file contains all single cell data and metadata, including marker expression levels, spatial information, and neighborhood information.

```{r import-cell-data, results='hide'}
url_cells <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/f1e3b8dc-56be-4172-bbc4-3a6f9de97563/file_downloaded")

importData(url_cells, workdir, "CellSubset.zip")
```

### Image metadata

The `Image` file contains all image metadata, such as image width and height, or the number of cells per image.

```{r import-image-metadata, results='hide'}
# Download the zipped folder image and unzip it
url_image_meta <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/0b236273-d21b-4566-84a2-f1c56324a900/file_downloaded")

importData(url_image_meta, workdir, "Image.zip")
```

### Cell type information

In the original publication, cells were phenotyped based on informative marker expression.
We also import these phenotype labels from the online repository.

```{r import-celltypes, results='hide'}
url_celltypes <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/59e8da72-5bfe-4289-b95b-28348a6e1222/file_downloaded")

importData(url_celltypes, workdir, "CellTypes.zip")
```

### Neighbors information

The `Object relationship` file contains information about cell neighborhoods.

```{r import-neighbors, results='hide'}
url_neigbhors <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/ecd72f68-4cf5-4121-9d3d-ebb3c608220f/file_downloaded")

importData(url_neigbhors, workdir, "ObjectRelationships.zip")
```

### Clinical information

Last, the `Donors` file contains clinical information about organ donors.

```{r download-donor-info, results='hide'}
url_donors <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/9074990e-1b93-4c79-8c49-1db01a66398b/file_downloaded")

importData(url_donors, workdir, "Donors.zip")
```


## Read in single-cell data

We  first read in the `.csv` file containing cell data and metadata, and order it by image and object (cell) number. Then, we read in the other files that contain image metadata, cell types, neighboring cells, and clinical data.

```{r read-in-cell-data}
# Single cell data and metadata
cells <- fread(file.path(workdir, "CellSubset.csv"), stringsAsFactors = FALSE)
cells <- cells[order(cells$ImageNumber, cells$ObjectNumber), ]

# Image metadata
image_metadata <- fread(file.path(workdir, "All_Image.csv"),
                        stringsAsFactors = FALSE)

# Cell types
celltypes <- fread(file.path(workdir, "CellTypes.csv"),
                   stringsAsFactors = FALSE)

# Neighbors
neighbors <- fread(file.path(workdir, "All_Object relationships.csv"),
                   stringsAsFactors = FALSE)

# Clinical data
donors <- fread(file.path(workdir, "Donors.csv"),
                stringsAsFactors = FALSE)
```


## Prepare data

### Cell-level metadata

Here, we will collect all cell, image and clinical metadata to generate the `colData` entry of the final `SingleCellExperiment` object.  

First, we collect all cell metadata. Columns are renamed for consistency with the other datasets.

```{r cell-metadata}
cell_metadata <- DataFrame(
  cell_number = cells$ObjectNumber,
  image_number = cells$ImageNumber,
  cell_x = cells$Location_Center_X,
  cell_y = cells$Location_Center_Y,
  cell_area = cells$AreaShape_Area,
  cell_perimeter = cells$AreaShape_Perimeter,
  cell_compactness = cells$AreaShape_Compactness,
  cell_eccentricity = cells$AreaShape_Eccentricity,
  cell_euler_number = cells$AreaShape_EulerNumber,
  cell_extent = cells$AreaShape_Extent,
  cell_major_axis_length = cells$AreaShape_MajorAxisLength,
  cell_minor_axis_length = cells$AreaShape_MinorAxisLength,
  cell_orientation = cells$AreaShape_Orientation,
  cell_solidity = cells$AreaShape_Solidity,
  neighbors_number = cells$Neighbors_NumberOfNeighbors_3,
  neighbors_percent_touching = cells$Neighbors_PercentTouching_3,
  islet_parent = cells$Parent_Islets,
  islet_closest = cells$Parent_ExpandedIslets,
  distance_to_islet = cells$Intensity_MedianIntensity_IsletDistance_c100,
  distance_to_bloodvessel = cells$Intensity_MedianIntensity_BVDistance_c101
)
cell_metadata$cell_number_absolute <- 1:nrow(cell_metadata)
```

We do the same with image metadata.

```{r image-metadata}
image_metadata <- image_metadata[, .(
  image_number = ImageNumber,
  image_name = Metadata_Core,
  image_filename = FileName_CleanStack,
  image_width = Width_CleanStack,
  image_height = Height_CleanStack,
  image_area = Width_CleanStack * Height_CleanStack,
  image_cells_per_image = Count_Cells,
  image_islets_per_image = Count_Islets,
  tissue_slide = Metadata_Slide
)]
```

We merge the cell and image metadata data frames.

```{r merge-metadata}
cell_metadata <- merge(cell_metadata, image_metadata, by = "image_number")
```

We add cell-type information to the metadata object. For this, we create a unique `cell_id`, which is combination of `image_name` and `cell_number`.

```{r import-cell-types}
# Add unique cell ids to cell metadata
cell_metadata$cell_id <- paste(cell_metadata$image_name,
                               cell_metadata$cell_number,
                               sep = "_")

# Merge cell metadata and cell type information
celltypes <- celltypes[, .(cell_id = id,
                           cell_type = CellType,
                           cell_category = CellCat)]
cell_metadata <- merge(cell_metadata, celltypes, by = "cell_id")
```

We add clinical (organ donor) information to the metadata object.

```{r import-donors-info}
# Rename columns
donors <- donors[, .(
  tissue_slide = slide,
  tissue_region = part,
  patient_id = case,
  patient_batch = group,
  patient_stage = stage,
  patient_disease_duration = duration,
  patient_age = Age,
  patient_gender = Gender,
  patient_ethnicity = Ethnicity,
  patient_BMI = BMI
)]

# Merge
cell_metadata <- merge(cell_metadata, donors, by = "tissue_slide")
```

We re-order the columns for consistency.

```{r reorder-columns}
col_order <- c(
  "cell_id", "image_name", "image_number", "cell_number",
  "cell_type", "cell_category", "cell_x", "cell_y", "cell_area",
  "cell_number_absolute", "neighbors_number", "islet_parent", "islet_closest",
  "distance_to_islet", "distance_to_bloodvessel",
  "image_width", "image_height", "image_filename",
  "tissue_slide", "tissue_region",
  colnames(cell_metadata)[grepl("patient_", colnames(cell_metadata))]
)

cell_metadata <- cell_metadata[, col_order]
```

Finally, we order the cell metadata object based on `image_number` and `cell_number` and add unique cell ids as row names.

```{r cell-rownames}
# Rows are ordered by image and cell numbers
cell_metadata <- cell_metadata[order(cell_metadata$image_number,
                                     cell_metadata$cell_number), ]

# Cell ids are used as row names
rownames(cell_metadata) <- cell_metadata$cell_id
```

### Neighbors

Here, we collect all cell neighborhood relationships. This information will be added to the `colPairs` slot of the `SingleCellExperiment` object. In the original publication, neighboring cells were defined by mask expansion.    

First, we subset object relationships to keep only cell neighborhood relationships for images in the current dataset. Then, we add a `cell_number_absolute` column that contains a unique interger per cell. This number will be used to generate cell pairings.

```{r prepare-neighbors}
# Keep only neighborhood relationships
neighbors <- neighbors[Relationship == "Neighbors", ]

# Subset to the 100 images in the dataset
setnames(neighbors, "First Image Number", "image_number")
neighbors <- neighbors[image_number %in% cell_metadata$image_number, ]

# Add image names
image_map <- as.data.frame(unique(
  cell_metadata[, c("image_number", "image_name")]))
neighbors <- merge.data.table(neighbors, image_map, by = "image_number")

# Add unique cell ids
neighbors[, cell_id_from := paste(image_name, `First Object Number`, sep = "_")]
neighbors[, cell_id_to := paste(image_name, `Second Object Number`, sep = "_")]

# Map to absolute cell numbers (one unique number per cell)
cell_map <- as.data.frame(unique(
  cell_metadata[, c("cell_id", "cell_number_absolute")]))

neighbors$cell_from <- cell_map$cell_number_absolute[
  match(neighbors$cell_id_from, cell_map$cell_id)]

neighbors$cell_to <- cell_map$cell_number_absolute[
  match(neighbors$cell_id_to, cell_map$cell_id)]
```

### Marker metadata

Here, we will collect all marker-related information and collect it in a `DataFrame` that will constitute the `rowData` slot of the `SingleCellExperiment` object.

We first download the panel file, which contains antibody-related metadata. For some datasets, however, the channel-order and the panel order do not match. For this reason, the channel-mass file is used to match panel information and image stack slices.

```{r import-panel}
# Import panel
url_panel <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/2f9fecfc-b98f-4937-bc38-ae1b959bd74d/file_downloaded")
download.file(url_panel, destfile = file.path(workdir, "panel.csv"))
panel <- fread(file.path(workdir, "panel.csv"))

# Import channel-mass file
url_channelmass <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/704312eb-377c-42e2-8227-44bb9aca0fb3/file_downloaded")
download.file(url_channelmass, destfile = file.path(workdir, "ChannelMass.csv"))
channel_mass <- fread(file.path(workdir, "ChannelMass.csv"), header = FALSE)
```

Then, we subset the channels that are relevant to data analysis (defined by the `keep` column in the panel file). Then, we order these channels based on isotope mass. Columns are renamed for consistency with the other datasets.

```{r load-panel}
# Read-in the panel
panel <- panel[, .(
  channel,
  metal = MetalTag,
  name = Target,
  short_name = clean_Target,
  antibody_clone = Clone,
  keep = full
)]

# Match panel and stack slice information
panel <- panel[keep == 1,]
panel <- panel[order(match(panel$metal, channel_mass$V1)), ]
panel[, keep := NULL]
```

We will also rename markers for consistency with other datasets.

```{r rename-markers}
panel[metal == "In115", `:=` (name = "Smooth muscle actin")]
panel[metal == "Pr141", `:=` (name = "Insulin")]
panel[metal == "Nd144", `:=` (name = "Prohormone convertase 2",
                              antibody_clone = "polyclonal_PC2")]
panel[metal == "Sm147", `:=` (name = "Myeloperoxidase",
                              antibody_clone = "polyclonal_MPO")]
panel[metal == "Nd148", `:=` (name = "Glucose transporter 1")]
panel[metal == "Nd150", `:=` (name = "Pancreatic amylase",
                              antibody_clone = "polyclonal_pancreatic_amylase")]
panel[metal == "Eu153", `:=` (name = "Pancreatic polypeptide")]
panel[metal == "Gd155", `:=` (name = "Programmed cell death protein 1",
                              short_name = "PD_1")]
panel[metal == "Gd158", `:=` (name = "Pancreatic and duodenal homeobox 1",
                              antibody_clone = "polyclonal_Pdx1")]
panel[metal == "Dy163", `:=` (name = "Forkhead box P3")]
panel[metal == "Ho165", `:=` (name = "CD8 alpha")]
panel[metal == "Er166", `:=` (name = "Carbonic anhydrase IX")]
panel[metal == "Er167", `:=` (name = "Islet amyloid polypeptide",
                              antibody_clone = "polyclonal_IAPP")]
panel[metal == "Er168", `:=` (name = "Ki-67", short_name = "Ki67")]
panel[metal == "Tm169", `:=` (name = "Homeobox protein Nkx-6.1",
                              short_name = "NKX6_1",
                              antibody_clone = "D8O4R")]
panel[metal == "Er170", `:=` (name = "p-Histone H3 [S28]",
                              short_name = "p_HH3")]
panel[metal == "Yb171", `:=` (antibody_clone = "polyclonal_CD4")]
panel[metal == "Yb173", `:=` (name = "E-Cadherin", short_name = "CDH1")]
panel[metal == "Yb174", `:=` (name = "IA-2", short_name = "PTPRN")]
panel[metal == "Lu175", `:=` (name = "phopsho-Rb [S807/S811]",
                              short_name = "p_Rb")]
panel[metal == "Yb176", `:=` (name = "cleaved-PARP + cleaved-Caspase3",
                              short_name = "cPARP_cCASP3")]
panel[metal == "Ir191", `:=` (name = "Ir191", short_name = "DNA1")]
panel[metal == "Ir193", `:=` (name = "Ir193", short_name = "DNA2")]
```

Finally, we convert the panel table to a `DataFrame` and add target short_names as row names.

```{r prepare-panel}
panel <- as(panel, "DataFrame")
rownames(panel) <- panel$short_name
```

### Counts matrix

Here, we will prepare the counts matrix that will be stored in the `assay` slot of the `SingleCellExperiment` object.  

CellProfiler measures a number of different statistics per marker and cell. We select the mean intensity per channel and per cell to obtain single-cell expression counts.

```{r select-counts}
counts_columns <- grepl("Intensity_MeanIntensity_CleanStack", colnames(cells))
counts <- cells[, ..counts_columns]
```

Finally, we reorder the channels based on channel number and convert the counts to a `matrix`.

```{r reorder-counts}
channel_number <- as.numeric(sub("^.*_c", "", colnames(counts)))
column_order <- order(channel_number, decreasing = FALSE)
counts <- counts[, ..column_order]
colnames(counts) <- NULL
counts <- as.matrix(counts, rownames = NULL)
```


## Create SingleCellExperiment object

### Create the object

We have now obtained all data and metadata required to create the `SingleCellExperiment` object.

```{r create-SCE}
sce <- SingleCellExperiment(
  assays = list(counts = t(counts)),
  rowData = panel,
  colData = cell_metadata
)
mainExpName(sce) <- "Damond_2019_Pancreas"
```

### Counts transformations

We apply two different counts transformations:
- `exprs`: arcsinh-transformed counts (cofactor = 1).
- `quant_norm`: censored + quantile-normalized counts.

```{r transform-counts}
assay(sce, "exprs") <- asinh(counts(sce) / 1)

quant <- apply(assay(sce, "counts"), 1, quantile, probs = 0.99)
assay(sce, "quant_norm") <- apply(assay(sce, "counts"), 2,
                                  function(x) x / quant)
assay(sce, "quant_norm")[assay(sce, "quant_norm") > 1] <- 1
assay(sce, "quant_norm")[assay(sce, "quant_norm") < 0] <- 0
```

### Add neighborhood information

We generate a [SelfHits](https://bioconductor.org/packages/devel/bioc/vignettes/SingleCellExperiment/inst/doc/intro.html#6_Storing_row_or_column_pairings) object containing pairings of neighboring cells and store it in the `colPairs` slot of the `SingleCellExperiment` object.

Integers in the `colPairs(sce, "neighborhood")` object are unique cell numbers that map to `colData(sce)$cell_number_absolute`.

```{r add-colpairs}
colPair(sce, "neighborhood") <- SelfHits(from = neighbors$cell_from,
                                         to = neighbors$cell_to,
                                         nnode = ncol(sce))
```

### Save on disk

We save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
saveRDS(sce, file.path(outdir, "sce.rds"))
```

### Clean up

Finally, we remove the downloaded files and generated objects to save storage space.

```{r clean-up-cell-data, results='hide'}
remove(counts, cells, celltypes, cell_metadata, neighbors, image_metadata)

file.remove(file.path(workdir, "All_Image.csv"),
            file.path(workdir, "CellSubset.csv"),
            file.path(workdir, "CellTypes.csv"),
            file.path(workdir, "All_Object relationships.csv"),
            file.path(workdir, "Donors.csv"))
```


# **Images and cell masks**

Here, we will download a subset of a hundred images from [@Damond-2019-Pancreas], as well as the corresponding cell segmentation masks. Images and masks correspond to the data in the `SingleCellExperiment` object and will be formatted into `CytoImageList` objects.


## Import images and masks

### Multichannel images

We first download the image subset.

```{r import-images, results='hide'}
url_images <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/b37054d2-d5d0-4c48-a001-81ff77136f41/file_downloaded")

importData(url_images, workdir, "ImageSubset.zip")
```

We use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r read-in-images}
images <- loadImages(workdir, pattern = "_full_clean.tiff")
images
```

### Cell segmentation masks

We also download the associated cell segmentation masks and read them into a `CytoImageList` object.

```{r import-masks, results='hide'}
url_masks <- ("https://data.mendeley.com/public-files/datasets/cydmwsfztj/files/13679a61-e9b4-4820-9f09-a5bbc697647c/file_downloaded")

importData(url_masks, workdir, "Masks.zip")
```

```{r read-in-masks}
masks <- loadImages(workdir, pattern = "_full_mask.tiff")
masks
```

### Clean-up

We remove the downloaded image and mask `tiff` files to save storage space.

```{r clean-up-images-masks, results='hide'}
images_to_delete <- list.files(workdir, pattern = "_full_clean.tiff",
                               full.names = TRUE)
file.remove(images_to_delete)

# Remove masks
masks_to_delete <- list.files(workdir, pattern = "_full_mask.tiff",
                              full.names = TRUE)
file.remove(masks_to_delete)
```


## Prepare images and masks

We will now process the images and masks to make them compatible with the `cytomapper` package.

### Rescale masks

The masks are 16-bit images and need to be re-scaled in order to obtain integer cell ids.

```{r scale-masks}
# Before scaling
masks[[1]]

masks <- scaleImages(masks, value = (2 ^ 16) - 1)

# After scaling
masks[[1]]
```

### Add image names and numbers

Next, we add image names to the images and masks objects, these names correspond to the `image_name` column in `colData(sce)`. This information is stored in the metadata columns of the `CytoImageList` objects and is used by `cytomapper` to match single cell data, images and masks.

```{r add-image-names}
mcols(images)$image_name <- gsub("_a0_full_clean", "", names(images))
mcols(masks)$image_name <- gsub("_a0_full_mask", "", names(masks))
```

We downloaded the full set of segmentation masks, so we will subset them to retain only the masks corresponding to the image subset. As a sanity check, we will make sure that the `image_name` slots of the `masks` and `images` objects are identical.

```{r subset-masks}
masks <- masks[mcols(masks)$image_name %in% mcols(images)$image_name]
identical(mcols(images)$image_name, mcols(masks)$image_name)
```

We will also add `image_numbers` to the metadata columns of the `images` and `masks` objects.

```{r add-image-numbers}
ref_images <- unique(colData(sce)[, c("image_name", "image_number")])
mcols(images) <- merge(mcols(images), ref_images, by = "image_name")
mcols(masks) <- merge(mcols(masks), ref_images, by = "image_name")
identical(mcols(images)$image_number, mcols(masks)$image_number)
```

### Add channel names

Finally, we will add protein short names as channel names of the `images` object with , corresponding to the row names of the `SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names}
# Match panel and stack slice information
panel <- rowData(sce)
panel <- panel[order(match(panel$metal, channel_mass$V1)), ]

# Add channel names to the "images" object
channelNames(images) <- panel$short_name
```

### Save on disk

Finally, we will save the generated `CytoImageList` images and masks objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-masks}
saveRDS(masks, file.path(outdir, "masks.rds"))
```

```{r save-images}
saveRDS(images, file.path(outdir, "images.rds"))
```

### Clean up

Remove all files from the temporary working directory.

```{r clean-up, results='hide'}
downloaded_files <- list.files(workdir)
downloaded_files <- downloaded_files[!downloaded_files %in% "BiocStyle"]
unlink(file.path(workdir, downloaded_files), recursive = TRUE)

# Reset original timeout value
options(timeout = timeout)
```


# **Session information**

```{r session-info}
sessionInfo()
```


# **References**