---
title: "Preparation of the Zanotelli et al dataset"
author: "Nicolas Damond"
date: "Created: 06 April 2021; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: inline
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script downloads 517 images, as well as the associated single-cell data and cell segmentation masks from the Imaging Mass Cytometry (IMC) cell line spheroid dataset described in the following publication:  

[Zanotelli et al. A quantitative analysis of the interplay of environment, neighborhood, and cell state in 3D spheroids. Mol Syst Biol (2020)16:e9798](https://doi.org/10.15252/msb.20209798).  

All data are openly available from [zenodo](https://doi.org/10.5281/zenodo.4271910). The code used to generate this dataset is available from the [SpheroidPublication GitHub repository](https://github.com/BodenmillerGroup/SpheroidPublication). Additional data associated with the same publication is available from the zenodo parent repository: https://doi.org/10.5281/zenodo.4055781.  

Here, we will download processed single-cell data and metadata, and process them to create a [SingleCellExperiment](https://bioconductor.org/packages//SingleCellExperiment.html) object. We will then download the corresponding IMC images and cell segmentation masks and format them into `CytoImageList` objects using the [cytomapper](https://bioconductor.org/packages/cytomapper) package.


# **Settings**

```{r libraries, include = FALSE}
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
cat("Dataset version: v1")
```

Set the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
outdir <- file.path(".", "Zanotelli_2020_Spheroids")
if(!(dir.exists(outdir)))
    dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


## Download the dataset

We download the full cell lines dataset from [@Zanotelli-2020-Spheroids], including single cell data and images from [zenodo](https://doi.org/10.5281/zenodo.4271910).

```{r download-dataset, results='hide'}
url_dat <- ("https://zenodo.org/record/4271910/files/phys_analysis_export_v3.zip")

download.file(url_dat, destfile = file.path(workdir, "CellLines.zip"))
unzip(file.path(workdir, "CellLines.zip"), exdir = workdir)
file.remove(file.path(workdir, "CellLines.zip"))

# List unzipped files
writeLines(list.files(workdir))
```


# **Single-cell data**

## Read in single-cell data

We read in the following files:

* `cell_X.csv`:  mean count intensities per cell.
* `cell_obs.csv`: cell metadata.
* `cell_var.csv`: marker and antibody metadata. 
* `image_meta.csv`: image metadata. 
* `relations_cell_neighbors.csv`: neighborhood information. 

```{r read-in-data}
cell_X <- fread(file.path(workdir, "cell_X.csv"))
cell_meta <- fread(file.path(workdir, "cell_obs.csv"))
cell_var <- fread(file.path(workdir, "cell_var.csv"))
image_meta <- fread(file.path(workdir, "image_meta.csv"))
cell_neighbors <- fread(file.path(workdir, "relations_cell_neighbors.csv"))
```

Add row and column names to the data tables and count matrix.

```{r row-col-names}
# Set object_id as cell metadata row names
cell_meta <- cell_meta[, !duplicated(colnames(cell_meta)), with = FALSE]

# Make values in the "goodname" column as unique
# (for barcoding channels and GFP)
cell_var[goodname == "BC", goodname := paste("barcoding", metal, sep = "_")]
cell_var[goodname == "GFP", goodname := paste("GFP", metal, sep = "_")]

# Set row and column names for the count matrix
cell_X <- as.matrix(cell_X)
rownames(cell_X) <- cell_meta$object_id
colnames(cell_X) <- cell_var$measurement_id
```


## Prepare data

### Cell-level metadata

Here, we will collect all cell and image metadata to generate the `colData` entry of the final `SingleCellExperiment` object.  

We first extract spatial measurements such as cell area, distance to rim, or cell location, and add them to cell metadata. Units are `pixels` (with 1 px = 1 um), unless otherwise specified.

- `cell_x`, `cell_y`: object centroid position in image.  
- `cell_area` : area of the cell (units = um^2).  
- `distance_rim`: estimated distance to spheroid border.  
- `distance_sphere`: distance to spheroid section border.  
- `distance_other_sphere`: distance to other spheroid section in the same image.  
- `distance_background`: distance to background pixels.  

Columns are renamed for consistency with the other datasets.

```{r extract-spatial-measurements}
# Fix measurement type and names for "dist-sphere", "dist-other" and "dist-bg"
cell_var[goodname %in% c("dist-sphere", "dist-other", "dist-bg"),
         measurement_type := "Location"]
cell_var[goodname == "dist-sphere", measurement_name := "dist-sphere"]
cell_var[goodname == "dist-other", measurement_name := "dist-other"]
cell_var[goodname == "dist-bg", measurement_name := "dist-bg"]

# Extract relevant columns
spatial_meas <- cell_var[measurement_type != "Intensity", measurement_id]
spatial <- cell_X[, c(colnames(cell_X)  %in% spatial_meas)]

# Add column names
spatial_names <- cell_var[measurement_type != "Intensity", goodname]
colnames(spatial) <- spatial_names

# Add the data to cell observations
spatial <- DataFrame(spatial)
cell_meta <- DataFrame(cell_meta)
rownames(cell_meta) <- cell_meta$object_id
cell_meta <- merge(cell_meta, spatial, by = "row.names")

# Rename columns
cell_meta <- DataFrame(
  cell_number = cell_meta$object_number,
  image_number = cell_meta$image_id,
  cell_obj_id = cell_meta$object_id,
  cell_x = cell_meta$Center_X,
  cell_y = cell_meta$Center_Y,
  cell_area = cell_meta$Area,
  distance_rim = cell_meta$dist.rim,
  distance_sphere = cell_meta$dist.sphere,
  distance_other_sphere = cell_meta$dist.other,
  distance_background = cell_meta$dist.bg
)
```

We then read in image metadata. See the [original dataset description](https://zenodo.org/record/4271910#.YGWWHD8kz-j) for description of the columns content. Columns are renamed for consistency with the other datasets.

```{r import-image-meta}
# Select relevant columns
image_meta <- image_meta[, .(
  image_number,
  image_x = image_pos_x,
  image_y = image_pos_y,
  image_width = image_shape_w,
  image_height = image_shape_h,
  image_filename = image_stack_filename_FullStackComp,
  mask_filename = mask_filename_cell,
  cell_line = cellline,
  treatment_id = condition_id,
  treatment_name = condition_name,
  treatment_concentration = concentration,
  treatment_time_point = time_point,
  treatment_hasTelox = hastelox,
  plate_id,
  plate_well_name = well_name,
  slide_id,
  sampleblock_id,
  sampleblock_name,
  acquisition_id,
  site_id
)]
```

We then merge the cell and image metadata data frames.

```{r merge-metadata}
# Merge cell and image metadata
cell_meta <- merge(cell_meta, DataFrame(image_meta), by = "image_number")
cell_meta$cell_id <- paste(cell_meta$image_number,
                           cell_meta$cell_number, sep = "_")
cell_meta$image_name <- gsub("_cell.tiff", "", cell_meta$mask_filename)

# Add cell_ids as row names
rownames(cell_meta) <- cell_meta$cell_id

# Remove merged data frames
remove(spatial, image_meta)
```

We re-order the columns for consistency and order the table by `image_number` and `cell_number`.

```{r reorder-columns-rows}
# Order columns
col_order <- c(
  "cell_id", "image_name", "image_number", "cell_number", "cell_line",
  "cell_x", "cell_y", "cell_area", "cell_obj_id",
  "distance_rim", "distance_background", "distance_sphere",
  "distance_other_sphere", "treatment_name", "treatment_id",
  "treatment_concentration", "treatment_time_point", "treatment_hasTelox",
  "image_width", "image_height", "image_x", "image_y",
  "image_filename", "mask_filename", "acquisition_id", "site_id", "slide_id",
  "plate_id", "plate_well_name", "sampleblock_id", "sampleblock_name"
)

cell_meta <- cell_meta[, col_order]

# Order rows
cell_meta <- cell_meta[order(cell_meta$image_number, cell_meta$cell_number), ]

# Add unique integers as cell identifiers
cell_meta$cell_number_absolute <- 1:nrow(cell_meta)
```

### Neighbors

Here, we collect all cell neighborhood relationships. This information will be added to the `colPairs` slot of the `SingleCellExperiment` object.

We map the object ids in the `cell_neighbors` data frame to `cell_number_absolute` integers in the `cell_meta` data frame.

```{r prepare-neighbors}
# Subset cell neighbors to cells present in cell_meta
cell_neighbors <- cell_neighbors[
  cell_neighbors$object_id_cell %in% cell_meta$cell_obj_id &
  cell_neighbors$object_id_neighbor %in% cell_meta$cell_obj_id, ]

# Map to absolute cell numbers (one unique integer per cell)
cell_map <- as.data.frame(unique(
  cell_meta[, c("cell_obj_id", "cell_number_absolute")]))

cell_neighbors$cell_from <- cell_map$cell_number_absolute[
  match(cell_neighbors$object_id_cell, cell_map$cell_obj_id)]

cell_neighbors$cell_to <- cell_map$cell_number_absolute[
  match(cell_neighbors$object_id_neighbor, cell_map$cell_obj_id)]
```

### Marker metadata

Here, we will collect all marker-related information and collect it in a `DataFrame` that will be the `rowData` of the `SingleCellExperiment` object.

We first select the `Intensity` and `MeanIntensityComp` measurements, which means spillover-compensated mean marker intensity per cell. Columns are renamed for consistency with the other datasets.

```{r assemble-panel}
# Select intensity measurement columns
panel <- cell_var[measurement_type == "Intensity" &
                    measurement_name == "MeanIntensityComp", ]
panel <- panel[order(ref_plane_number), ]

# Prepare the data frame
panel <- panel[, .(
  channel = panel$ref_plane_number,
  metal = panel$metal,
  name = panel$goodname,
  short_name = panel$goodname,
  antibody_clone = panel$`Antibody Clone`,
  antibody_working = panel$working,
  antibody_cell_cycle = panel$is_cc,
  measurement_id = panel$measurement_id
)]
```

We also rename markers for consistency with other datasets.

```{r rename-markers}
barcoding_channels <- c(
  "Pd102", "Rh103", "Pd104", "Pd105", "Pd106", "Pd108", "Pd108", "Pd110",
  "In113", "In115", "Pt194", "Pt195", "Bi209", "Y89")
panel[metal %in% barcoding_channels, `:=` (short_name = paste0("BC_", metal),
                                           antibody_clone = NA)]
panel[metal == "Te125", `:=` (name = "Telox2 probe", antibody_clone = NA)]
panel[metal == "La139", `:=` (name = "methylated-Histone H3 [K27]",
                              short_name = "me_H3")]
panel[metal == "Pr141", `:=` (name = "Histone H3", short_name = "H3")]
panel[metal == "Nd142", `:=` (name = "Epidermal growth factor receptor")]
panel[metal == "Nd143", `:=` (name = "phospho-FAK [Y397]",
                              short_name = "p_FAK")]
panel[metal == "Nd144", `:=` (name = "phospho-MEK1/2 [S217/S221]",
                              short_name = "p_MEK1_2")]
panel[metal == "Nd145", `:=` (name = "phospho-MAPKAPK2 [T334]",
                              short_name = "p_MAPKAPK2")]
panel[metal == "Nd146", `:=` (name = "phospho-p70S6K [T389]",
                              short_name = "p_p70S6K")]
panel[metal == "Sm147", `:=` (name = "phospho-Aurora kinase B [T232]",
                              short_name = "p_AURKB")]
panel[metal == "Nd148", `:=` (name = "phospho-STAT1 [S727]",
                              short_name = "p_STAT1")]
panel[metal == "Sm149", `:=` (name = "phospho-p53 [S15]",
                              short_name = "p_p53")]
panel[metal == "Eu151", `:=` (name = "phospho-EGFR [Tyr1173]",
                              short_name = "p_EGFR")]
panel[metal == "Sm152", `:=` (name = "phospho-AMPK alpha [T172]",
                              short_name = "p_AMPKa")]
panel[metal == "Eu153", `:=` (name = "phospho-Histone H3 [S28]",
                              short_name = "p_H3")]
panel[metal == "Sm154", `:=` (name = "phospho-ERK1/2 [T202/Y204]",
                              short_name = "p_ERK1_2")]
panel[metal == "Gd155", `:=` (name = "phospho-HER2 [Y1196]",
                              short_name = "p_HER2")]
panel[metal == "Gd156", `:=` (name = "phospho-p38 [T180/Y182]",
                              short_name = "p_p38")]
panel[metal == "Gd158", `:=` (name = "phospho-GSK3 [S9]",
                              short_name = "p_GSK3")]
panel[metal == "Gd160", `:=` (short_name = "BIRC5")]
panel[metal == "Dy161", `:=` (name = "Cyclin B1",
                              short_name = "CCNB1")]
panel[metal == "Dy162", `:=` (short_name = "VIM")]
panel[metal == "Dy163", `:=` (name = "phospho-AKT [S473]",
                              short_name = "p_AKT")]
panel[metal == "Ho165", `:=` (name = "phospho-CDK1 [Y15]",
                              short_name = "p_CDK1")]
panel[metal == "Er166", `:=` (name = "Carbonic anhydrase IX",
                              short_name = "CA9")]
panel[metal == "Er168", `:=` (short_name = "Ki67")]
panel[metal == "Er170", `:=` (name = "phospho-JNK [T183/Y185]",
                              short_name = "p_JNK")]
panel[metal == "Yb171", `:=` (name = "phospho-S6 [S235/S236]",
                              short_name = "p_S6")]
panel[metal == "Yb172", `:=` (name = "cleaved-Caspase 3",
                              short_name = "c_CASP3")]
panel[metal == "Yb173", `:=` (name = "phospho-STAT3 [Y705]",
                              short_name = "p_STAT3")]
panel[metal == "Yb174", `:=` (short_name = "c_PARP")]
panel[metal == "Lu175", `:=` (name = "phopsho-Rb [S807/S811]",
                              short_name = "p_Rb")]
panel[metal == "Yb176", `:=` (name = "DYKDDDDK tag")]
panel[metal == "Ir193", `:=` (antibody_clone = NA)]
```

Finally, we convert the panel table to a `DataFrame` and add target short_names as row names.

```{r prepare-panel}
panel <- as(panel, "DataFrame")
rownames(panel) <- panel$short_name
```

### Counts matrix

Here, we will prepare the counts matrix that will be stored in the `assay` slot of the `SingleCellExperiment` object.  

**Select measurements**

The measurement matrix contains different cell-level measurements:  
- `MeanIntensityComp`: mean intensity per cell, spillover-compensated.  
- `NbMeanMeanIntensityComp`:  mean intensity of neighboring cells, spillover-compensated.  

We will not retain the other measurements in the final SCE object:  
- `MeanIntensity`:mean intensity measured on compensated images.   
- `MinIntensity`: min intensity measured on compensated images.  
- `MaxIntensity`: max intensity measured on compensated images.  
- `StdIntensity`: intensity std measured on compensated images.  

```{r select-measurements}
# Select intensity measurement columns in counts matrix
cell_var <- cell_var[measurement_name %in% c("MeanIntensityComp",
                                             "NbMeanMeanIntensityComp"), ]
cell_X <- cell_X[, colnames(cell_X)  %in% cell_var$measurement_id]
```

**Prepare count matrices**

Here, we create one count matrix for mean cell intensities (`MeanIntensityComp`) and one count matrix for mean intensities of neighboring cells (`NbMeanMeanIntensityComp`).

```{r extract-counts}
# Select the ids of measurements to subset
cell_meas <- cell_var[
  cell_var$measurement_name == "MeanIntensityComp", ]$measurement_id
neighb_meas <- cell_var[
  cell_var$measurement_name == "NbMeanMeanIntensityComp", ]$measurement_id

# Create the two matrices
counts_X <- cell_X[, colnames(cell_X) %in% cell_meas]
counts_neighb_X <- cell_X[, colnames(cell_X) %in% neighb_meas]

# Make panel for neighboring cell counts
panel_neighb <- cell_var[
  cell_var$measurement_name == "NbMeanMeanIntensityComp", ]
panel_neighb <- merge(panel_neighb, panel[, c("metal", "short_name")],
                      by = "metal")

# Make sure the panels and counts matrices are in the same order
counts_X <- counts_X[, order(match(colnames(counts_X), panel$measurement_id))]
counts_X <- counts_X[order(match(rownames(counts_X), cell_meta$object_id)), ]

counts_neighb_X <- counts_neighb_X[, order(match(colnames(counts_neighb_X),
                                                 panel_neighb$measurement_id))]
counts_neighb_X <- counts_neighb_X[order(match(rownames(counts_neighb_X),
                                               cell_meta$object_id)), ]

# Rename the rows and columns of the count matrices
colnames(counts_X) <- panel$short_name
rownames(counts_X) <- cell_meta$cell_id
colnames(counts_neighb_X) <- panel_neighb$short_name
rownames(counts_neighb_X) <- cell_meta$cell_id
cell_meta$object_id <- NULL

# Remove original counts matrix
remove(cell_X)
```


## Create SingleCellExperiment object

### Create the object

We have now obtained all data and metadata required to create the `SingleCellExperiment` object.

```{r create-SCE}
sce <- SingleCellExperiment(
  assays = list(counts = t(counts_X)),
  rowData = panel,
  colData = cell_meta
)
mainExpName(sce) <- "Zanotelli_2020_Spheroids"
```

### Counts transformations

We apply two different counts transformations:
- `exprs`: arcsinh-transformed counts (cofactor = 1).
- `quant_norm`: censored + quantile-normalized counts.

```{r transform-counts}
assay(sce, "exprs") <- asinh(counts(sce) / 1)

quant <- apply(assay(sce, "counts"), 1, quantile, probs = 0.99)
assay(sce, "quant_norm") <- apply(assay(sce, "counts"), 2,
                                  function(x) x / quant)
assay(sce, "quant_norm")[assay(sce, "quant_norm") > 1] <- 1
assay(sce, "quant_norm")[assay(sce, "quant_norm") < 0] <- 0
```

### Neighboring cells counts

We create a new `SingleCellExperiment` object containing counts for neighboring cells and add it to the `altExp` slot. The data can be retrieved using `altExp(sce, "neighboring_cells")`

```{r add-neighboring-cells-counts}
# Order the count matrix
counts_neighb_X <- counts_neighb_X[, order(match(colnames(counts_neighb_X),
                                                 rownames(panel)))]
sce_neighb <- SingleCellExperiment(
  assays = list(counts = t(counts_neighb_X)),
  rowData = panel,
  colData = cell_meta
)

# Re-order the SCE objects by image and cell number
sce <- sce[, order(sce$image_number, sce$cell_number)]
sce_neighb <- sce_neighb[, order(
  sce_neighb$image_number, sce_neighb$cell_number)]

altExp(sce, "neighboring_cells") <- sce_neighb
```

### Add neighborhood information

We generate a [SelfHits](https://bioconductor.org/packages/devel/bioc/vignettes/SingleCellExperiment/inst/doc/intro.html#6_Storing_row_or_column_pairings) object containing pairings of neighboring cells and store it in the `colPairs` slot of the `SingleCellExperiment` object.

Integers in the `colPairs(sce, "neighborhood")` object are unique cell numbers that map to `colData(sce)$cell_number_absolute`.

```{r add-colpairs}
colPair(sce, "neighborhood") <- SelfHits(from = cell_neighbors$cell_from,
                                         to = cell_neighbors$cell_to,
                                         nnode = ncol(sce))
```

### Save on disk

We save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
saveRDS(sce, file.path(outdir, "sce.rds"))
```

### Clean up

Finally, we remove the downloaded files and generated objects to save storage space.

```{r clean-up-cell-data, results='hide'}
remove(sce_neighb, cell_meta, cell_neighbors, cell_var, cell_map,
       counts_X, counts_neighb_X)
```


# **Images and cell masks**

## Import images and masks

### Multichannel images

We first remove images that are not the `SingleCellExperiment` object.

```{r select-images, results='hide'}
image_files <- list.files(file.path(workdir, "images"))
image_files <- image_files[!(image_files %in% sce$image_filename)]
unlink(file.path(workdir, "images", image_files), recursive = TRUE)
```

We use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r read-in-images}
images <- loadImages(file.path(workdir, "images"), pattern = "_comp.tiff")
```

### Cell segmentation masks

We also remove masks that are not the `SingleCellExperiment` object and read the remaining masks into a `CytoImageList` object.

```{r select-masks, results='hide'}
mask_files <- list.files(file.path(workdir, "masks"))
mask_files <- mask_files[!(mask_files %in% sce$mask_filename)]
unlink(file.path(workdir, "masks", mask_files), recursive = TRUE)
```

```{r read-in-masks}
masks <- loadImages(file.path(workdir, "masks"), pattern = "_cell.tiff")
```


## Prepare images and masks

We will now process the images and masks to make them compatible with the `cytomapper` package.

### Rescale

The masks are 16-bit images and need to be re-scaled in order to obtain integer cell ids.

```{r scale-masks}
# Before scaling
masks[[1]]

masks <- scaleImages(masks, value = (2 ^ 16) - 1)

# After scaling
masks[[1]]
```

Images are scaled in the same way.

```{r scale-images}
# Before scaling
images[[1]]

images <- scaleImages(images, value = (2 ^ 16) - 1)

# After scaling
images[[1]]
```

### Add image names and numbers

Next, we add image names to the images and masks objects, these names correspond to the `image_name` column in `colData(sce)`. This information is stored in the metadata columns of the `CytoImageList` objects and is used by `cytomapper` to match single cell data, images and masks.

```{r add-image-names}
mcols(images)$image_name <- gsub("_comp", "", names(images))
mcols(masks)$image_name <- gsub("_cell", "", names(masks))
identical(mcols(images)$image_name, mcols(masks)$image_name)
```

We will also add `image_numbers` to the metadata columns of the `images` and `masks` objects.

```{r add-image-numbers}
ref_images <- unique(colData(sce)[, c("image_name", "image_number")])
mcols(images) <- merge(mcols(images), ref_images, by = "image_name")
mcols(masks) <- merge(mcols(masks), ref_images, by = "image_name")
identical(mcols(images)$image_number, mcols(masks)$image_number)
```

### Add channel names

Finally, we will add protein short names as channel names of the `images` object with , corresponding to the row names of the `SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names}
channelNames(images) <- rowData(sce)$short_name
```

### Save on disk

Finally, we will save the generated `CytoImageList` images and masks objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r clean-up-memory, echo=FALSE, results='hide'}
gc()
```

```{r save-masks}
saveRDS(masks, file.path(outdir, "masks.rds"))
```

```{r save-images}
saveRDS(images, file.path(outdir, "images.rds"))
```

### Clean up

Remove all files from the temporary working directory.

```{r clean-up, results='hide'}
downloaded_files <- list.files(workdir)
downloaded_files <- downloaded_files[!downloaded_files %in% "BiocStyle"]
unlink(file.path(workdir, downloaded_files), recursive = TRUE)

# Reset original timeout value
options(timeout = timeout)
```


# **Session information**

```{r session-info}
sessionInfo()
```


# **References**