---
title: "Preparation of the IMMUcan example dataset"
author: "Nils Eling and Nicolas Damond"
date: "Created: 14 September 2022; Compiled: `r BiocStyle::doc_date()`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script loads example data acquired by the [IMMUcan](https://immucan.eu) consortium. The data was processed as part of the [IMC data analysis book](https://bodenmillergroup.github.io/IMCDataAnalysis/) and is hosted at [zenodo.org/record/7079294](https://zenodo.org/record/7079294).   


# **Settings**

Loading required R packages

```{r libraries, include = FALSE}
library(data.table)
library(dplyr)
library(SpatialExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
dataset_name <- "IMMUcan_2022_CancerExample"
dataset_version <- "v1"
cat("Dataset version:", dataset_version)
```

Setting the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
dataset_dir <- file.path(".", dataset_name)
if(!(dir.exists(dataset_dir))) dir.create(dataset_dir)

outdir <- file.path(dataset_dir, dataset_version)
if(!(dir.exists(outdir))) dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


# **Single cell data**

## Import single cell data

We will download the already processed data and convert the SpatialExperiment into a SingleCellExperiment object.

```{r download-cell-data}
download.file("https://zenodo.org/record/7079294/files/spe.rds", 
              destfile = file.path(workdir, "spe.rds"))
```

```{r convert-SPE-to-SCE}
spe <- readRDS(file.path(workdir, "spe.rds"))
sce <- as(spe, "SingleCellExperiment")
```

## Prepare data

Here, we modify the SingleCellExperiment object for consistency with other datasets.

### Cell-level metadata

```{r cell-metadata}
sce$image_name <- sce$sample_id
sce$cell_number <- sce$ObjectNumber

# Add image numbers and cell ids
image_names <- unique(colData(sce)$image_name)
image_numbers <- 1:length(image_names)
names(image_numbers) <- image_names
sce$image_number <- image_numbers[colData(sce)$image_name]

sce$cell_id <- paste(sce$image_number, sce$cell_number, sep = "_")

# Synchronise additional columns
sce$cell_type <- sce$celltype
sce$image_width <- sce$width_px
sce$image_height <- sce$height_px
 
sce$cell_area <- sce$area
sce$cell_major_axis_length <- sce$major_axis_length
sce$cell_minor_axis_length <- sce$minor_axis_length
sce$cell_eccentricity <- sce$eccentricity

# Add cell ids as row names
colnames(sce) <- colData(sce)$cell_id
```

Store cell coordinates in `colData(sce)`

```{r cell-coords}
colData(sce)$cell_x <- spatialCoords(spe)[, "Pos_X"]
colData(sce)$cell_y <- spatialCoords(spe)[, "Pos_Y"]
```

### Marker metadata

Here, we will make the marker metadata (`rowData(sce)`) consistent with other datasets.

First, add `short_name` and `full_name` columns and rename other columns.

```{r marker-metadata}
# Add short names, metal and channel numbers
rowData(sce)$short_name <- rowData(sce)$name
rowData(sce)$metal <- rowData(sce)$channel
rowData(sce)$channel <- 1:nrow(rowData(sce))

# Rename columns
renaming_vector <- c(
  channel = "channel",
  metal = "metal",
  short_name = "short_name",
  full_name = "Target",
  name = "name",
  marker_class = "marker_class",
  antibody_clone = "Antibody.Clone",
  antibody_tube_number = "Tube.Number",
  antibody_stock_conc = "Stock.Concentration",
  antibody_final_conc_or_dilution = "Final.Concentration...Dilution",
  antibody_ul_to_add = "uL.to.add",
  use_channel = "use_channel",
  used_for_clustering = "used_for_clustering",
  channel_name = "channel_name",
  keep = "keep",
  ilastik = "ilastik",
  deepcell = "deepcell"
)

panel <- rowData(sce) %>%
  as_tibble() %>%
  dplyr::rename(all_of(renaming_vector)) %>%
  as.data.table()

# Re-order columns
setcolorder(panel, names(renaming_vector))
```

We then modify marker short names and full names for consistency.

```{r rename-markers}
panel[metal == "Y89", `:=` (full_name = "Myeloperoxidase")]
panel[metal == "In113", `:=` (short_name = "H3")]
panel[metal == "In115", `:=` (full_name = "Smooth muscle actin")]
panel[metal == "Nd143", `:=` (full_name = "HLA_DR", short_name = "HLA_DR")]
panel[metal == "Nd144", `:=` (antibody_clone = "polyclonal_CD27",
                              antibody_ul_to_add = 1.5)]
panel[metal == "Nd148", `:=` (full_name = "Beta-2-microglobulin")]
panel[metal == "Eu151", `:=` (full_name = "Indoleamine 2,3-dioxygenase 1",
                              short_name = "IDO1")]
panel[metal == "Sm152", `:=` (full_name = "CD3 epsilon", short_name = "CD3e")]
panel[metal == "Eu153", `:=` (
  full_name = "Lymphocyte activation gene 3 protein",
  short_name = "LAG3",
  antibody_ul_to_add = 1.4)]
panel[metal == "Gd155", `:=` (
  full_name = "Programmed cell death protein 1",
  short_name = "PD_1")]
panel[metal == "Gd156", `:=` (
  full_name = "Platelet-derived growth factor receptor beta",
  short_name = "PDGFRB")]
panel[metal == "Tb159", `:=` (full_name = "Granzyme B", short_name = "GZMB")]
panel[metal == "Gd160", `:=` (full_name = "Programmed death-ligand 1",
                              short_name = "PD_L1")]
panel[metal == "Dy163", `:=` (full_name = "Forkhead box P3")]
panel[metal == "Dy164", `:=` (full_name = "Inducible T-cell costimulator",
                              short_name = "ICOS")]
panel[metal == "Ho165", `:=` (full_name = "CD8 alpha")]
panel[metal == "Er166", `:=` (full_name = "Carbonic anhydrase IX",
                              short_name = "CA9")]
panel[metal == "Er168", `:=` (full_name = "Ki-67", short_name = "Ki67")]
panel[metal == "Tm169", `:=` (
  full_name = "V-type Ig domain-containing suppressor of T-cell activation")]
panel[metal == "Yb173", `:=` (full_name = "E-Cadherin", short_name = "CDH1")]
panel[metal == "Yb174", `:=` (antibody_clone = "polyclonal_CD303")]
panel[metal == "Lu175", `:=` (full_name = "CD206")]
panel[metal == "Yb176", `:=` (full_name = "cleaved-PARP",
                              short_name = "c_PARP")]
panel[metal == "Ir191", `:=` (full_name = "Iridium 191", short_name = "DNA1",
                              antibody_clone = NA,
                              antibody_tube_number = NA)]
panel[metal == "Ir193", `:=` (full_name = "Iridium 193", short_name = "DNA2",
                              antibody_clone = NA,
                              antibody_tube_number = NA)]
```

Finally, we convert the panel table to a `DataFrame` and add target short_names as row names.

```{r prepare-panel}
panel <- as(panel, "DataFrame")
panel$antibody_ul_to_add <- as.numeric(panel$antibody_ul_to_add)
rownames(panel) <- panel$short_name
rowData(sce) <- panel
```

## SingleCellExperiment object

### Finalize the object

```{r finalize-sce}
mainExpName(sce) <- paste(dataset_name, dataset_version, sep = "_")
```

### Save on disk

We save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
saveRDS(sce, file.path(outdir, "sce.rds"))
print(sce)
```


# **Images and cell masks**

## Import images and masks

```{r import-images}
download.file("https://zenodo.org/record/7079294/files/images.rds", 
              destfile = file.path(workdir, "images.rds"))

images <- readRDS(file.path(workdir, "images.rds"))
```

```{r import-masks}
download.file("https://zenodo.org/record/7079294/files/masks.rds", 
              destfile = file.path(workdir, "masks.rds"))

masks <- readRDS(file.path(workdir, "masks.rds"))
```

## Prepare images and masks

We will now process the images and masks to make them consistent with other datasets.

### Add image names and numbers

Next, we add image names and numbers to the images and masks objects, these names correspond to the `image_name` (respectively, `image_number`) column in `colData(sce)`. This information is stored in the metadata columns of the `CytoImageList` objects and is used by `cytomapper` to match single cell data, images and masks.

```{r add-image-names}
# Add image names
mcols(images)$image_name <- mcols(images)$sample_id
mcols(masks)$image_name <- mcols(masks)$sample_id

# Add image numbers
mcols(images)$image_number <- image_numbers[mcols(images)$image_name]
mcols(masks)$image_number <- image_numbers[mcols(masks)$image_name]

# Fix patient_id column name
names(mcols(images))[names(mcols(images)) == "patient_id.V1"] <- "patient_id"
names(mcols(masks))[names(mcols(masks)) == "patient_id.V1"] <- "patient_id"
```

### Add channel names

Finally, we will add protein short names as channel names of the `images` object with , corresponding to the row names of the `SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names}
if (! identical(rowData(sce)$name, channelNames(images)))
  stop("image names in 'sce' and 'images' objects are different")

# Add channel names to the "images" object
channelNames(images) <- rowData(sce)$short_name
```

### Save on disk

Finally, we will save the generated `CytoImageList` images and masks objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-masks}
saveRDS(masks, file.path(outdir, "masks.rds"))
print(head(masks))
```

```{r save-images}
saveRDS(images, file.path(outdir, "images.rds"))
print(head(images))
```

### Clean up

Remove all files from the temporary working directory.

```{r clean-up, results='hide'}
downloaded_files <- list.files(workdir)
downloaded_files <- downloaded_files[!downloaded_files %in% "BiocStyle"]
unlink(file.path(workdir, downloaded_files), recursive = TRUE)

# Reset original timeout value
options(timeout = timeout)
```

# Session information

```{r session-info}
sessionInfo()
```
