---
title: "Preparation of the Jackson, Fischer et al dataset"
author: "Jana Fischer and Nicolas Damond"
date: "Created: 15 March 2021; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: inline
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script downloads a hundred images (one image per patient), as well as the associated single-cell data and cell segmentation masks from the breast tumour Imaging Mass Cytometry (IMC) dataset described in the following publication:  

[Jackson, H.W., Fischer, J.R. et al. The single-cell pathology landscape of breast cancer. Nature 578, 615–620 (2020)](https://doi.org/10.1038/s41586-019-1876-x).  

All data are openly available from [zenodo](https://doi.org/10.5281/zenodo.3518284).  

Here, we will download single cell data and metadata, and process them to create a [SingleCellExperiment](https://bioconductor.org/packages//SingleCellExperiment.html) object. We will then download the corresponding multichannel IMC images and cell segmentation masks and format them into `CytoImageList` objects using the [cytomapper](https://bioconductor.org/packages/cytomapper) package.


# **Settings**

```{r libraries, include = FALSE}
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
cat("Dataset version: v1")
```

Setting the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
outdir <- file.path(".", "JacksonFischer_2020_BreastCancer")
if(!(dir.exists(outdir)))
    dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


# **Single cell data**

We will download a subset of single-cell data corresponding to 100 images from [@JacksonFischer-2020-BreastCancer] from [zenodo](https://doi.org/10.5281/zenodo.3518284).

## Download single cell data

### Import function

Function to download and unzip files.

```{r import-function}
importData <- function(url, output_dir, filename) {
  # Download
  download.file(url, destfile = file.path(output_dir, filename))
  
  # Unzip
  system2("unzip", args = c("-o",
                            file.path(output_dir, filename),
                            "-d", output_dir),
          stdout = TRUE)
  
  # Remove zipped folder
  file.remove(file.path(output_dir, filename))
}
```

### Single cell data

The first zip folder contains the main single-cell, sample and patient metadata. The single-cell locations and cluster labels are downloaded as separate zip folders because they were uploaded to zenodo separately at a later time point.

```{r import-cell-data, results='hide'}
# Download dataset
url_cells <- ("https://zenodo.org/record/4607374/files/SingleCell_and_Metadata.zip?download=1")
dataset_name <- "SingleCell_and_Metadata.zip"
download.file(url_cells, destfile = file.path(workdir, dataset_name))

# Unzip only required files
system2("unzip", args = c("-o", file.path(workdir, dataset_name),
                          "Data_publication/BaselTMA/SC_dat.csv",
                          "-d", workdir))
system2("unzip", args = c("-o", file.path(workdir, dataset_name),
                          "Data_publication/BaselTMA/Basel_PatientMetadata.csv",
                          "-d", workdir))
system2("unzip", args = c("-o", file.path(workdir, dataset_name),
                          "Data_publication/Basel_Zuri_StainingPanel.csv",
                          "-d", workdir))

file.remove(file.path(workdir, dataset_name))
```

Download single cell labels and locations

```{r import-cell-metadata, results='hide'}
# Clusters
url_cluster <- ("https://zenodo.org/record/4607374/files/singlecell_cluster_labels.zip?download=1")
importData(url_cluster, workdir, "singlecell_cluster_labels.zip")

# Cell locations
url_locations <- ("https://zenodo.org/record/4607374/files/singlecell_locations.zip?download=1")
importData(url_locations, workdir, "singlecell_locations.zip")
```


## Read in single-cell data

We read in single cell data linked to the Basel cohort, including single cell data, cell metadata, clinical data, antibody panel information, and cell clusters.

```{r read-in-cell-data}
# Single cell expressions and spatial features
cells <- fread(file.path(
  workdir, "Data_publication/BaselTMA/SC_dat.csv"))

# Sample and clinical metadata
cell_meta <- fread(file.path(
  workdir, "Data_publication/BaselTMA/Basel_PatientMetadata.csv"))

# Panel information
panel <- fread(file.path(
  workdir, "Data_publication/Basel_Zuri_StainingPanel.csv"))

# Cluster labels (merge PhenoGraph cluster labels and metacluster labels)
phenograph_clusters <- fread(file.path(
  workdir, "Cluster_labels/PG_basel.csv"), header = TRUE)

meta_clusters <- fread(file.path(
  workdir, "Cluster_labels/Basel_metaclusters.csv"), header = TRUE)

clusters <- merge(phenograph_clusters, meta_clusters, by = "id", all.x = TRUE)
remove(phenograph_clusters, meta_clusters)

# Single-cell locations
locations <- fread(file.path(
  workdir, "Basel_SC_locations.csv"), header = TRUE)
```

### Subset patients and images

We select a set of a hundred patients as an example dataset and sample one primary tumor core (one image) per patient.

```{r subset-images}
# Exclude healthy tissue samples and controls
cell_meta <- cell_meta[diseasestatus == "tumor", ]

# Exclude patients for which the tumor core was split across 2 images (machine failure)
cell_meta[, ncores := .N, by = "PID"]
cell_meta <- cell_meta[ncores == 1,]

# Randomly sample 100 patients
set.seed(2)
cell_meta <- cell_meta[PID %in% sample(cell_meta$PID, 100, replace = FALSE), ]
```

We then subset single cell data to selected images.

```{r subset-cell-data}
cells <- cells[core %in% cell_meta$core, ]
locations <- locations[core %in% cell_meta$core, ]
clusters <- clusters[core %in% cell_meta$core, ]
```


## Prepare data

### Extract spatial information

We first extract channels related to spatial information, such as cell area or number of neighbors.

```{r split-cell-metadata}
spatial_channels <- c(
  "Area", "Eccentricity", "Solidity", "Extent", "EulerNumber", "Perimeter",
  "MajorAxisLength","MinorAxisLength", "Orientation", "Percent_Touching",
  "Number_Neighbors"
)

# Spatial channels will go to the colData slot of the SCE
spatial <- cells[channel %in% spatial_channels, ]

# Marker expression levels will go to the assay slot of the SCE
cells <- cells[!channel %in% spatial_channels,]
```

### Cell-level metadata

Here, we will collect all cell-specific metadata in a single `DataFrame`, which will constitute the `colData` entry of the final `SingleCellExperiment` object.

Columns are renamed for consistency with the other datasets.

```{r cell-metadata}
# Wide format spatial single-cell info
cell_metadata <- dcast.data.table(
  spatial, "core + CellId + id ~ channel", value.var = "mc_counts")

# Subset and rename columns
cell_metadata <- cell_metadata[, .(
  image_name = core,
  cell_id = id,
  neighbors_number = Number_Neighbors,
  neighbors_percent_touching = Percent_Touching,
  cell_area = Area,
  cell_perimeter = Perimeter,
  cell_eccentricity = Eccentricity,
  cell_euler_number = EulerNumber,
  cell_extent = Extent,
  cell_major_axis_length = MajorAxisLength,
  cell_minor_axis_length = MinorAxisLength,
  cell_orientation = Orientation,
  cell_solidity = Solidity
)]

# Add single-cell locations
locations <- locations[, .(
  cell_id = id,
  cell_x = Location_Center_X,
  cell_y = Location_Center_Y
)]
cell_metadata <- merge(cell_metadata, locations, by = "cell_id")

# Add clusters
clusters <- clusters[, .(
  cell_id = id,
  cell_cluster_phenograph = PhenoGraphBasel,
  cell_metacluster = cluster
)]
cell_metadata <- merge(cell_metadata, clusters, by = "cell_id")

# Add sample and patient metadata
cell_meta <- cell_meta[, image_number := 1:.N]

# Rename columns
cell_meta <- cell_meta[, .(
  image_name = core,
  image_number = image_number,
  image_width = Width_FullStack,
  image_height = Height_FullStack,
  image_area = area,
  image_filename = FileName_FullStack,
  image_cells_per_image = Count_Cells,
  image_sum_area_cells = sum_area_cells,
  image_percent_tumor_cells = `%tumorcells`,
  image_percent_normal_epithelial_cells = `%normalepithelialcells`,
  image_percent_stroma = `%stroma`,
  image_percent_inflammatory_cells = `%inflammatorycells`,
  patient_id = PID,
  patient_age = age,
  patient_gender = gender,
  patient_status = Patientstatus,
  patient_disease_status = diseasestatus,
  patient_DFS_months = DFSmonth,
  patient_OS_months = OSmonth,
  patient_year_sample_collection = Yearofsamplecollection,
  TMA_location = TMALocation,
  TMA_x = TMAxlocation,
  TMA_y = yLocation,
  TMA_block_label = TMABlocklabel,
  TMA_UBTMA_location = UBTMAlocation,
  tumor_grade = grade,
  tumor_size = tumor_size,
  tumor_primary_site = PrimarySite,
  tumor_primary_diagnosis = Ptdiagnosis,
  tumor_clinical_type = clinical_type,
  tumor_subtype = Subtype,
  tumor_ER_status = ERStatus,
  tumor_HER2_status = HER2Status,
  tumor_HR_status = HR,
  tumor_PR_status = PRStatus,
  tumor_ERpos_ductal_ca = `ER+DuctalCa`,
  tumor_triple_neg_ductal = TripleNegDuctal,
  tumor_hormone_sensitive = hormonesensitive,
  tumor_hormone_resistant_after_sensitive = hormoneresistantaftersenstive,
  tumor_I_plus_neg = I_plus_neg,
  tumor_PTNM_M = PTNM_M,
  tumor_PTNM_N = PTNM_N,
  tumor_PTNM_T = PTNM_T,
  tumor_PTNM_radicality = PTNM_Radicality,
  tumor_microinvasion = microinvasion,
  tumor_lymphatic_invation = Lymphaticinvasion,
  tumor_venous_invasion = Venousinvasion,
  tumor_SN = SN,
  tumor_histology = histology,
  tumor_pre_surgery_Tx_type = `Pre-surgeryTx`,
  tumor_post_surgery_Tx_type = `Post-surgeryTx`,
  tumor_post_surgery_Tx = Post_surgeryTx,
  tumor_response = response
)]
```

We merge the two cell metadata data frames.

```{r merge-metadata}
cell_metadata <- merge(cell_metadata, cell_meta,
                       by = "image_name", all.x = TRUE)
cell_metadata <- DataFrame(cell_metadata)
```

Finally, we add unique cell ids as row names, add cell numbers, and order the cell metadata object based on `image_number` and `cell_number`.

```{r cell-rownames}
# Cell ids are used as row names
rownames(cell_metadata) <- cell_metadata$cell_id
cell_metadata$cell_number <- as.integer(sub(".*_", "", cell_metadata$cell_id))

# Rows are ordered by image and cell numbers
cell_metadata <- cell_metadata[order(cell_metadata$image_number,
                                     cell_metadata$cell_number), ]
```

### Marker metadata

Here, we will collect all marker-related information and collect it in a `DataFrame` that will constitute the `rowData` slot of the `SingleCellExperiment` object.  

We first rename markers for consistency with other datasets.

```{r rename-markers}
# Exclude gas channels
gas_channels <- c("Hg", "In115", "I127", "Pb", "Xe", "Ar")
cells <- cells[!grepl(paste(gas_channels, collapse = "|"), cells$channel),
               . (image_name = core, cell_id = id, channel, counts = mc_counts)]

# Fix marker and metal names
cells[, name := sub(".*Di ", "", channel)]
cells[, metal := sub("Di .*", "", channel)]
cells[, weight := sub(".*[A-Za-z ]", "", metal)]
cells[, metal := gsub("[0-9]+", "", metal)]
cells[, metal := paste0(metal, weight)]

cells[name == "Rutheni", `:=` (short_name = metal, name = metal)]
cells[name == "Iridium", `:=` (short_name = metal, name = metal)]

# Add missing metal info
cells[name == "phospho Histone", `:=` (
  metal = "Eu153", name = "phospho-Histone H3 [S28]", short_name = "p_H3")]
cells[name == "phospho S6", `:=` (
  metal = "Er170", name = "phospho-S6 [S235/S236]", short_name = "p_S6")]
cells[name == "phospho mTOR", `:=` (
  metal = "Yb173", name = "phospho-mTOR [S2448]", short_name = "p_mTOR")]

# Clarify unclear names
cells[name == "cleaved", `:=` (
  name = "cleaved-PARP + cleaved-Caspase3", short_name = "cPARP_cCASP3")]
cells[name == "cerbB", `:=` (
  name = "Epidermal growth factor receptor-2", short_name = "HER2")]
cells[name == "Carboni", `:=` (
  name = "Carbonic anhydrase IX", short_name = "CA9")]

cells[metal == "In113", `:=` (name = "Histone H3", short_name = "H3")]
cells[metal == "La139", `:=` (name = "H3K27me3", short_name = "H3K27me3")]
cells[metal == "Pr141", `:=` (name = "Cytokeratin 5", short_name = "KRT5")]
cells[metal == "Nd142", `:=` (name = "Fibronectin", short_name = "FN1")]
cells[metal == "Nd143", `:=` (name = "Cytokeratin 19", short_name = "KRT19")]
cells[metal == "Nd144", `:=` (name = "Cytokeratin 8/18", short_name = "KRT8_18")]
cells[metal == "Nd145", `:=` (name = "Twist", short_name = "TWIST1")]
cells[metal == "Sm147", `:=` (name = "Cytokeratin 14", short_name = "KRT14")]
cells[metal == "Nd148", `:=` (name = "Smooth muscle actin", short_name = "SMA")]
cells[metal == "Sm149", `:=` (name = "Vimentin", short_name = "VIM")]
cells[metal == "Nd150", `:=` (name = "c-Myc", short_name = "c_Myc")]
cells[metal == "Sm152", `:=` (name = "CD3 epsilon", short_name = "CD3e")]
cells[metal == "Gd155", `:=` (name = "Slug", short_name = "SNAI2")]
cells[metal == "Tb159", `:=` (name = "p53", short_name = "p53")]
cells[metal == "Gd156", `:=` (name = "Estrogen receptor alpha",
                              short_name = "ERa")]
cells[metal == "Gd158", `:=` (name = "Progesterone receptor A/B",
                              short_name = "PGR")]
cells[metal == "Er167", `:=` (name = "E-Cadherin", short_name = "CDH1")]
cells[metal == "Er168", `:=` (name = "Ki-67", short_name = "Ki67")]
cells[metal == "Tm169", `:=` (name = "Epidermal growth factor receptor",
                         short_name = "EGFR")]
cells[metal == "Yb172", `:=` (name = "von Willebrand factor",
                              short_name = "vWF")]
cells[metal == "Yb174", `:=` (name = "Cytokeratin 7", short_name = "KRT7")]
cells[metal == "Lu175", `:=` (name = "Pan-cytokeratin", short_name = "PanCK")]
cells[metal == "Ir191", `:=` (name = "Ir191", short_name = "DNA1")]
cells[metal == "Ir193", `:=` (name = "Ir193", short_name = "DNA2")]

# Add missing short names
cells[name %in% c("CD68", "p53", "CD44", "CD45", "GATA3", "CD20"),
      short_name := name]
```

We then import the panel and fix antibody clone names.

```{r assemble-panel}
# Select columns
panel <- panel[, .(
  channel = FullStack,
  metal = `Metal Tag`,
  antibody_clone = `Antibody Clone`
)]

# Fix antibody clones names
panel[metal == "Gd158", antibody_clone := "EP2 + SP2"]
panel[metal == "Yb176", antibody_clone := "F21-852 + C92-605"]
panel <- panel[!duplicated(metal), ]
panel <- panel[metal %in% unique(cells$metal)]
panel[antibody_clone == "", antibody_clone := NA]
panel[metal == "Sm147", `:=` (antibody_clone = "polyclonal_CK14")]
panel[metal == "Eu153", `:=` (antibody_clone = "HTA28")]
panel[metal == "Gd156", `:=` (antibody_clone = "polyclonal_anti_rabbit_IgG")]
panel[metal == "Er167", `:=` (antibody_clone = "36/E-Cadherin")]
panel[metal == "Yb172", `:=` (antibody_clone = "polyclonal_vWF")]

# Merge all panel information
panel <- merge(unique(cells[, .(metal, name, short_name)]), panel, by = "metal")
```

Finally, we convert the panel table to a `DataFrame` and add target short_names as row names.

```{r prepare-panel}
panel <- panel[order(panel$channel), ]
panel <- as(panel, "DataFrame")
rownames(panel) <- panel$short_name
```

### Counts matrix

Here, we will prepare the counts matrix that will be stored in the `assay` slot of the `SingleCellExperiment` object.

We extract marker expression values from the `cells` table and convert them to a matrix. We then order the counts matrix by `cell_id` and `short_name`.

```{r prepare-counts}
counts <-  dcast.data.table(cells, "cell_id ~ short_name",
                            value.var = "counts")
row_names <- counts$cell_id
counts[, cell_id := NULL]
counts <- as.matrix(counts, rownames = NULL)
rownames(counts) <- row_names

counts <- counts[order(match(rownames(counts), rownames(cell_metadata))),
                 order(match(colnames(counts), rownames(panel)))]
```


## Create SingleCellExperiment object

### Create the object

We have now obtained all metadata and feature data to create the `SingleCellExperiment` object.

```{r create-SCE}
sce <- SingleCellExperiment(
  assays = list(counts = t(counts)),
  rowData = panel,
  colData = cell_metadata
)
mainExpName(sce) <- "JacksonFischer_2020_BreastCancer"
```

### Counts transformations

We apply two different counts transformations:
- `exprs`: arcsinh-transformed counts (cofactor = 1).
- `quant_norm`: censored + quantile-normalized counts.

```{r transform-counts}
assay(sce, "exprs") <- asinh(counts(sce) / 1)

quant <- apply(assay(sce, "counts"), 1, quantile, probs = 0.99)
assay(sce, "quant_norm") <- apply(assay(sce, "counts"), 2,
                                  function(x) x / quant)
assay(sce, "quant_norm")[assay(sce, "quant_norm") > 1] <- 1
assay(sce, "quant_norm")[assay(sce, "quant_norm") < 0] <- 0
```

### Save on disk

We save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
saveRDS(sce, file.path(outdir, "sce.rds"))
```

### Clean up

Finally, we remove the downloaded files and generated objects to save storage space.

```{r clean-up-cell-data, results='hide'}
remove(cells, cell_metadata, spatial, counts, locations, clusters, row_names)

file.remove(file.path(workdir, "Basel_SC_locations.csv"),
            file.path(workdir, "Zurich_SC_locations.csv"))
unlink(file.path(workdir, "Data_publication"), recursive = TRUE)
unlink(file.path(workdir, "Cluster_labels"), recursive = TRUE)
```


# **Images and cell masks**

Here, we will download a subset of a hundred images from [@JacksonFischer-2020-BreastCancer], as well as the corresponding cell segmentation masks. Images and masks correspond to the data in the `SingleCellExperiment` object and will be formatted into `CytoImageList` objects.


## Import images and masks

### Multichannel images

We first download and unzip the image subset.

```{r import-images, results='hide'}
# Download images and masks
url_images <- ("https://zenodo.org/record/4607374/files/OMEandSingleCellMasks.zip?download=1")
importData(url_images, workdir, "OMEandSingleCellMasks.zip")

# Unzip image folder
fn <- file.path("OMEnMasks", "ome.zip")
system2("unzip", args = c("-o", file.path(workdir, fn),
                          "*BaselTMA_*.tiff",
                          "-d", workdir),
        stdout = TRUE)
file.remove(file.path(workdir, fn))
```

Loading all tiffs would require too much memory, so we delete all the `tiff` files that we do not want to keep. We then use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r load-images, results = 'hide', warning=FALSE}
tiffs_delete <- list.files(file.path(workdir, "ome/"))
tiffs_delete <-  tiffs_delete[!tiffs_delete %in% unique(sce$image_filename)]
file.remove(file.path(workdir, "ome/", tiffs_delete))

# Load the images as a CytoImageList object
images <- loadImages(file.path(workdir, "ome/"), pattern = "_full.tiff")
```

### Cell segmentation masks

We also unzip the associated cell segmentation masks and read them into a `CytoImageList` object.

```{r import-masks, results='hide'}
fn <- file.path("OMEnMasks", "Basel_Zuri_masks.zip")
system2("unzip", args = c("-o", file.path(workdir, fn),
                          "*BaselTMA_*.tiff",
                          "-d", workdir),
        stdout = TRUE)
file.remove(file.path(workdir, fn))
```

```{r read-in-masks}
masks <- loadImages(file.path(workdir, "Basel_Zuri_masks"),
                    pattern = "_full_maks.tiff")
masks
```

### Clean-up

We remove the downloaded image and mask `tiff` files to save storage space.

```{r clean-up-images-masks, results='hide'}
images_to_delete <- list.files(file.path(workdir, "ome"),
                               pattern = "*.tiff",
                               full.names = TRUE)
file.remove(images_to_delete)

# Remove masks
masks_to_delete <- list.files(file.path(workdir, "Basel_Zuri_masks"),
                              pattern = "*.tiff",
                              full.names = TRUE)
file.remove(masks_to_delete)
```


## Prepare images and masks

We will now process the images and masks to make them compatible with the `cytomapper` package.

### Rescale masks

The masks are 16-bit images and need to be re-scaled in order to obtain integer cell ids.

```{r scale-masks}
# Before scaling
range(masks[[1]])

masks <- scaleImages(masks, value = (2 ^ 16) - 1)

# After scaling
range(masks[[1]])
```

### Subset and fix masks

We subset the masks to the hundred images contained in the `SingleCellExperimentObject`

```{r subset-masks}
mcols(masks)$image_filename <- gsub("_full_maks", "_full.tiff", names(masks))
masks <- masks[mcols(masks)$image_filename %in% unique(sce$image_filename)]
```

Occasionally a single-cell ID is skipped in the masks but in the single-cell data the cell numbers were renamed sequentially. Therefore, the single-cell IDs in the masks also have to renamed sequentially in order to correspond to the cell numbers from the single-cell data.

```{r fix-masks}
# Rename single-cell IDs sequentially in each mask
for (n in names(masks)){
  imageData(masks[[n]]) = plyr::mapvalues(
    imageData(masks[[n]]),
    sort(unique(as.integer(imageData(masks[[n]])))),
    0:(length(unique(as.integer(imageData(masks[[n]]))))-1)
  )
}
```

### Add image names and numbers

Next, we add image names to the images and masks objects, these names correspond to the `image_name` column in `colData(sce)`. This information is stored in the metadata columns of the `CytoImageList` objects and is used by `cytomapper` to match single cell data, images and masks.

```{r add-image-names}
# Images
mcols(images)$image_filename <- paste0(names(images), ".tiff")
images <- images[mcols(images)$image_filename %in% sce$image_filename]
mcols(images) <- merge(
  mcols(images),
  unique(colData(sce)[, c("image_filename", "image_name", "image_number")]),
  by = "image_filename")

# Masks
mcols(masks) <- merge(
  mcols(masks),
  unique(colData(sce)[, c("image_filename", "image_name", "image_number")]),
  by = "image_filename")

identical(mcols(masks)$image_name, mcols(images)$image_name)
```

### Subset channels

Here, we exclude channels that are not present in the `SingleCellExperiment` object.

```{r subset-channels}
panel <- rowData(sce)
panel <- panel[order(panel$channel), ]
images <- getChannels(images, panel$channel)
```

### Add channel names

Finally, we will add protein short names as channel names of the `images` object with , corresponding to the row names of the `SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names}
channelNames(images) <- rownames(panel)
```

### Save on disk

Finally, we will save the generated `CytoImageList` images and masks objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r clean-up-memory, echo=FALSE, results='hide'}
gc()
```

```{r save-masks}
saveRDS(masks, file.path(outdir, "masks.rds"))
```

```{r save-images}
saveRDS(images, file.path(outdir, "images.rds"))
```

## Clean up

Remove all files from the temporary working directory.

```{r clean-up, results='hide'}
downloaded_files <- list.files(workdir)
downloaded_files <- downloaded_files[!downloaded_files %in% "BiocStyle"]
unlink(file.path(workdir, downloaded_files), recursive = TRUE)

# Reset original timeout value
options(timeout = timeout)
```


# **Session information**

```{r session-info}
sessionInfo()
```


# **References**