---
title: "Preparation of the Hoch, Schulz et al dataset"
author: "Tobias Hoch and Nicolas Damond"
date: "Created: 17 June 2022; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: console
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script downloads two example Imaging Mass Cytometry (IMC) datasets (RNA and protein), each comprising fifty corresponding images and the associated cell segmentation masks and single cell data. The datasets are described in the following publication:

[Hoch, Schulz et al. Multiplexed imaging mass cytometry of the chemokine milieus in melanoma characterizes features of the response to immunotherapy. Sci immunol. 2022 Apr;7(70):eabk1692](http://doi.org/10.1126/sciimmunol.abk1692)

All data are available [from zenodo](https://zenodo.org/record/5994136). After obtaining the raw data, we will further process them to create a [SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) object. We will then use the [cytomapper](https://www.bioconductor.org/packages/release/bioc/html/cytomapper.html) package to read in the images and masks and create `CytoImageList` objects.


# **Settings**

```{r libraries, include = FALSE}
library(data.table)
library(dplyr)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
dataset_version <- "v1"
cat("Dataset version:", dataset_version)
```

Set the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
dataset_dir <- file.path(".", "HochSchulz_2022_Melanoma")
if(!(dir.exists(dataset_dir))) dir.create(dataset_dir)

outdir <- file.path(dataset_dir, dataset_version)
if(!(dir.exists(outdir))) dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


# **Single cell data**

## Download single cell data

Here, a subset of single-cell data corresponding to two datasets (protein and RNA) from [@HochSchulz-2022-Melanoma] is downloaded.

```{r import-cell-data, results='hide'}
# Download the zipped folder
url_data <- "https://zenodo.org/record/5994136/files/data_for_analysis.zip"
dataset_name <- "sceFiles"

system2("curl", args = c(
  "-o", 
  file.path(workdir, paste0(dataset_name, ".zip")), 
  url_data,
  "--max-time 1000"))

# Unzip SCE files
system2("unzip", args = c(
  "-o", file.path(workdir, paste0(dataset_name, ".zip")),
  "*sce_*.rds",
  "-d", workdir))

file.remove(file.path(workdir, paste0(dataset_name, ".zip")))
```


## Read in single-cell data

We read in the `SingleCellExperiment` objects from both datasets and subset 50 (corresponding) images based on the highest fraction of B cells.

```{r read-in-cell-data}
# RNA dataset
sce_rna <- readRDS(
  file.path(workdir, "data_for_analysis_zenodo/sce_RNA.rds"))

# Protein dataset
sce_protein <- readRDS(
  file.path(workdir, "data_for_analysis_zenodo/sce_protein.rds"))
```

### Subset images

Select the top 50 images with the most B cells.

```{r select-images}
images_to_keep <- as.data.frame(colData(sce_protein)) %>%
  group_by(Description, celltype) %>%
  summarise(numberOfcells = n(), .groups = "keep") %>%
  filter(celltype == "B cell") %>%
  ungroup() %>%
  slice_max(numberOfcells, n = 50) %>%
  pull(Description)
```

Subset single cell data to selected images.

```{r subset-cell-data}
sce_protein <- sce_protein[, sce_protein$Description %in% images_to_keep]
sce_rna <- sce_rna[, sce_rna$Description %in% images_to_keep]
```

## Prepare data

### Cell-level metadata

Here, we rename the columns of the `colData` entry for consistency with the other datasets.

```{r rename-cell-metadata}
renaming_vector <- c(
  cell_id = "cellID",
  cell_number = "CellNumber",
  image_number = "ImageNumber",
  cell_x = "Center_X",
  cell_y = "Center_Y",
  cell_type = "celltype",
  cell_cluster = "celltype_clustered",
  cell_area = "Area",
  cell_major_axis_length = "MajorAxisLength",
  cell_minor_axis_length = "MinorAxisLength",
  cell_in_tumour = "in_tumor",
  milieu_Bcell_patch_score = "bcell_patch_score",
  milieu_Tcell_density_score = "Tcell_density_score_image",
  milieu_dysfunction_score = "dysfunction_score",
  milieu_dysfunction_density = "dysfunction_density",
  neighbors_number = "NumberOfNeighbors",
  sample_block_id = "BlockID",
  sample_location = "Location",
  sample_description = "Description",
  sample_tissue_type = "TissueType",
  sample_MM_location = "MM_location",
  sample_MM_location_short = "MM_location_simplified",
  patient_id = "PatientID",
  patient_age = "Age",
  patient_gender = "Gender",
  patient_cancer_stage = "Cancer_Stage",
  patient_adjuvant_therapy = "Adjuvant",
  patient_E_I_D = "E_I_D",
  patient_mutation = "Mutation",
  patient_status_at_3months = "Status_at_3m",
  patient_relapse = "relapse",
  patient_treatment_status_before_surgery = "treatment_status_before_surgery",
  patient_treatment_group_before_surgery = "treatment_group_before_surgery",
  patient_treatment_group_after_surgery = "treatment_group_after_surgery",
  patient_death = "Death",
  patient_death_date = "Date_death"
)

renaming_vector_protein = c(
  cell_parent_nucleus = "Parent_nuclei",
  milieu_Bcell_patch = "bcell_patch",
  milieu_Bcell_milieu = "bcell_milieu",
  milieu_TCF7 = "TCF7",
  milieu_PD1 = "PD1"
)

renaming_vector_rna = c(
  cell_type_rf = "celltype_rf",
  cell_annotation = "cellAnnotation",
  chemokine_expressor = "expressor"
)

# Rename protein dataset
protein_coldata <- as_tibble(colData(sce_protein))
protein_coldata <- protein_coldata %>%
  dplyr::rename(all_of(renaming_vector)) %>%
  dplyr::rename(all_of(renaming_vector_protein)) %>%
  as("DataFrame")

protein_coldata$cell_id <- gsub("protein_", "", protein_coldata$cell_id)
protein_coldata$image_name <- as.character(protein_coldata$image_number)
rownames(protein_coldata) <- protein_coldata$cell_id

# Rename RNA dataset
rna_coldata <- as_tibble(colData(sce_rna))
rna_coldata <- rna_coldata %>%
  dplyr::rename(all_of(renaming_vector)) %>%
  dplyr::rename(all_of(renaming_vector_rna)) %>%
  as("DataFrame")

rna_coldata$cell_id <- gsub("RNA_", "", rna_coldata$cell_id)
rna_coldata$image_name <- as.character(rna_coldata$image_number)
rownames(rna_coldata) <- rna_coldata$cell_id
```

### Marker metadata

Here, we rename the columns of the `rowData` entry for consistency with the other datasets.

```{r rename-marker-metadata}
renaming_vector <- c(
  channel = "channel",
  metal = "Metal.Tag",
  name = "Target",
  short_name = "clean_target",
  marker_class = "marker_class",
  antibody_clone = "Antibody.Clone",
  antibody_tube_number = "Tube.Number",
  antibody_stock_conc = "Stock.Concentration",
  antibody_final_conc_or_dilution = "Final.Concentration...Dilution",
  antibody_ul_to_add = "uL.to.add",
  full = "full",
  keep = "good_marker",
  ilastik = "ilastik"
)

# Rename protein dataset
protein_rowdata <- as_tibble(rowData(sce_protein))
protein_rowdata <- protein_rowdata %>%
  dplyr::rename(all_of(renaming_vector)) %>%
  dplyr::rename(c(tumor_mask = "tumorMask")) %>%
  as.data.table()
protein_rowdata$full_name <- protein_rowdata$name

# Rename RNA dataset
rna_rowdata <- as_tibble(rowData(sce_rna))
rna_rowdata <- rna_rowdata %>%
  dplyr::rename(all_of(renaming_vector)) %>%
  dplyr::rename(c(marker_no_RNA = "noRNA_marker")) %>%
  as.data.table()
rna_rowdata$full_name <- rna_rowdata$name
```

We also rename markers for consistency with other datasets.

```{r rename-markers}
# Protein dataset
protein_rowdata[metal == "Cd111", `:=` (full_name = "Vimentin",
                                         short_name = "VIM")]
protein_rowdata[metal == "Cd112", `:=` (short_name = "CAV1")]
protein_rowdata[metal == "In113", `:=` (short_name = "H3")]
protein_rowdata[metal == "In115", `:=` (full_name = "Smooth muscle actin")]
protein_rowdata[metal == "Nd142", `:=` (full_name = "CXCR2 IL-8 RB")]
protein_rowdata[metal == "Nd143", `:=` (short_name = "HLA_DR")]
protein_rowdata[metal == "Sm147", `:=` (full_name = "SOX-9",
                                        short_name = "SOX9")]
protein_rowdata[metal == "Eu151", `:=` (
  full_name = "phospho-ERK1/2 [T202/Y204]",
  short_name = "p_ERK1_2")]
protein_rowdata[metal == "Sm152", `:=` (full_name = "CD3 epsilon",
                                        short_name = "CD3e")]
protein_rowdata[metal == "Gd155", `:=` (
  full_name = "Programmed cell death protein 1",
  short_name = "PD_1")]
protein_rowdata[metal == "Gd156", `:=` (short_name = "MITF")]
protein_rowdata[metal == "Tb159", `:=` (short_name = "GZMB")]
protein_rowdata[metal == "Gd160", `:=` (full_name = "Programmed death-ligand 1",
                                        short_name = "PD_L1")]
protein_rowdata[metal == "Dy163", `:=` (full_name = "Forkhead box P3")]
protein_rowdata[metal == "Dy164", `:=` (
  full_name = "Inducible T-cell costimulator",
  short_name = "ICOS")]
protein_rowdata[metal == "Ho165", `:=` (full_name = "Beta-catenin",
                                        short_name = "CTNNB1")]
protein_rowdata[metal == "Er166", `:=` (full_name = "CD8 alpha",
                                        short_name = "CD8a")]
protein_rowdata[metal == "Er167", `:=` (
  short_name = "COL1A1",
  antibody_clone = "polyclonal_Collagen I")]
protein_rowdata[metal == "Er168", `:=` (full_name = "Ki-67_Er168",
                                        short_name = "Ki67_Er168")]
protein_rowdata[metal == "Er170", `:=` (full_name = "phospho-S6 [S235/S236]",
                                        short_name = "p_S6")]
protein_rowdata[metal == "Yb172", `:=` (
  full_name = "Indoleamine 2,3-dioxygenase 1")]
protein_rowdata[metal == "Yb173", `:=` (full_name = "SOX-10",
                                        short_name = "SOX10")]
protein_rowdata[metal == "Yb174", `:=` (
  antibody_clone = "polyclonal_DLEC_CLEC4C_BDCA-2")]
protein_rowdata[metal == "Lu175", `:=` (full_name = "CD206")]
protein_rowdata[metal == "Yb176", `:=` (full_name = "cleaved-PARP",
                                        short_name = "c_PARP")]
protein_rowdata[metal == "Ir191", `:=` (full_name = "Iridum 191",
                                        antibody_clone = NA)]
protein_rowdata[metal == "Ir193", `:=` (full_name = "Iridium 193",
                                        antibody_clone = NA)]
protein_rowdata[metal == "Pt198", `:=` (full_name = "Ki-67_Pt198",
                                        short_name = "Ki67_Pt198")]
protein_rowdata[metal == "Y89", `:=` (full_name = "Myeloperoxidase",
                                      antibody_clone = "polyclonal_MPO")]

# RNA dataset
rna_rowdata[metal == "Cd111", `:=` (full_name = "Vimentin",
                                    short_name = "VIM")]
rna_rowdata[metal == "In113", `:=` (short_name = "H3")]
rna_rowdata[metal == "In115", `:=` (full_name = "Smooth muscle actin")]
rna_rowdata[metal == "Pr141", `:=` (full_name = "Cytokeratin 5",
                                    short_name = "KRT5")]
rna_rowdata[metal == "Nd143", `:=` (short_name = "HLA_DR")]
rna_rowdata[metal == "Nd145", `:=` (full_name = "Cadherin-11",
                                    short_name = "CDH11")]
rna_rowdata[metal == "Nd146", `:=` (
  full_name = "Fibroblast activation protein alpha",
  short_name = "FAP")]
rna_rowdata[metal == "Nd148", `:=` (full_name = "Beta-2-microglobulin")]
rna_rowdata[metal == "Eu151", `:=` (full_name = "Glucose transporter 1",
                                    short_name = "SLC2A12")]
rna_rowdata[metal == "Sm152", `:=` (full_name = "CD3 epsilon",
                                    short_name = "CD3e")]
rna_rowdata[metal == "Eu153", `:=` (
  full_name = "Lymphocyte activation gene 3 protein",
  short_name = "LAG3")]
rna_rowdata[metal == "Gd155", `:=` (
  full_name = "Programmed cell death protein 1",
  short_name = "PD_1")]
rna_rowdata[metal == "Gd156", `:=` (short_name = "CCL4_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Gd158", `:=` (short_name = "CCL18_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Tb159", `:=` (full_name = "C-C chemokine receptor type 2",
                                    short_name = "CCR2")]
rna_rowdata[metal == "Gd160", `:=` (full_name = "Programmed death-ligand 1",
                                    short_name = "PD_L1")]
rna_rowdata[metal == "Dy161", `:=` (short_name = "CXCL8_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Dy162", `:=` (short_name = "CXCL10_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Dy163", `:=` (short_name = "CXCL12_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Dy164", `:=` (short_name = "CXCL13_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Ho165", `:=` (full_name = "CD8 alpha",
                                    short_name = "CD8a")]
rna_rowdata[metal == "Er166", `:=` (short_name = "CCL2_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Er167", `:=` (short_name = "CCL22_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Er168", `:=` (short_name = "CXCL9_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Tm169", `:=` (short_name = "DapB_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Er170", `:=` (full_name = "SOX-10",
                                    short_name = "SOX10")]
rna_rowdata[metal == "Yb171", `:=` (short_name = "CCL8_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Yb173", `:=` (short_name = "CCL19_mRNA",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Yb174", `:=` (full_name = "MelanA / MART1",
                                    short_name = "MLANA")]
rna_rowdata[metal == "Lu175", `:=` (full_name = "phopsho-Rb [S807/S811]",
                                    short_name = "p_Rb")]
rna_rowdata[metal == "Yb176", `:=` (full_name = "cleaved-PARP",
                                    short_name = "c_PARP")]
rna_rowdata[metal == "Ir191", `:=` (full_name = "Iridium 191",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Ir193", `:=` (full_name = "Iridium 193",
                                    antibody_clone = NA)]
rna_rowdata[metal == "Y89", `:=` (full_name = "Myeloperoxidase",
                                  antibody_clone = "polyclonal_MPO")]
```

We convert the panel table to a `DataFrame` and add target short_names as row names.

```{r prepare-panel}
protein_rowdata <- as(protein_rowdata, "DataFrame")
rownames(protein_rowdata) <- protein_rowdata$short_name

rna_rowdata <- as(rna_rowdata, "DataFrame")
rownames(rna_rowdata) <- rna_rowdata$short_name
```


## Update SingleCellExperiment object

We now update the `SingleCellExperiment` objects with new `rowData` and `colData` entries.

```{r update-colData}
# Protein dataset
colData(sce_protein) <- protein_coldata
rowData(sce_protein) <- protein_rowdata

# RNA dataset
colData(sce_rna) <- rna_coldata
rowData(sce_rna) <- rna_rowdata
```

Finally, we rename the row and column names of the `SingleCellExperiment` objects.

```{r rename-count-matrices}
# Protein dataset
rownames(sce_protein) <- rownames(protein_rowdata)
colnames(sce_protein) <- rownames(protein_coldata)

# RNA dataset
rownames(sce_rna) <- rownames(rna_rowdata)
colnames(sce_rna) <- rownames(rna_coldata)

# MainExpName
mainExpName(sce_protein) <- "HochSchulz_2022_Melanoma_protein"
mainExpName(sce_rna) <- "HochSchulz_2022_Melanoma_RNA"
```

We also rename the assays for consistency with other datasets.

```{r rename-assays}
assayNames(sce_protein) <- c("counts", "exprs", "scaled_counts", "scaled_exprs")
assayNames(sce_rna) <- c("counts", "exprs", "scaled_counts", "scaled_exprs")
```

### Save on disk

We save the `SingleCellExperiment` objects for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
saveRDS(sce_protein, file.path(outdir, "sce_protein.rds"))
print(sce_protein)

saveRDS(sce_rna, file.path(outdir, "sce_rna.rds"))
print(sce_rna)
```

### Clean up

Finally, we remove the downloaded files and generated objects to save storage space.

```{r clean-up-cell-data, results='hide'}
remove(protein_coldata, protein_rowdata, rna_coldata, rna_rowdata)

unlink(file.path(workdir, "data_for_analysis_zenodo"), recursive = TRUE)
```


# **Images and cell masks**

## Import images and masks - Protein dataset

### Download the dataset

Here, we download a subset of 50 images from the [zenodo repository](https://zenodo.org/record/5994136#.YyLNv7RBz-h).

```{r download-images-protein, message=FALSE}
# Download the zipped folder image and unzip it
system2("curl", args = c(
  "-o", 
  file.path(workdir, "ImageSubset.zip"), 
  "https://zenodo.org/record/6004986/files/full_data.zip?download=1",
  "--max-time 900"))

# Unzip selected files
system2("unzip", args = c(
  "-o", file.path(workdir, "ImageSubset.zip"),
  "full_data/protein/cpout/*.tiff",
  "-d", workdir))

system2("unzip", args = c(
  "-o", file.path(workdir, "ImageSubset.zip"),
  "full_data/protein/cpout/Image.csv",
  "-d", workdir))
```

We select the fifty images that correspond to the dat in the `SingleCellExperiment` objects.

```{r select-images-protein}
protein_directory = file.path(
  workdir, file.path("full_data", "protein", "cpout"))
protein_info <- fread(file.path(protein_directory, "Image.csv"))
protein_info <- protein_info[
  protein_info$Metadata_Description %in% images_to_keep, ]
```

### Multichannel images

We use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r read-in-images-protein}
images_protein <- loadImages(
  protein_directory, pattern = protein_info$FileName_SpillCorrected)
```

### Cell segmentation masks

We also read the masks into a `CytoImageList` object.

```{r read-in-masks-protein, results='hide'}
masks_protein <- loadImages(
  protein_directory, pattern = protein_info$FileName_cellmask)
```

### Clean-up

We remove the downloaded image and mask `tiff` files to save storage space.

```{r clean-up-images-mask-proteins, results='hide'}
files_to_delete <- list.files(protein_directory, full.names = TRUE)
file.remove(files_to_delete)
```


## Import images and masks - RNA dataset

### Unzip the dataset

The dataset has been already been downloaded from zenodo, together with the protein dataset.

```{r load-images-rna, results='hide'}
system2("unzip", args = c(
  "-o", file.path(workdir, "ImageSubset.zip"),
  "full_data/rna/cpout/*.tiff",
  "-d", workdir)) 

system2("unzip", args = c(
  "-o", file.path(workdir, "ImageSubset.zip"),
  "full_data/rna/cpout/Image.csv",
  "-d", workdir))

# Clean-up
file.remove(file.path(workdir, "ImageSubset.zip"))
```

We select the fifty images that correspond to the dat in the `SingleCellExperiment` objects.

```{r select-images-rna}
rna_directory <- file.path(workdir, file.path("full_data", "rna", "cpout"))
rna_info <- fread(file.path(rna_directory, "Image.csv"))
rna_info <- rna_info[rna_info$Metadata_Description %in% images_to_keep, ]
```

### Multichannel images

We use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r read-in-images-rna}
images_rna <- loadImages(
  rna_directory, pattern = rna_info$FileName_SpillCorrected)
```

### Cell segmentation masks

We also read the masks into a `CytoImageList` object.

```{r read-in-masks-rna, results='hide'}
masks_rna <- loadImages(
  rna_directory, pattern = rna_info$FileName_cellmask)
```

### Clean-up

We remove the downloaded image and mask `tiff` files to save storage space.

```{r clean-up-images-masks-rna, results='hide'}
files_to_delete <- list.files(rna_directory, full.names = TRUE)
file.remove(files_to_delete)
```


## Prepare images and masks

We will now process the images and masks to make them compatible with the `cytomapper` package.

### Rescale masks

The masks are 16-bit images and need to be re-scaled in order to obtain integer cell ids.

```{r scale-masks}
masks_protein <- scaleImages(masks_protein,
                             value = protein_info$Scaling_cellmask[[1]])
masks_rna <- scaleImages(masks_rna,
                         value = rna_info$Scaling_cellmask[[1]])
```

### Add image names and numbers

Next, we add image names to the images and masks objects, these names correspond to the `image_name` column in `colData(sce_*)`. This information is stored in the metadata columns of the `CytoImageList` objects and is used by `cytomapper` to match single cell data, images and masks.

```{r add-image-names-protein}
# We extract the file names of the masks into data frames
df_masks_protein <- data.frame(
  cell_mask = protein_info$FileName_cellmask,
  image_number = protein_info$ImageNumber,
  description = protein_info$Metadata_Description)

df_images_protein <- data.frame(
  cell_mask = protein_info$FileName_SpillCorrected,
  image_number = protein_info$ImageNumber,
  description = protein_info$Metadata_Description)

# We match the data frames row names with the names of the 'masks' object
rownames(df_masks_protein) <- gsub(".tiff", "",
                                   protein_info$FileName_cellmask)
rownames(df_images_protein) <- gsub(".tiff", "", 
                                    protein_info$FileName_SpillCorrected)

# We add the extracted information to the metadata of the 'masks' object
mcols(masks_protein) <- df_masks_protein[names(masks_protein), ]
mcols(images_protein) <- df_images_protein[names(images_protein), ]

# Add image names
mcols(masks_protein)$image_name <- as.character(
  mcols(masks_protein)$image_number)
mcols(images_protein)$image_name <- as.character(
  mcols(images_protein)$image_number)
names(masks_protein) <- mcols(masks_protein)$image_name
names(images_protein) <- mcols(images_protein)$image_name

# Add dataset specification
mcols(images_protein)$dataset <- rep("protein", nrow(mcols(images_protein)))
mcols(masks_protein)$dataset <- rep("protein", nrow(mcols(masks_protein)))
```

```{r add-image-names-rna}
# We extract the file names of the masks into data frames
df_masks_rna <- data.frame(
  cell_mask = rna_info$FileName_cellmask,
  image_number = rna_info$ImageNumber,
  description = rna_info$Metadata_Description)

df_images_rna <- data.frame(
  cell_mask = rna_info$FileName_SpillCorrected,
  image_number = rna_info$ImageNumber,
  description = rna_info$Metadata_Description)

# We match the data frames row names with the names of the 'masks' object
rownames(df_masks_rna) <- gsub(".tiff", "", rna_info$FileName_cellmask)
rownames(df_images_rna) <- gsub(".tiff", "", rna_info$FileName_SpillCorrected)

# We add the extracted information to the metadata of the 'masks' object
mcols(masks_rna) <- df_masks_rna[names(masks_rna), ]
mcols(images_rna) <- df_images_rna[names(images_rna), ]

# Add image names
mcols(masks_rna)$image_name <- as.character(
  mcols(masks_rna)$image_number)
mcols(images_rna)$image_name <- as.character(
  mcols(images_rna)$image_number)
names(masks_rna) <- mcols(masks_rna)$image_name
names(images_rna) <- mcols(images_rna)$image_name

# Add dataset specification
mcols(images_rna)$dataset <- rep("RNA", nrow(mcols(images_rna)))
mcols(masks_rna)$dataset <- rep("RNA", nrow(mcols(masks_rna)))
```

Sanity check

```{r sanity-check}
identical(mcols(masks_rna)$Description, mcols(images_rna)$Description)
identical(mcols(masks_protein)$Description, mcols(images_protein)$Description)
```

### Add channel names

Finally, we will add protein short names as channel names of the `images` objects with , corresponding to the row names of the `SingleCellExperiment` objects and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names}
# Protein dataset
panel_protein <- rowData(sce_protein)
panel_protein <- panel_protein[order(panel_protein$channel), ]
channelNames(images_protein) <- rownames(panel_protein)

# RNA dataset
panel_rna <- rowData(sce_rna)
panel_rna <- panel_rna[order(panel_rna$channel), ]
channelNames(images_rna) <- rownames(panel_rna)
```

### Save on disk

Finally, we will save the generated `CytoImageList` images and masks objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-masks}
saveRDS(masks_protein, file.path(outdir, "masks_protein.rds"))
print(masks_protein)

saveRDS(masks_rna, file.path(outdir, "masks_rna.rds"))
print(masks_rna)
```

```{r save-images}
saveRDS(images_protein, file.path(outdir, "images_protein.rds"))
print(images_protein)

saveRDS(images_rna, file.path(outdir, "images_rna.rds"))
print(images_rna)
```

### Clean up

Remove all files from the temporary working directory.

```{r clean-up, results='hide'}
downloaded_files <- list.files(workdir)
downloaded_files <- downloaded_files[!downloaded_files %in% "BiocStyle"]
unlink(file.path(workdir, downloaded_files), recursive = TRUE)

# Reset original timeout value
options(timeout = timeout)
```


# **Session information**

```{r session-info}
sessionInfo()
```


# **References**
