---
title: "Preparation of the Meyer et al dataset"
author: "Lasse Meyer"
date: "Created: 22 December 2025; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: inline
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script downloads single-cell data, multichannel images and cell segmentation masks from the triple-negative breast cancer Imaging Mass Cytometry (IMC) dataset described in the following publication:  

[Meyer et al. A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Cancer Cell (2025)](https://doi.org/10.1016/j.ccell.2025.06.019).  

Raw data is provided on **zenodo** (IMC: <https://zenodo.org/records/10890543>, IF: <https://zenodo.org/records/10942403>, R data: <https://zenodo.org/records/15304181>) and **GEO** (scRNA-seq organoid data - GEO Series number: GSE298343). Here, we will download [SingleCellExperiment](https://bioconductor.org/packages//SingleCellExperiment.html) and [CytoImageList](https://bioconductor.org/packages/release/bioc/vignettes/cytomapper/inst/doc/cytomapper.html) objects from [zenodo](https://zenodo.org/records/15304181). 

# **Settings**

```{r libraries, include = FALSE}
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
dataset_name <- "Meyer_2025_TripleNegativeBreastCancer"
dataset_version <- "v1"
cat("Dataset version:", dataset_version)
```

Setting the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
dataset_dir <- file.path(".", dataset_name)
if(!(dir.exists(dataset_dir))) dir.create(dataset_dir)

outdir <- file.path(dataset_dir, dataset_version)
if(!(dir.exists(outdir))) dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


# **Single cell data**

We will download single-cell data corresponding from
[@Meyer-2025-TripleNegativeBreastCancer] from 
[zenodo](https://zenodo.org/records/15304181).

## Download SingleCellExperiment object

### Import function

Function to download and read RDS files.

```{r import-function}
importAndReadData <- function(url, output_dir, filename) {
  # Create full path
  file_path <- file.path(output_dir, filename)
  
  # Download the RDS file
  download.file(url, destfile = file_path, mode = "wb")
  
  # Read and return the RDS object
  readRDS(file_path)
}
```


## Update SingleCellExperiment

### Download

Here, we download a `SingleCellExperiment` object from zenodo that contains all single-cell, sample and patient metadata. 

```{r download-sce}
# Download SCE
# TODO: Update record
url_sce <- "https://zenodo.org/record/15396067/files/sce_ALL.rds?download=1"

sce <- importAndReadData(url_sce, workdir, "sce_ALL.rds")
```

### colData

```{r format-col}
colData <- as.data.table(colData(sce))
```

Here, we will rename and select columns of the `colData` entry in the downloaded `SingleCellExperiment` object, which will constitute the final version. 

Columns are renamed for consistency with the other datasets.

```{r rename-col}
# Rename columns
colData_new <- colData[, .(
  image_name = sample_id,        
  image_width = width_px,
  image_height = height_px,
  image_TMA_number = TMANr,
  image_n = n_samples, 
  image_tissue_area = tissue_area,
  
  cell_number = ObjectNumber,
  cell_id = cell_id,
  cell_x = Pos_X,
  cell_y = Pos_Y,
  cell_area = area,
  cell_major_axis_length = major_axis_length,
  cell_minor_axis_length = minor_axis_length,
  cell_eccentricity = eccentricity,
  cell_n_neighbors = n_neighbors,
  cell_category_tumor_nontumor = analysis_cat,
  cell_metacluster = metacluster_combined,
  cell_CD8Tcell_spatialpattern = CD8Tcell_type,
  cell_Bcell_patch = patch_B_cell,
  cell_Ki67_status = GMM_Ki67,
  
  patient_id = PID,
  patient_age = age,
  patient_age_bin = age_bin,
  patient_OS_status = status_OS,
  patient_OS_months = OS_months,
  patient_DFS_status = status_DFS,
  patient_DFS_months = DFS_months,
  patient_pretreatment = pretreatment,
  patient_Bcell_score = B_cell_score,
  patient_CD8Tcell_spatialgroup = spatial_CD8_pheno,
  patient_CK5CK7_group = CK5_7_group_name,
  patient_patientgroup = BL_CD8_group_merg,
  patient_riskgroup = PID_group_risk,
  
  tumor_size_cm = tumor_size_cm,
  tumor_grade = grade,
  tumor_PTNM_T = pT,
  tumor_PTNM_T_simple = pT_simple, 
  tumor_PTNM_N = pN,
  tumor_PTNM_N_simple = pN_simple,
  tumor_PTNM_M = pM,
  tumor_PTNM_M_simple = pM_simple
)]

rownames(colData_new) <- colData_new$cell_id
```

### rowData

```{r format-row}
rowData <- as.data.table(rowData(sce))
```

Here, we will rename and select columns of the `rowData` entry in the downloaded `SingleCellExperiment` object, which will constitute the final version. 

Columns are renamed for consistency with the other datasets.

```{r rename-row}
rowData_new <- rowData

rowData_new$metal <- rowData_new$channel
rowData_new$channel <- rownames(rowData_new)
rowData_new$channel_ilastik <- rowData_new$ilastik
rowData_new$channel_deepcell <- rowData_new$deepcell
rowData_new$channel_astir <- rowData_new$use_astir
rowData_new$channel_cluster_TME <- rowData_new$TME_cluster_channel
rowData_new$channel_cluster_tumor <- rowData_new$tumor_cluster_channel
rowData_new$channel_cluster_combined <- rowData_new$cluster_channel_combined

# Update metal and protein info
rowData_new[clean_target == "HistoneH3", `:=` (
  name = "Histone H3", 
  full_name = "Histone H3",
  short_name = "H3"
  )]

rowData_new[clean_target == "H3K27me3", `:=`(
  name = "H3K27me3",
  full_name = "trimethyl-Histone H3 [K27]",
  short_name = "trime_H3"
  )]

rowData_new[clean_target == "CK5", `:=`(
  name = "CK5",
  full_name = "Cytokeratin 5",
  short_name = "CK5"
  )]

rowData_new[clean_target == "Fibronectin", `:=`(
  name = "Fibronectin",
  full_name = "Fibronectin",
  short_name = "Fibronectin"
  )]

rowData_new[clean_target == "HLADR", `:=`(
  name = "HLADR",
  full_name = "HLADR",
  short_name = "HLADR"
  )]

rowData_new[clean_target == "CK8_18", `:=`(
  name = "CK8/18",
  full_name = "Cytokeratin 8/18",
  short_name = "CK8_18"
  )]

rowData_new[clean_target == "CD15", `:=`(
  name = "CD15",
  full_name = "CD15",
  short_name = "CD15"
  )]

rowData_new[clean_target == "CD68", `:=`(
  name = "CD68",
  full_name = "CD68",
  short_name = "CD68"
  )]

rowData_new[clean_target == "KRT14", `:=`(
  name = "CK14",
  full_name = "Cytokeratin 14",
  short_name = "CK14"
  )]

rowData_new[clean_target == "SMA", `:=`(
  name = "SMA",
  full_name = "Smooth muscle actin",
  short_name = "SMA"
  )]

rowData_new[clean_target == "Vimentin", `:=`(
  name = "Vimentin",
  full_name = "Vimentin",
  short_name = "Vimentin"
  )]

rowData_new[clean_target == "cMYC", `:=`(
  name = "cMYC",
  full_name = "cMYC",
  short_name = "cMYC"
  )]

rowData_new[clean_target == "HER2", `:=`(
  name = "HER2",
  full_name = "Epidermal growth factor receptor-2",
  short_name = "HER2"
  )]

rowData_new[clean_target == "CD3", `:=`(
  name = "CD3",
  full_name = "CD3",
  short_name = "CD3"
  )]

rowData_new[clean_target == "pH3", `:=`(
  name = "p-H3",
  full_name = "phospho-Histone H3 [S28]",
  short_name = "p_H3"
  )]

rowData_new[clean_target == "CD11c", `:=`(
  name = "CD11c",
  full_name = "CD11c",
  short_name = "CD11c"
  )]

rowData_new[clean_target == "Bcl2", `:=`(
  name = "Bcl2",
  full_name = "Bcl2",
  short_name = "Bcl2"
  )]

rowData_new[clean_target == "ER", `:=`(
  name = "ER",
  full_name = "Estrogen receptor",
  short_name = "ER"
  )]

rowData_new[clean_target == "PR", `:=`(
  name = "PR",
  full_name = "Progesterone receptor",
  short_name = "PR"
  )]

rowData_new[clean_target == "p53", `:=`(
  name = "p53",
  full_name = "p53",
  short_name = "p53"
  )]

rowData_new[clean_target == "CD44", `:=`(
  name = "CD44",
  full_name = "CD44",
  short_name = "CD44"
  )]

rowData_new[clean_target == "AR", `:=`(
  name = "AR",
  full_name = "Androgen receptor",
  short_name = "AR"
  )]

rowData_new[clean_target == "CD45", `:=`(
  name = "CD45",
  full_name = "CD45",
  short_name = "CD45"
  )]

rowData_new[clean_target == "GATA3", `:=`(
  name = "GATA3",
  full_name = "GATA3",
  short_name = "GATA3"
  )]

rowData_new[clean_target == "CD20", `:=`(
  name = "CD20",
  full_name = "CD20",
  short_name = "CD20"
  )]

rowData_new[clean_target == "CD8a", `:=`(
  name = "CD8a",
  full_name = "CD8 alpha",
  short_name = "CD8a"
  )]

rowData_new[clean_target == "CAIX", `:=`(
  name = "CAIX",
  full_name = "Carbonic anhydrase IX",
  short_name = "CA9"
  )]

rowData_new[clean_target == "Ecad", `:=`(
  name = "Ecad",
  full_name = "E-Cadherin/P-Cadherin",
  short_name = "Ecad"
  )]

rowData_new[clean_target == "Ki67", `:=`(
  name = "Ki67",
  full_name = "Ki67",
  short_name = "Ki67"
  )]

rowData_new[clean_target == "EGFR", `:=`(
  name = "EGFR",
  full_name = "Epidermal growth factor receptor",
  short_name = "EGFR"
  )]

rowData_new[clean_target == "S6", `:=`(
  name = "p-S6",
  full_name = "phospho-S6 [S235/S236]",
  short_name = "p_S6"
  )]

rowData_new[clean_target == "CD4", `:=`(
  name = "CD4",
  full_name = "CD4",
  short_name = "CD4"
  )]

rowData_new[clean_target == "CD31_vWF", `:=`(
  name = "CD31/vWF",
  full_name = "CD31 + von Willebrand factor",
  short_name = "CD31_vWF"
  )]

rowData_new[clean_target == "p_mTOR", `:=`(
  name = "p-mTOR",
  full_name = "phospho-mTOR [S2448]",
  short_name = "p_mTOR"
  )]

rowData_new[clean_target == "CK7", `:=`(
  name = "CK7",
  full_name = "Cytokeratin 7",
  short_name = "CK7"
  )]

rowData_new[clean_target == "panCK", `:=`(
  name = "panCK",
  full_name = "Pan-cytokeratin",
  short_name = "panCK"
  )]

rowData_new[clean_target == "c_Cas3_PARP", `:=`(
  name = "cCas3/PARP",
  full_name = "cleaved-Caspase3 + cleaved-PARP",
  short_name = "cCASP3_cPARP"
  )]

rowData_new[clean_target == "DNA_1", `:=`(
  name = "DNA1",
  full_name = "Iridium 191",
  short_name = "DNA1"
  )]

rowData_new[clean_target == "DNA_2", `:=`(
  name = "DNA2",
  full_name = "Iridium 193",
  short_name = "DNA2"
  )]

# Keep specific columns
keep_vector <- c("channel", "metal", "name", "full_name", "short_name", "channel_ilastik", "channel_deepcell", "channel_astir", "channel_cluster_TME", "channel_cluster_tumor", "channel_cluster_combined")

rowData_new <- rowData_new[, ..keep_vector]

rownames(rowData_new) <- rowData_new$short_name
```


### assays 

We adjust the `assay` slots of the `SingleCellExperiment` to match the updated `colData` and `rowData` information. 

```{r assay}
# Raw mean ion counts per cell
counts <- assay(sce, "counts")

# Arcsinh-transformed counts (cofactor of 1)
exprs <- assay(sce, "exprs")

# Min-max normalized arcsinh-transformed counts
min_max <- assay(sce, "min_max")

rownames(counts) <- rownames(exprs) <- rownames(min_max) <- rownames(rowData_new)
colnames(counts) <- colnames(exprs) <- colnames(min_max) <- rownames(colData_new)
```

### colPairs 

We extract the `colPairs` slot of the `SingleCellExperiment` with the steinbock-based neighborhood graph. 

```{r colpair}
# steinbock based neighborhood graph based on pixel expansion
neighborhood <- colPairs(sce)[1]
```

### Combine

We generate the final `SingleCellExperiment` object.

```{r final-sce}
# Create final SCE object based on updated entries
sce_new <- SingleCellExperiment(
  assays = list(counts = counts, exprs = exprs, min_max = min_max),
  rowData = rowData_new,
  colData = colData_new,
  colPairs = neighborhood
)
```


### Subset

We also save subsets of the `SingleCellExperiment` objects. The subset corresponds to 125 images from 60 patients sampled across proposed patient groups. This includes all images visually presented in the publication. 

```{r subset-sce}
# Patients selected 
patients_keep <- c("B07.17017", "B08.40119", "B12.26323", "B06.22442", "B06.38699", 
"B08.24635", "B09.04541", "B08.38569", "B09.09794", "B16.33745", 
"B09.30766", "B12.01576", "B06.11749", "B08.36428", "B11.17149", 
"B13.59896", "B07.11296", "B12.15315", "B07.22827", "B09.42460", 
"B05.41292", "B10.59762", "B16.21103", "B12.39340", "B08.28600", 
"B05.34236", "B06.25613", "B10.23573", "B12.05893", "B07.47034", 
"B12.35711", "B16.78991", "B17.15400", "B12.04518", "B07.29298", 
"B11.03307", "B07.45135", "B10.34352", "B06.02565", "B10.18650", 
"B10.03716", "B14.30715", "B08.33039", "B16.74059", "B07.41959", 
"B10.46561", "B08.42871", "B16.23579", "B10.02341", "B13.06581", 
"B10.45389", "B10.21451", "B17.16725", "B09.18572", "B12.50460", 
"B13.06577", "B07.08412", "B15.54100", "B08.07613", "B11.57286"
)

sce_new_sub <- sce_new[,sce_new$patient_id %in% patients_keep]
```

### Save on disk 

We save the `SingleCellExperiment` objects for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
# Full dataset
saveRDS(sce_new, file.path(outdir, "sce_full.rds"))
print(sce_new)

# Subset
saveRDS(sce_new_sub, file.path(outdir, "sce.rds"))
print(sce_new_sub)
```

### Clean up 

```{r clean-sce, results='hide'}
remove(colData, colData_new, counts, exprs, min_max, neighborhood, rowData, rowData_new, sce, keep_vector, url_sce, patients_keep)
```

# **Images and masks**

We will download multichannel images and corresponding cell segmentation masks from
[@Meyer-2025-TripleNegativeBreastCancer] from [zenodo](https://zenodo.org/records/15304181). Images and masks are stored as `CytoImageList` objects. The subsetted objects contain 125 images/masks from 60 patients sampled across proposed patient groups. This includes all images visually presented in the publication and corresponds to the subsetted `SingleCellExperiment` object. Due to memory space limitations, multichannel images for the full cohort are not available. 


## Download

```{r import-images-masks}
# # Download image subset
# # TODO: Add record number
# url_images_sub <- "https://zenodo.org/record/?/files/images_sub.rds?download=1"
# images_sub <- importAndReadData(url_images_sub, workdir, "images_sub.rds")
# 
# # Download masks and mask subset
# # TODO: Add record number
# url_masks_sub <- "https://zenodo.org/record/?/files/masks_sub.rds?download=1"
# masks_sub <- importAndReadData(url_masks_sub, workdir, "masks_sub.rds")
# 
# url_masks <- "https://zenodo.org/record/?/files/masks.rds?download=1"
# masks <- importAndReadData(url_masks, workdir, "masks.rds")
```


```{r}
# TODO: This chunk will be removed when record number is added
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_new.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")

masks <- masks[mcols(masks)$sample_id %in% sce_new$image_name]
#saveRDS(masks, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/masks.rds")

masks_sub <- masks[mcols(masks)$sample_id %in% sce_new_sub$image_name]
#saveRDS(masks_sub, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/masks_sub.rds")

images_sub <- images[mcols(images)$sample_id %in% sce_new_sub$image_name]
images_sub <- CytoImageList(images_sub, on_disk = FALSE)
#saveRDS(images_sub, "/mnt/central_nas/tnbc_volume/TNBC/data_analysis/00_final_files/OTHER/images_sub.rds")
```


### mcols

We will adjust and rename the `mcols` entries of the `CytoImageList` objects for consistency with the other datasets.

```{r mcols}
mcols(images_sub)$image_name <- mcols(images_sub)$sample_id
mcols(masks_sub)$image_name <- mcols(masks_sub)$sample_id
mcols(masks)$image_name <- mcols(masks)$sample_id

mcols(images_sub)$sample_id <- mcols(masks_sub)$sample_id <- mcols(masks)$sample_id <- NULL 
mcols(images_sub)$PID <- mcols(masks_sub)$PID <- mcols(masks)$PID <- NULL 

mcols(images_sub)$patient_id <- sce_new_sub$patient_id[match(mcols(images_sub)$image_name, sce_new_sub$image_name)]
mcols(masks_sub)$patient_id <- sce_new_sub$patient_id[match(mcols(masks_sub)$image_name, sce_new_sub$image_name)]
mcols(masks)$patient_id <- sce_new$patient_id[match(mcols(masks)$image_name, sce_new$image_name)]
```

### channelNames

We will adjust `channelNames` of the `CytoImageList` image subset to match the `SingleCellExperiment` object.

```{r channelnames}
# Update channelNames, so they match the SCE object
channelNames(images_sub) <- rownames(sce_new_sub)
```

## Save on disk

Finally, we will save the generated `CytoImageList` objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-images-masks}
# Masks and masks subset
saveRDS(masks, file.path(outdir, "masks_full.rds"))
print(head(masks))
#remove(masks)

saveRDS(masks_sub, file.path(outdir, "masks.rds"))
print(head(masks_sub))
#remove(masks_sub)

# Image subset
saveRDS(images_sub, file.path(outdir, "images.rds"))
print(head(images_sub))
#remove(images_sub)
```

## Clean up

```{r clean-rest, results='hide'}
# Remove all environment variables
# TODO: Remove images from this list
remove(images, images_sub, masks, masks_sub, sce_new, sce_new_sub)

# Reset original timeout value
options(timeout = timeout)
```


# **Session information**

```{r session-info}
sessionInfo()
```


# **References**