---
title: "Preparation of the Meyer et al dataset"
author: "Lasse Meyer"
date: "Created: 22 December 2025; Compiled: `r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcdatasets')`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: inline
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path("..", "extdata"))
```


# **Introduction**

This script downloads TODO images (around two images per patient) from 215 patients, as well as the associated single-cell data and cell segmentation masks from the triple-negative breast cancer Imaging Mass Cytometry (IMC) dataset described in the following publication:  

[Meyer et al. A stratification system for breast cancer based on basoluminal tumor cells and spatial tumor architecture, Cancer Cell (2025)](https://doi.org/10.1016/j.ccell.2025.06.019).  

Raw data is provided on **zenodo** (IMC: <https://zenodo.org/records/10890543>, IF: <https://zenodo.org/records/10942403>, R data: <https://zenodo.org/records/15304181>) and **GEO** (scRNA-seq organoid data - GEO Series number: GSE298343). Here, we will download a [SingleCellExperiment](https://bioconductor.org/packages//SingleCellExperiment.html) object from [zenodo](https://zenodo.org/records/15304181). 

We will then download the corresponding multichannel IMC images and cell segmentation masks and format them into `CytoImageList` objects using the [cytomapper](https://bioconductor.org/packages/cytomapper) package.


# **Settings**

```{r libraries, include = FALSE}
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(cytomapper)
```

```{r dataset-version, echo=FALSE}
dataset_name <- "Meyer_2025_TripleNegativeBreastCancer"
dataset_version <- "v1"
cat("Dataset version:", dataset_version)
```

Setting the working and output directories

```{r directories}
# Temporary directory to unzip files
workdir <- tempdir()
Sys.setenv(workdir = workdir)

# Output directory
dataset_dir <- file.path(".", dataset_name)
if(!(dir.exists(dataset_dir))) dir.create(dataset_dir)

outdir <- file.path(dataset_dir, dataset_version)
if(!(dir.exists(outdir))) dir.create(outdir)

# Increase timeout period so that large files can be downloaded
timeout <- getOption('timeout')
options(timeout = 1000)
```


# **Single cell data**

We will download single-cell data corresponding from
[@Meyer-2025-TripleNegativeBreastCancer] from 
[zenodo](https://zenodo.org/records/15304181).

## Download SingleCellExperiment object

### Import function

Function to download and read RDS files.

```{r import-function}
importAndReadData <- function(url, output_dir, filename) {
  # Create full path
  file_path <- file.path(output_dir, filename)
  
  # Download the RDS file
  download.file(url, destfile = file_path, mode = "wb")
  
  # Read and return the RDS object
  readRDS(file_path)
}
```

### Object download

TODO: The first zip folder contains the main single-cell, sample and patient metadata. The single-cell locations and cluster labels are downloaded as separate zip folders because they were uploaded to zenodo separately at a later time point.

```{r}
# Download SCE
url_sce <- "https://zenodo.org/record/15396067/files/sce_ALL.rds?download=1"

sce <- importAndReadData(url_sce, workdir, "sce_ALL.rds")
```

## Update SingleCellExperiment

### colData

```{r}
colData <- as.data.table(colData(sce))
```

Here, we will rename and select columns of the `colData` entry in the downloaded `SingleCellExperiment` object, which will constitute the final version. 

Columns are renamed for consistency with the other datasets.

```{r}
# Rename columns
colData_new <- colData[, .(
  image_name = sample_id,        
  image_width = width_px,
  image_height = height_px,
  image_TMA_number = TMANr,
  image_n = n_samples, 
  image_tissue_area = tissue_area,
  
  cell_number = ObjectNumber,
  cell_id = cell_id,
  cell_x = Pos_X,
  cell_y = Pos_Y,
  cell_area = area,
  cell_major_axis_length = major_axis_length,
  cell_minor_axis_length = minor_axis_length,
  cell_eccentricity = eccentricity,
  cell_n_neighbors = n_neighbors,
  cell_category_tumor_nontumor = analysis_cat,
  cell_metacluster = metacluster_combined,
  cell_CD8Tcell_spatialpattern = CD8Tcell_type,
  cell_Bcell_patch = patch_B_cell,
  cell_Ki67_status = GMM_Ki67,
  
  patient_id = PID,
  patient_age = age,
  patient_age_bin = age_bin,
  patient_OS_status = status_OS,
  patient_OS_months = OS_months,
  patient_DFS_status = status_DFS,
  patient_DFS_months = DFS_months,
  patient_pretreatment = pretreatment,
  patient_Bcell_score = B_cell_score,
  patient_CD8Tcell_spatialgroup = spatial_CD8_pheno,
  patient_CK5CK7_group = CK5_7_group_name,
  patient_patientgroup = BL_CD8_group_merg,
  patient_riskgroup = PID_group_risk,
  
  tumor_size_cm = tumor_size_cm,
  tumor_grade = grade,
  tumor_PTNM_T = pT,
  tumor_PTNM_T_simple = pT_simple, 
  tumor_PTNM_N = pN,
  tumor_PTNM_N_simple = pN_simple,
  tumor_PTNM_M = pM,
  tumor_PTNM_M_simple = pM_simple
)]

rownames(colData_new) <- colData_new$cell_id
```

### rowData

```{r}
rowData <- as.data.table(rowData(sce))
```

Here, we will rename and select columns of the `rowData` entry in the downloaded `SingleCellExperiment` object, which will constitute the final version. 

Columns are renamed for consistency with the other datasets.

```{r}
rowData_new <- rowData

rowData_new$metal <- rowData_new$channel
rowData_new$channel <- rownames(rowData_new)
rowData_new$channel_ilastik <- rowData_new$ilastik
rowData_new$channel_deepcell <- rowData_new$deepcell
rowData_new$channel_astir <- rowData_new$use_astir
rowData_new$channel_cluster_TME <- rowData_new$TME_cluster_channel
rowData_new$channel_cluster_tumor <- rowData_new$tumor_cluster_channel
rowData_new$channel_cluster_combined <- rowData_new$cluster_channel_combined

# Update metal and protein info
rowData_new[clean_target == "HistoneH3", `:=` (
  name = "Histone H3", 
  full_name = "Histone H3",
  short_name = "H3"
  )]

rowData_new[clean_target == "H3K27me3", `:=`(
  name = "H3K27me3",
  full_name = "trimethyl-Histone H3 [K27]",
  short_name = "trime_H3"
  )]

rowData_new[clean_target == "CK5", `:=`(
  name = "CK5",
  full_name = "Cytokeratin 5",
  short_name = "CK5"
  )]

rowData_new[clean_target == "Fibronectin", `:=`(
  name = "Fibronectin",
  full_name = "Fibronectin",
  short_name = "Fibronectin"
  )]

rowData_new[clean_target == "HLADR", `:=`(
  name = "HLADR",
  full_name = "HLADR",
  short_name = "HLADR"
  )]

rowData_new[clean_target == "CK8_18", `:=`(
  name = "CK8/18",
  full_name = "Cytokeratin 8/18",
  short_name = "CK8_18"
  )]

rowData_new[clean_target == "CD15", `:=`(
  name = "CD15",
  full_name = "CD15",
  short_name = "CD15"
  )]

rowData_new[clean_target == "CD68", `:=`(
  name = "CD68",
  full_name = "CD68",
  short_name = "CD68"
  )]

rowData_new[clean_target == "KRT14", `:=`(
  name = "CK14",
  full_name = "Cytokeratin 14",
  short_name = "CK14"
  )]

rowData_new[clean_target == "SMA", `:=`(
  name = "SMA",
  full_name = "Smooth muscle actin",
  short_name = "SMA"
  )]

rowData_new[clean_target == "Vimentin", `:=`(
  name = "Vimentin",
  full_name = "Vimentin",
  short_name = "Vimentin"
  )]

rowData_new[clean_target == "cMYC", `:=`(
  name = "cMYC",
  full_name = "cMYC",
  short_name = "cMYC"
  )]

rowData_new[clean_target == "HER2", `:=`(
  name = "HER2",
  full_name = "Epidermal growth factor receptor-2",
  short_name = "HER2"
  )]

rowData_new[clean_target == "CD3", `:=`(
  name = "CD3",
  full_name = "CD3",
  short_name = "CD3"
  )]

rowData_new[clean_target == "pH3", `:=`(
  name = "p-H3",
  full_name = "phospho-Histone H3 [S28]",
  short_name = "p_H3"
  )]

rowData_new[clean_target == "CD11c", `:=`(
  name = "CD11c",
  full_name = "CD11c",
  short_name = "CD11c"
  )]

rowData_new[clean_target == "Bcl2", `:=`(
  name = "Bcl2",
  full_name = "Bcl2",
  short_name = "Bcl2"
  )]

rowData_new[clean_target == "ER", `:=`(
  name = "ER",
  full_name = "Estrogen receptor",
  short_name = "ER"
  )]

rowData_new[clean_target == "PR", `:=`(
  name = "PR",
  full_name = "Progesterone receptor",
  short_name = "PR"
  )]

rowData_new[clean_target == "p53", `:=`(
  name = "p53",
  full_name = "p53",
  short_name = "p53"
  )]

rowData_new[clean_target == "CD44", `:=`(
  name = "CD44",
  full_name = "CD44",
  short_name = "CD44"
  )]

rowData_new[clean_target == "AR", `:=`(
  name = "AR",
  full_name = "Androgen receptor",
  short_name = "AR"
  )]

rowData_new[clean_target == "CD45", `:=`(
  name = "CD45",
  full_name = "CD45",
  short_name = "CD45"
  )]

rowData_new[clean_target == "GATA3", `:=`(
  name = "GATA3",
  full_name = "GATA3",
  short_name = "GATA3"
  )]

rowData_new[clean_target == "CD20", `:=`(
  name = "CD20",
  full_name = "CD20",
  short_name = "CD20"
  )]

rowData_new[clean_target == "CD8a", `:=`(
  name = "CD8a",
  full_name = "CD8 alpha",
  short_name = "CD8a"
  )]

rowData_new[clean_target == "CAIX", `:=`(
  name = "CAIX",
  full_name = "Carbonic anhydrase IX",
  short_name = "CA9"
  )]

rowData_new[clean_target == "Ecad", `:=`(
  name = "Ecad",
  full_name = "E-Cadherin/P-Cadherin",
  short_name = "Ecad"
  )]

rowData_new[clean_target == "Ki67", `:=`(
  name = "Ki67",
  full_name = "Ki67",
  short_name = "Ki67"
  )]

rowData_new[clean_target == "EGFR", `:=`(
  name = "EGFR",
  full_name = "Epidermal growth factor receptor",
  short_name = "EGFR"
  )]

rowData_new[clean_target == "S6", `:=`(
  name = "p-S6",
  full_name = "phospho-S6 [S235/S236]",
  short_name = "p_S6"
  )]

rowData_new[clean_target == "CD4", `:=`(
  name = "CD4",
  full_name = "CD4",
  short_name = "CD4"
  )]

rowData_new[clean_target == "CD31_vWF", `:=`(
  name = "CD31/vWF",
  full_name = "CD31 + von Willebrand factor",
  short_name = "CD31_vWF"
  )]

rowData_new[clean_target == "p_mTOR", `:=`(
  name = "p-mTOR",
  full_name = "phospho-mTOR [S2448]",
  short_name = "p_mTOR"
  )]

rowData_new[clean_target == "CK7", `:=`(
  name = "CK7",
  full_name = "Cytokeratin 7",
  short_name = "CK7"
  )]

rowData_new[clean_target == "panCK", `:=`(
  name = "panCK",
  full_name = "Pan-cytokeratin",
  short_name = "panCK"
  )]

rowData_new[clean_target == "c_Cas3_PARP", `:=`(
  name = "cCas3/PARP",
  full_name = "cleaved-Caspase3 + cleaved-PARP",
  short_name = "cCASP3_cPARP"
  )]

rowData_new[clean_target == "DNA_1", `:=`(
  name = "DNA1",
  full_name = "Iridium 191",
  short_name = "DNA1"
  )]

rowData_new[clean_target == "DNA_2", `:=`(
  name = "DNA2",
  full_name = "Iridium 193",
  short_name = "DNA2"
  )]

# Keep specific columns
keep_vector <- c("channel", "metal", "name", "full_name", "short_name", "channel_ilastik", "channel_deepcell", "channel_astir", "channel_cluster_TME", "channel_cluster_tumor", "channel_cluster_combined")

rowData_new <- rowData_new[, ..keep_vector]

rownames(rowData_new) <- rowData_new$short_name
```


### assays 

```{r}
# Raw mean ion counts per cell
counts <- assay(sce, "counts")

# Arcsinh-transformed counts (cofactor of 1)
exprs <- assay(sce, "exprs")

# Min-max normalized arcsinh-transformed counts
min_max <- assay(sce, "min_max")

rownames(counts) <- rownames(exprs) <- rownames(min_max) <- rownames(rowData_new)
colnames(counts) <- colnames(exprs) <- colnames(min_max) <- rownames(colData_new)
```

### colPairs 

```{r}
# steinbock based neighborhood graph based on pixel expansion
neighborhood <- colPairs(sce)[1]
```

### Combine

```{r}
# Create final SCE object based on updated entries
sce_new <- SingleCellExperiment(
  assays = list(counts = counts, exprs = exprs, min_max = min_max),
  rowData = rowData_new,
  colData = colData_new,
  colPairs = neighborhood
)
```


### Subset

TODO: We also save subsets of the `SingleCellExperiment` objects. 

The subset corresponds to 125 images from 60 patients sampled across proposed patient groups. This includes all images visually presented in the publication. 

```{r}
# Patients selected 
patients_keep <- c("B07.17017", "B08.40119", "B12.26323", "B06.22442", "B06.38699", 
"B08.24635", "B09.04541", "B08.38569", "B09.09794", "B16.33745", 
"B09.30766", "B12.01576", "B06.11749", "B08.36428", "B11.17149", 
"B13.59896", "B07.11296", "B12.15315", "B07.22827", "B09.42460", 
"B05.41292", "B10.59762", "B16.21103", "B12.39340", "B08.28600", 
"B05.34236", "B06.25613", "B10.23573", "B12.05893", "B07.47034", 
"B12.35711", "B16.78991", "B17.15400", "B12.04518", "B07.29298", 
"B11.03307", "B07.45135", "B10.34352", "B06.02565", "B10.18650", 
"B10.03716", "B14.30715", "B08.33039", "B16.74059", "B07.41959", 
"B10.46561", "B08.42871", "B16.23579", "B10.02341", "B13.06581", 
"B10.45389", "B10.21451", "B17.16725", "B09.18572", "B12.50460", 
"B13.06577", "B07.08412", "B15.54100", "B08.07613", "B11.57286"
)

sce_new_sub <- sce_new[,sce_new$patient_id %in% patients_keep]
```



### Save on disk 

We save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save-sce}
# Full dataset
saveRDS(sce_new, file.path(outdir, "sce_full.rds"))
print(sce_new)

# Subset
saveRDS(sce_new_sub, file.path(outdir, "sce.rds"))
print(sce_new_sub)
```



### Clean up 

```{r}
remove(colData, colData_new, counts, exprs, min_max, neighborhood, rowData, rowData_new, sce, keep_vector, url_sce, patients_keep)
```


# CONTINUE HERE 

# **Images and masks**

TODO Here, we will download multichannel images from [@JacksonFischer-2020-BreastCancer], as well as the corresponding cell segmentation masks. Images and masks correspond to the data in the `SingleCellExperiment` object and will be formatted into `CytoImageList` objects.

## Download images and masks

We download the zip file containing images and masks.

```{r import-images, results='hide'}
# TODO: Download images and masks
#url_images <- ("https://zenodo.org/record/4607374/files/OMEandSingleCellMasks.zip?download=1")
#importData(url_images, workdir, "OMEandSingleCellMasks.zip")
```

```{r}
images <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/images_TNBC_new.rds")
masks <- readRDS("/mnt/central_nas/tnbc_volume/TNBC/data_analysis/masks_segtest_4.rds")
```


## Cell segmentation masks

```{r}


mcols(images)


images_sub <- images[mcols(images)$sample_id %in% sce_new$image_name]
images_sub_onDisk <- CytoImageList(images_sub, on_disk = FALSE)

masks_sub <- masks[mcols(masks)$sample_id %in% sce_new$image_name]




channelNames(images)

#images_onDisk <- CytoImageList(images, on_disk = FALSE)

imagesimagesTRUE
```





## Multichannel images 

```{r}

```

### Add mask names

Next, we add image names to masks objects, these names correspond to the 
`image_name` column in `colData(sce)`. This information is stored in the 
metadata columns of the `CytoImageList` objects and is used by `cytomapper` to 
match single cell data, images and masks.

```{r add-mask-names}
names(masks) <- gsub("_full_maks", "_full.tiff", names(masks))
mcols(masks)$image_filename <- names(masks)
masks <- masks[names(masks) %in% sce$image_filename]

# Add image metadata
mcols(masks) <- merge(
  mcols(masks),
  unique(colData(sce)[, c(
      "image_filename", "image_name", "image_number")]),
  by = "image_filename")
```

### Create cohort-specific subsets

We create two objects containing the masks from the Basel and Zurich cohorts.

We subset the masks to the hundred images contained in the
`SingleCellExperiment` objects.

```{r subset-masks-cohorts}
# Basel cohort
masks_basel <- masks[grep("BaselTMA", names(masks))]
masks_basel <- masks_basel[names(masks_basel) %in% unique(
    sce_basel$image_filename)]

names(masks_basel) <- mcols(masks_basel)$image_name

# Zurich cohort
masks_zurich <- masks[grep("ZTMA", names(masks))]
masks_zurich <- masks_zurich[names(masks_zurich) %in% unique(
    sce_zurich$image_filename)]

names(masks_zurich) <- mcols(masks_zurich)$image_name

# All masks
names(masks) <- mcols(masks)$image_name
```

### Save on disk

Finally, we will save the generated `CytoImageList` objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-masks}
# All masks
saveRDS(masks, file.path(outdir, "masks_full.rds"))
print(head(masks))
remove(masks)

# Basel cohort
saveRDS(masks_basel, file.path(outdir, "masks_basel.rds"))
print(head(masks_basel))

# Zurich cohort
saveRDS(masks_zurich, file.path(outdir, "masks_zurich.rds"))
print(head(masks_zurich))
```


## Multichannel images

We will import and process multichannel images to make them compatible with the `cytomapper` package.

For memory space reasons, we will generate one image subset for the Basel cohort and one for the Zurich cohort. The images will be subsetted to correspond to the single cell data in the `sce_basel` and `sce_zurich` objects. We will first process images from the Basel cohort and then repeat the same steps for the Zurich cohort.

### Import images

We unzip images and read them into a `CytoImageList` object.

```{r unzip-images-basel, results='hide'}
fn <- file.path("OMEnMasks", "ome.zip")
system2("unzip", args = c("-o", file.path(workdir, fn),
                          "*BaselTMA_*.tiff",
                          "-d", workdir),
        stdout = TRUE)
```

Loading all tiffs would require too much memory, so we delete all the `tiff` files that we do not want to keep. We then use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r load-images-basel, results='hide', warning=FALSE}
tiffs_all <- list.files(file.path(workdir, "ome/"))
tiff_delete <- tiffs_all[!tiffs_all %in% sce_basel$image_filename]
file.remove(file.path(workdir, "ome/", tiff_delete))

# Load the images as a CytoImageList object
images_basel <- loadImages(file.path(workdir, "ome/"), pattern = "_full.tiff")
file.remove(file.path(workdir, "ome/", tiffs_all))
```

### Add image names and numbers

Next, we add image names to image objects, these names correspond to the 
`image_name` column in `colData(sce)`. This information is stored in the 
metadata columns of the `CytoImageList` objects and is used by `cytomapper` 
to match single cell data, images and masks.

```{r add-image-names-basel}
mcols(images_basel)$image_filename <- paste0(names(images_basel), ".tiff")
images_basel <- images_basel[
    mcols(images_basel)$image_filename %in% sce_basel$image_filename]
mcols(images_basel) <- merge(
  mcols(images_basel),
  unique(colData(sce_basel)[, c(
      "image_filename", "image_name", "image_number")]),
  by = "image_filename")
```

### Subset channels

Here, we exclude image channels that are not present in the `SingleCellExperiment` object.

```{r subset-channels-basel}
panel <- rowData(sce)
panel <- panel[order(panel$channel), ]

images_basel <- getChannels(images_basel, panel$channel)
gc()
```

### Add channel names

Finally, we will add protein short names as channel names of the `images` 
object. These short names correspond to the row names of the 
`SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names-basel}
channelNames(images_basel) <- rownames(panel)
```

### Check image names

We make sure that image names and masks names are the same so that they can be matched with `cytomapper`.

```{r check-mask-image-names-basel}
print(identical(mcols(masks_basel)$image_name,
                mcols(images_basel)$image_name))
names(images_basel) <- mcols(images_basel)$image_name
```

### Save on disk

Finally, we save the generated `CytoImageList` images for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-images-basel}
saveRDS(images_basel, file.path(outdir, "images_basel.rds"))
print(head(images_basel))

remove(images_basel)
gc()
```

### Zurich cohort

To import and format images from the Zurich cohort, we perform the exact same steps as we just did for the Basel cohort.

Unzip images

```{r unzip-images-zurich, results='hide'}
system2("unzip", args = c("-o", file.path(workdir, fn),
                          "*ZTMA*.tiff",
                          "-d", workdir),
        stdout = TRUE)
```

Loading all tiffs would require too much memory, so we delete all the `tiff` files that we do not want to keep. We then use the `loadImages` function of the `cytomapper` package to read the images into a `CytoImageList` object.

```{r load-images-zurich, results='hide', warning=FALSE}
tiffs_all <- list.files(file.path(workdir, "ome/"))
tiff_delete <- tiffs_all[!tiffs_all %in% sce_zurich$image_filename]
file.remove(file.path(workdir, "ome/", tiff_delete))

# Load the images as a CytoImageList object
images_zurich <- loadImages(file.path(workdir, "ome/"), pattern = "_full.tiff")
file.remove(file.path(workdir, "ome/", tiffs_all))
```

Next, we add image names to image objects, these names correspond to the 
`image_name` column in `colData(sce)`. This information is stored in the 
metadata columns of the `CytoImageList` objects and is used by `cytomapper` 
to match single cell data, images and masks.

```{r add-image-names-zurich}
mcols(images_zurich)$image_filename <- paste0(names(images_zurich), ".tiff")
images_zurich <- images_zurich[
    mcols(images_zurich)$image_filename %in% sce_zurich$image_filename]
mcols(images_zurich) <- merge(
  mcols(images_zurich),
  unique(colData(sce_zurich)[, c(
      "image_filename", "image_name", "image_number")]),
  by = "image_filename")
```

Here, we exclude image channels that are not present in the `SingleCellExperiment` object.

```{r subset-channels-zurich}
panel <- rowData(sce)
panel <- panel[order(panel$channel), ]

images_zurich <- getChannels(images_zurich, panel$channel)
gc()
```

We add protein short names as channel names of the `images` 
object. These short names correspond to the row names of the 
`SingleCellExperiment` object and to the `short_name` column of `rowData(sce)`.

```{r add-channel-names-zurich}
channelNames(images_zurich) <- rownames(panel)
```

We make sure that image names and masks names are the same so that they can be matched with `cytomapper`.

```{r check-mask-image-names-zurich}
print(identical(mcols(images_zurich)$image_name,
                mcols(masks_zurich)$image_name))
names(images_zurich) <- mcols(images_zurich)$image_name
```

Finally, we save the generated `CytoImageList` images and masks objects for uploading to `r Biocpkg("ExperimentHub")`.

```{r save-images-zurich}
saveRDS(images_zurich, file.path(outdir, "images_zurich.rds"))
print(head(images_zurich))

remove(images_zurich)
```

## Clean up

Remove all files from the temporary working directory.

```{r clean-up, results='hide'}
downloaded_files <- list.files(workdir)
downloaded_files <- downloaded_files[!downloaded_files %in% "BiocStyle"]
unlink(file.path(workdir, downloaded_files), recursive = TRUE)

# Reset original timeout value
options(timeout = timeout)
```


# **Session information**

```{r session-info}
sessionInfo()
```


# **References**