---
title: "Loading single cell data, images and masks from the Zanotelli et al dataset"
author: "Nicolas Damond"
date: "Created: Apr 06, 2021; Compiled: `r BiocStyle::doc_date()`"
output:
  BiocStyle::html_document:
    titlecaps: false
    toc_float: true
editor_options:
  chunk_output_type: inline
bibliography: "`r system.file('scripts', 'ref.bib', package='imcdatasets')`"
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

Load required R packages

```{r load-packages, message=FALSE}
library(S4Vectors)
library(SingleCellExperiment)
library(data.table)
library(cytomapper)
library(igraph)
```

# **Introduction**

This script downloads 517 images, as well as the associated single-cell data and cell segmentation masks from the Imaging Mass Cytometry (IMC) cell line spheroid dataset, which is available from zenodo: https://doi.org/10.5281/zenodo.4271910. This dataset is associated with the following publication:  

[Zanotelli et al. A quantitative analysis of the interplay of environment, neighborhood, and cell state in 3D spheroids. Mol Syst Biol (2020)16:e9798](https://doi.org/10.15252/msb.20209798)  

The code used to generate this dataset is available from the [SpheroidPublication GitHub repository]https://github.com/BodenmillerGroup/SpheroidPublication). Additional data associated with the same publication is available from the zenodo parent repository: https://doi.org/10.5281/zenodo.4055781.  

The current script downloads the processed single-cell data from zenodo and processes these data to create a [SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) object. The [cytomapper](https://www.bioconductor.org/packages/release/bioc/html/cytomapper.html) package is used to load the multichannel IMC images and cell segmentation masks and save them as `CytoImageList` objects.


# **Download the dataset**

The cell lines dataset from [@zanotelli2020spheroids] is downloaded from https://zenodo.org/record/4271910#.YGWWHD8kz-j.

## Set working and output directories

```{r directories}
workdir <- tempdir()
Sys.setenv(workdir = workdir)

outdir <- file.path("inst", "extdata", "zanotelli-spheroids-2020")
if(!(dir.exists(outdir)))
    dir.create(outdir)
```

## Download the dataset

```{r download}
timeout <- getOption('timeout')
options(timeout=1000)
url_dat <- ("https://zenodo.org/record/4271910/files/phys_analysis_export_v3.zip")

download.file(url_dat, destfile = file.path(workdir, "CellLines.zip"))
unzip(file.path(workdir, "CellLines.zip"), exdir = workdir)
file.remove(file.path(workdir, "CellLines.zip"))

# List unzipped files
writeLines(list.files(workdir))
```


# **Single-cell data**

## Import and format single-cell data

### Read-in the data

```{r read-in-data}
cell_X <- fread(file.path(workdir, "cell_X.csv"))
cell_obs <- fread(file.path(workdir, "cell_obs.csv"))
cell_var <- fread(file.path(workdir, "cell_var.csv"))
image_meta <- fread(file.path(workdir, "image_meta.csv"))
cell_neighbors <- fread(file.path(workdir, "relations_cell_neighbors.csv"))
```

### Add row and column names

```{r row-col-names}
# Set object_id as cell_obs row names
cell_obs <- cell_obs[, !duplicated(colnames(cell_obs)), with = FALSE]
cell_obs <- DataFrame(cell_obs)
rownames(cell_obs) <- cell_obs$object_id

# Set measurement_id as cell_var row names
cell_var <- DataFrame(cell_var)
rownames(cell_var) <- cell_var$V1
cell_var$V1 <- NULL

# Make values in the "goodname" column as unique
# (for barcoding channels and GFP)
cell_var[cell_var$goodname == "BC", ]$goodname <-
  paste0("BC_", cell_var[cell_var$goodname == "BC", ]$metal)

cell_var[cell_var$goodname == "GFP", ]$goodname <-
  paste0("GFP_", cell_var[cell_var$goodname == "GFP", ]$metal)

# Set row and column names for the measurement matrix
cell_X <- as.matrix(cell_X)
rownames(cell_X) <- rownames(cell_obs)
colnames(cell_X) <- rownames(cell_var)
```


## Cell metadata

### Extract spatial measurements

Measurements such as cell area, distance to rim, cell location, ... are extracted and added to cell observations. Units are `um`, unless otherwise specified.
- `Center_X/Y`: object centroid position in image.  
- `Area` : area of the cell (units = um^2).  
- `dist-rim`: estimated distance to spheroid border.  
- `dist-sphere`: distance to spheroid section border.  
- `dist-other`: distance to other spheroid section in the same image.  
- `dist-bg`: distance to background pixels.    

```{r spatial}
# Fix measurement_type for "dist-sphere", "dist-other" and "dist-bg"
cell_var[cell_var$goodname %in% c("dist-sphere", "dist-other", "dist-bg"),
         ]$measurement_type <- "Location"
cell_var[cell_var$goodname == "dist-sphere", ]$measurement_name <- "dist-sphere"
cell_var[cell_var$goodname == "dist-other", ]$measurement_name <- "dist-other"
cell_var[cell_var$goodname == "dist-bg", ]$measurement_name <- "dist-bg"

# Extract relevant columns
spatial_meas <- cell_var[cell_var$measurement_type != "Intensity",
                         "measurement_id"]
spatial <- cell_X[, colnames(cell_X)  %in% spatial_meas]

# Add column names
spatial_names <- cell_var[cell_var$measurement_type != "Intensity", "goodname"]
colnames(spatial) <- spatial_names

# Add the data to cell observations
spatial <- DataFrame(spatial)
cell_obs <- merge(cell_obs, spatial, by = "row.names")
```

### Import image metadata

See the [original dataset description](https://zenodo.org/record/4271910#.YGWWHD8kz-j) for description of the columns content.

```{r import-image-meta}
# Select relevant columns
keep_col <- c("image_id", "cellline", "condition_name", "concentration",
              "time_point", "condition_id", "plate_id", "well_name", "hastelox",
              "sampleblock_name", "acquisition_id", "site_id", "slide_id",
              "sampleblock_id", "image_shape_h", "image_shape_w",
              "image_stack_filename_FullStackComp", "mask_filename_cell")

image_meta <- DataFrame(image_meta[, ..keep_col])

# Merge cell and image metadata
cell_obs <- merge(cell_obs, image_meta, by = "image_id")
rownames(cell_obs) <- cell_obs$object_id


# Rename "object_number" as "CellNumber" for consistency with other datasets
colnames(cell_obs)[colnames(cell_obs) == "object_number"] <- "CellNumber"

# # Rename "image_id" as "ImageNumber" for consistency with other datasets
colnames(cell_obs)[colnames(cell_obs) == "image_id"] <- "ImageNumber"

# Remove merged data frames
remove(spatial, image_meta)
```


## Intensity measurements

### Select measurements

The measurement matrix contains different cell-level measurements:  
- `MeanIntensityComp`: mean intensity per cell, spillover-compensated.  
- `NbMeanMeanIntensityComp`:  mean intensity of neighboring cells, spillover-compensated.  

The other measurements are not retained in the final SCE object:  
- `MeanIntensity`:mean intensity measured on compensated images.   
- `MinIntensity`: min intensity measured on compensated images.  
- `MaxIntensity`: max intensity measured on compensated images.  
- `StdIntensity`: intensity std measured on compensated images.  

```{r select-measurements}
# Subset intensity columns in cell variables and measurement matrix
cell_var <- cell_var[cell_var$measurement_type == "Intensity", ]
cell_X <- cell_X[, colnames(cell_X)  %in% cell_var$measurement_id]

# Print the different cell intensity measurements
writeLines(unique(cell_var$measurement_name))
```

### Extract counts

Create one matrix for cell counts and one for average counts of neighboring cells.

```{r extract-counts}
# Select the ids of measurements to subset
cell_meas <- cell_var[cell_var$measurement_name == "MeanIntensityComp",
                      ]$measurement_id
neighb_meas <- cell_var[cell_var$measurement_name == "NbMeanMeanIntensityComp",
                        ]$measurement_id

# Create the two matrices
counts_X <- cell_X[, colnames(cell_X)  %in% cell_meas]
counts_neighb_X <- cell_X[, colnames(cell_X)  %in% neighb_meas]

# Make sure the rows of "cell_var" and of counts matrices are in the same order
counts_X <- counts_X[, order(match(
  colnames(counts_X),
  rownames(cell_var[cell_var$measurement_id %in% cell_meas, ])))]

counts_neighb_X <- counts_neighb_X[, order(match(
  colnames(counts_neighb_X),
  rownames(cell_var[cell_var$measurement_id %in% neighb_meas, ])))]

# Rename the columns of the matrices
colnames(counts_X) <-
  cell_var[cell_var$measurement_id %in% cell_meas, ]$goodname
colnames(counts_neighb_X) <-
  cell_var[cell_var$measurement_id %in% neighb_meas, ]$goodname

# Remove original counts matrix
remove(cell_X)
```


## Feature metadata

### Reformat cell variables data frame

```{r reformat-cell_var}
# Use the "goodname" column as rownames
rownames(cell_var) <- cell_var$goodname

# Keep only relevant columns for features ("cell_var")
col_keep <- c("metal", "goodname", "ref_plane_number", "working",
              "Antibody.Clone", "is_cc")
cell_var <- unique(cell_var[, colnames(cell_var) %in% col_keep])
```


## Generate a SingleCellExperiment object

### Add counts

Add cell mean intensities (as `counts`) and mean intensities of neighbouring cells (as `counts_neighb`).

```{r create-SCE}
sce <- SingleCellExperiment(assays = list(counts = t(counts_X)))
assay(sce, "counts_neighb") <- t(counts_neighb_X)
```

### Add cell and feature metadata

```{r row-col-Data}
# Match row and column names
cell_obs <- cell_obs[order(match(rownames(cell_obs), colnames(sce))), ]
cell_var <- cell_var[order(match(rownames(cell_var), rownames(sce))), ]

# Add metadata
colData(sce) <- cell_obs
rowData(sce) <- cell_var

# Order the dataset by ImageNumber and CellNumber
sce <- sce[, order(sce$ImageNumber, sce$CellNumber)]

# Add image names
sce$ImageName <- gsub("_cell.tiff", "", sce$mask_filename_cell)
```

### Add cell neighborhood information

Generate a neighborhood graph and add it to the metadata of the SCE object.

```{r neighb-graph}
g <- graph_from_data_frame(cell_neighbors)
metadata(sce) <- list(graph = g)
sce
```

### Save the SCE object

We will save the `SingleCellExperiment` object for upload to `r Biocpkg("ExperimentHub")`.

```{r save-SCE}
saveRDS(sce, file.path(outdir, "sce.rds"))

# Remove unused data objects
remove(g, cell_obs, cell_neighbors, cell_var, counts_X, counts_neighb_X)
```


# **Masks and Images**

## Cell segmentation masks

### Select masks

Remove mask images that are not in the SCE object

```{r select-masks}
mask_files <- list.files(file.path(workdir, "masks"))

# delete unneeded masks
mask_files <- mask_files[!(mask_files %in% sce$mask_filename_cell)]
unlink(file.path(workdir, "masks", mask_files), recursive = TRUE)
```


### Load the masks as a CytoImageList object

```{r load-masks}
masks <- loadImages(file.path(workdir, "masks"), pattern = "_cell.tiff")
```

### Scale masks

```{r scale-masks}
masks <- scaleImages(masks, value = (2 ^ 16) - 1)
```

### Add metadata

```{r mask-metadata}
mcols(masks)$ImageName <- gsub("_cell", "", names(masks))

# Add ImageNumber to te metadata columns of the `masks` object
masks <- masks[order(match(mcols(masks)$ImageName, unique(sce$ImageName)))]
mcols(masks)$ImageNumber <- unique(sce$ImageNumber)
```

### Save the masks as CytoImageList objects

Save the generated `CytoImageList` masks for upload to `r Biocpkg("ExperimentHub")`.

```{r save-masks-CIL}
saveRDS(masks, file.path(outdir, "masks.rds"))
```

## Multichannel images

### Select images

Remove multichannel images that are not in the SCE object

```{r select-images}
image_files <- list.files(file.path(workdir, "images"))

# delete unneeded images
image_files <- image_files[!(image_files %in%
                               sce$image_stack_filename_FullStackComp)]
unlink(file.path(workdir, "images", image_files), recursive = TRUE)
```

### Load the images as a CytoImageList object

```{r load-images}
images <- loadImages(file.path(workdir, "images"), pattern = "_comp.tiff")
```

### Scale images

```{r scale-images}
images <- scaleImages(images, value = (2 ^ 16) - 1)
```

### Add metadata

```{r images-metadata}
mcols(images)$ImageName <- gsub("_comp", "", names(images))

# Add ImageNumber to te metadata columns of the `images` object
images <- images[order(match(mcols(images)$ImageName, unique(sce$ImageName)))]
mcols(images)$ImageNumber <- unique(sce$ImageNumber)
```

### Add channel names

```{r add-channel-names}
channelNames(images) <- rowData(sce)[order(
  rowData(sce)[, "ref_plane_number"]), ]$goodname
```

### Save the images as CytoImageList objects

Save the generated `CytoImageList` images for upload to `r Biocpkg("ExperimentHub")`.

```{r save-images-CIL}
saveRDS(images, file.path(outdir, "images.rds"))
```

### Clean up

Remove the downloaded objects to save storage space.

```{r clean-up, message = FALSE}
downloaded_files <- list.files(workdir)
unlink(file.path(workdir, downloaded_files), recursive = TRUE)
options(timeout=timeout)
```

# Session information

```{r session-info}
sessionInfo()
```
